### 2017-02-22 manuscript/manuscript.tex
started methods

\documentclass{article}

\usepackage{color}
\usepackage{listings}

\begin{document}

\section*{Methods}
\subsection*{Data collection}
We have obtained monthly water temperature readings from 1978 through 2015 via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \colorbox{red}{\lstinline{cite}}. In all, 24 monitoring sites within the Puget Sound region (Fig. \colorbox{red}{\lstinline{1}}) were used in analyses. These represent 19 rivers across 9 counties, and range from 4 to 775 m in elevation. For one site at each river, monthly discharge time series were available, either for the same location as one of the temperature monitoring sites, or for another location within 30 km on the same major reach. Discharge data were aggregated by monthly mean from the USGS Washington Water Science Center (collected daily 1978-2007) and the USGS National Water Information System (collected at 15-minute intervals 2008-2015). At least one discharge monitoring site was available for every river represented in the temperature dataset.

Regional climate data were ... the Puget Sound Lowland and East Olympic/Cascade Foothills climate divisions

\end{document}

% https://wa.water.usgs.gov/data/realtime/adr/interactive/ #discharge daily
% https://waecy.maps.arcgis.com/apps/Viewer/index.html?appid=832e254169e640fba6e117780e137e7b #Q 15min
% http://www.horizon-systems.com/NHDPlus/NHDPlusV2_17.php #nhdplus
% ftp://newftp.epa.gov/EPADataCommons/ORD/NHDPlusLandscapeAttributes/StreamCat/HydroRegions/ #streamca
% http://www.ncdc.noaa.gov/cag/time-series/us/45/3/tavg/ytd/12/1895-2016?base_prd=true&firstbaseyear=1901&lastbaseyear=2000 #climate data

### 2017-02-22 03_model_fitting.R

@@ -24,10 +24,10 @@
 rm(list=ls()); cat('\014') #clear env and console
 
 # 0 - setup ####
-# setwd('C:/Users/Mike/git/stream_nuts_DFA/data/')
+setwd('C:/Users/Mike/git/stream_nuts_DFA/data/')
 # setwd('~/git/puget_sound_rivers_DFA/data')
 # setwd('Z:/stream_nuts_DFA/data/')
-setwd("C:/Users/vlahm/Desktop/stream_nuts_DFA/data")
+# setwd("C:/Users/vlahm/Desktop/stream_nuts_DFA/data")
 load('chemPhys_data/yys_bymonth.rda')
 DISCHARGE <- read.csv('discharge_data/discharge.csv', stringsAsFactors=FALSE, colClasses=c('date'='Date'))
 snowmelt <- read.csv('climate_data/snow_data/snowmelt.csv')
@@ -60,7 +60,7 @@ if (is.null(dev.list()) == TRUE){
 
 # response choices: COND FC NH3_N NO2_NO3 OP_DIS OXYGEN PH PRESS SUSSOL TEMP TP_P TURB
 # also DISCHARGE (from USGS)
-y_choice = 'DISCHARGE'
+y_choice = 'TEMP'
 # cov choices: meantemp meantemp_anom precip precip_anom hydroDrought hydroDrought_anom
 # maxtemp maxtemp_anom hdd hdd_anom, snowmelt
 #(snowmelt only available 1978-2015. also I haven't actually used snowmelt
@@ -88,7 +88,7 @@ scale = FALSE
 na_thresh = 0.55
 #transformations are 'log' and 'none' from here. can also explore 'power' and 'boxcox' in section 3.1
 #run function transformables() to see whether your response needs to be transformed.
-transform = 'log'
+transform = 'none'
 
 # 1.1 - subset data according to choices, remove problematic columns ####
 library(stringr)

### 2017-02-23 manuscript/manuscript.tex
added bayes dfa plotter. abandoning across-time approach. added equation to tex file

@@
 \usepackage{listings}
+\usepackage{bm}
 
@@
 
+todo:
+
+include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
+
+introduce terminology:
+responses = water temp, Q
+climatic predictors = air temp, precip, drought
+landscape predictors = many
+
+\section*{Introduction}
+
 \section*{Methods}
+\subsection*{Study region}
+I'll probably mention anything relevant to this in the intro.
+
 \subsection*{Data collection}
-We have obtained monthly water temperature readings from 1978 through 2015 via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \colorbox{red}{\lstinline{cite}}.
+The two response variables used in our analyses were water temperature and discharge.
+We obtained monthly water temperature readings from 1978 through 2015 via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \colorbox{red}{\lstinline{cite}}.
 In all, 24 monitoring sites within the Puget Sound region (Fig.
-\colorbox{red}{\lstinline{1}}) were used in analyses.
-These represent 19 rivers across 9 counties, and range from 4 to 775 m in elevation.
+\colorbox{red}{\lstinline{1}}) were included, representing 19 rivers across 9 counties, and ranging from 4 to 775 m in elevation.
 For one site at each river, monthly discharge time series were available, either for the same location as one of the temperature monitoring sites, or for another location within 30 km on the same major reach.
@@
 
-Regional climate data were ... the Puget Sound Lowland and East Olympic/Cascade Foothills climate divisions
+Potential climatic predictors of water temperature and discharge included mean and max air temperature, precipitation, and hydrologic drought, averaged by month across the response variable time series.
+These data are available through the U.S.
+Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI)\colorbox{red}{\lstinline{cite}}.
+We acquired climatic predictor data grouped by Washington State climate division, and all but two of our sites fell within divisions 3 (Puget Sound Lowland) and 4 (East Olympic/Cascade Foothills).
+We therefore aggregated these data by monthly mean across the two regions (after verifying their post-standardization similarity), resulting in a single dataset of four regional, climatic predictors.
+
+For post-hoc analyses of potential sub-watershed-scale drivers of response-predictor coupling, we amassed an additional set of landscape predictor data.
+These were collected individually for each of the watersheds that correspond to our 24 river sites, using the EPA's StreamCat (stream-catchment) data collection \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}.
+Landscape predictor categories include lithology, land use, population and road density, and soil type, as well as other categories summarized in \colorbox{red}{\lstinline{Table A1}}.
+
+% A snowmelt time series was then added to this dataset, using monthly mean SNOTEL records from six sites the USDA's Natural Resources Conservation Service
+\subsection*{Time series analysis}
+Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al.
+2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
+In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), covariates (which can have unique effects on each response series), and random error.
+We chose DFA over a traditional multivariate state space approach for two reasons.
+First, it provides advantages in computational efficiency, as 1-5 shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost.
+Second, in terms of identifying what drives the shared trends, having fewer trends allows greater inferential parsimony.
+Being a multivariate technique, DFA also provides an advantage over univariate alternatives in that covariance structure among responses can be specified and compared.
+All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software \colorbox{red}{\lstinline{Kristensen et al.
+2015}}, which we called using package TMB in R \colorbox{red}{\lstinline{R Core team 2016...}}.
+
+DFA takes the following form:
+
+\begin{equation}
+    \textbf{x}_t = \textbf{x}_{t-1} + \textbf{w}_t\textrm{, where } \textbf{w}_t \sim \textrm{MVN}(0,\textbf{R})
+\end{equation}
+\begin{equation}
+    \textbf{y}_t = \textbf{Zx}_t + \textbf{Dd}_t + \textbf{v}_t\textrm{, where } \textbf{v}_t \sim \textrm{MVN}(0,\textbf{Q})
+\end{equation}
+\begin{equation}
+    \textbf{x}_0 \sim \textrm{MVN}(\bm{\pi},\bm{\Lambda})
+\end{equation}
+
+At each time step \it{t}, the response variable \textbf{y} is a vector of
+
+\section*{Appendix A}
+Table A1:
+...at the scale of individual stream reaches (segments bounded by sources, confluences, or mouths) and their corresponding watersheds.
+Watersheds are calculated as land contributing flow to a reach, and have been determined for 2.6 million reaches within the conterminous United States.
 
@@
 % ftp://newftp.epa.gov/EPADataCommons/ORD/NHDPlusLandscapeAttributes/StreamCat/HydroRegions/ #streamca
-% http://www.ncdc.noaa.gov/cag/time-series/us/45/3/tavg/ytd/12/1895-2016?base_prd=true&firstbaseyear=1901&lastbaseyear=2000 #climate data
+% ncdc.noaa.gov/cag/time-series/us #climate data
+% https://www.nrcs.usda.gov/wps/portal/nrcs/detail/or/snow/?cid=nrcs142p2_046350 #snow
+% https://arxiv.org/pdf/1509.00660.pdf #TMB

### 2017-02-23 02_testing_or_evaluation.R

@@ -536,7 +536,7 @@ cov_and_seas <- designer(design=design, sections=sections, focal_months=focal_mo
 #feel free to uncomment and experiment with this stuff just to verify that
 #the regions barely differ in climate
 
-# #visualize
+#visualize
 # par(mfrow=c(3,2))
 # for(i in 1:ncol(covs3)){
 #     plot(scale(covs3[,i]), scale(covs4[,i]), main=colnames(covs3)[i])
@@ -553,7 +553,7 @@ cov_and_seas <- designer(design=design, sections=sections, focal_months=focal_mo
 # #precip
 # dfa3 <- runDFA(obs=dat_z, NumStates=mm, ErrStruc='DE', EstCovar=TRUE, Covars=rbind(cc,covs3_z[3,]))
 # dfa4 <- runDFA(obs=dat_z, NumStates=mm, ErrStruc='DE', EstCovar=TRUE, Covars=rbind(cc,covs4_z[3,]))
-# process_plotter_TMB(dfa3, mm)
+# process_plotter_TMB(dfa3, mm) #this function is defined in section 5.1
 # process_plotter_TMB(dfa4, mm)
 # #hydrologic drought
 # dfa5 <- runDFA(obs=dat_z, NumStates=mm, ErrStruc='DE', EstCovar=TRUE, Covars=rbind(cc,covs3_z[4,]))

### 2017-02-23 04_model_fitting_BAYES.R

@@ -4,22 +4,11 @@
 
 #NOTES:
 
-#collapse folds with ALT+O (windows, linux) or CMD+OPT+O (Mac); might have to do it twice
-
-#in this script, all you need to do is choose your options in section 1, and then set a few more
-#things in section 6.1
-
-#to quickly access any of the function definitions you come across,
-#put the cursor on the function name and hit F2
-
-#you should have at least 4 datapoints (observations x streams) for each parameter in model.
-#you can see the datapoints:params ratio in the parallel loop output.
-
-#if R crashes when you try to use runDFA,
-#use apply(dat_z, 2, function(x) sum(is.na(x))/length(x)) to see if you have any timepoints with
-#no data or 1 data point. these timepoints must either be removed or imputed (see section 1.2).
-
-#you will probably have to restart Rstudio between runs, due to wonkiness associated with parallelization.
+#the notes from 03_model_fitting.R apply here as well. Here, there is an additional section at the
+#bottom that can be used to run DFA models in STAN. This allows for credible intervals on effect sizes.
+#i made little effort to explain what's going on in section 7, because I didn't end up using
+#those results.
+#note to self: model object for 4-trend, diagonal and unequal, byMonth_acrossTime is on udrive (saved structures)
 
 rm(list=ls()); cat('\014') #clear env and console
 
@@ -1180,9 +1169,9 @@ if(!require('statss')){
 library(statss)
 
 #Eric's dfa function, with parallelization option included
-fit_dfa2 = function (y = y, covar = NULL, covar_index = NULL, num_trends = 2, 
-          varIndx = NULL, zscore = TRUE, iter = 4000, chains = 1, 
-          control = list(adapt_delta = 0.99), cores=1) 
+fit_dfa2 = function (y = y, covar = NULL, covar_index = NULL, num_trends = 2,
+          varIndx = NULL, zscore = TRUE, iter = 4000, chains = 1,
+          control = list(adapt_delta = 0.99), cores=1)
 {
     stan_dir = find.package("statss")
     model = paste0(stan_dir, "/exec/dfa.stan")
@@ -1211,38 +1200,38 @@ fit_dfa2 = function (y = y, covar = NULL, covar_index = NULL, num_trends = 2,
     mat_indx = matrix(0, P, K)
     start = 1
     for (k in 1:K) {
[…]

### 2017-02-24 manuscript/manuscript.tex
methods matrix

@@
 \usepackage{bm}
+\usepackage{amsmath}
 
@@
 2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
-In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), covariates (which can have unique effects on each response series), and random error.
+In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error.
 We chose DFA over a traditional multivariate state space approach for two reasons.
@@
 \begin{equation}
-    \textbf{x}_t = \textbf{x}_{t-1} + \textbf{w}_t\textrm{, where } \textbf{w}_t \sim \textrm{MVN}(0,\textbf{R})
+    \textbf{x}_t = \textbf{x}_{t-1} + \textbf{w}_t\textrm{, where } \textbf{w}_t \sim \textrm{MVN}(0,\textbf{Q})
 \end{equation}
 \begin{equation}
-    \textbf{y}_t = \textbf{Zx}_t + \textbf{Dd}_t + \textbf{v}_t\textrm{, where } \textbf{v}_t \sim \textrm{MVN}(0,\textbf{Q})
+    \textbf{y}_t = \textbf{Zx}_t + \textbf{Dd}_t + \textbf{v}_t\textrm{, where } \textbf{v}_t \sim \textrm{MVN}(0,\textbf{R})
 \end{equation}
 \begin{equation}
-    \textbf{x}_0 \sim \textrm{MVN}(\bm{\pi},\bm{\Lambda})
+    \textbf{x}_0 \sim \textrm{MVN}(0,\bm{\Lambda})
 \end{equation}
 
-At each time step \it{t}, the response variable \textbf{y} is a vector of
+At each time step {\it t}, the $m \times 1$ vector of shared trends \textbf{x} is a function of \textbf{x} in the previous step, plus normal error \textbf{w} (Eq.
+1).
+This is the definition of a random walk.
+The $n\times 1$ response vector \textbf{y} at time {\it t} is a function of the shared trends and their factor loadings (\textbf{Z}; $n\times m$), a $q\times 1$ vector of covariates \textbf{d} and their river-specific effects effects (\textbf{D}; $n\times q$), and a second normal error term \textbf{v} (Eq.
+2).
+\textbf{R} and \textbf{Q} are variance-covariance matrices of order m, and \textbf{Q} is set to identity for model identifiability (\colorbox{red}{\lstinline{Harvey 1989}}).
+The initial state of the shared trend vector, $\bm{x}_0$, is multivariate-normally distributed with a mean of zero and a diagonal variance-covariance matrix with large variance (e.g.
+5).
+Predictor data were standardized to facilitate comparison of effect sizes.
+Response data were centered on 0, but not scaled, to avoid error inflation \colorbox{red}{\lstinline{cite}}.
+
+Because we were interested in isolating the effects of climatic predictors on river temperature and discharge, we used fixed factors to absorb recurring seasonal variation not related to the predictors, with one factor level for each month.
+These factors were incorporated into the predictor matrix \textbf{d}.
+Thus, the coefficient of \textbf{D} relating, say, air temperature and water temperature, represents the effect size of the former on the latter.
+In other words, it is the change in water temperature for a unit change in air temperature over the course of the time series.
+We call this effect "coupling." We were also interested in coupling by month for specific predictors, which required that the focal predictor in a particular model be treated like the fixed factors of residual seasonal variation.
+Concretely,
+
+$$
+\begin{bmatrix} 
+    \theta_{a1} & \theta_{a2} & \theta_{a3} & \cdots & \theta_{aT} \\
+    \theta_{b1} & 0 & 0 & \cdots & 0 \\ 
+    0 & \theta_{b2} & 0 & \cdots & 0 \\
+    0 & 0 & \theta_{b3} & \cdots & 0 \\
+    \vdots & \vdots & \vdots & \ddots & \vdots \\
+    0 & 0 & 0 & \cdots & \theta_{bT} \\
+    1 & 0 & 0 & \cdots & 0 \\ 
+    0 & 1 & 0 & \cdots & 0 \\
+    0 & 0 & 1 & \cdots & 0 \\
+    \vdots & \vdots & \vdots & \ddots & \vdots \\
+    0 & 0 & 0 & \cdots & 1 \\
+\end{bmatrix}
+$$
+
+The shared trends represent environmental variation not captured by the climatic predictors.
 
@@
 % https://arxiv.org/pdf/1509.00660.pdf #TMB
+% Harvey, A. C. 1989. Forecasting, structural time series models and the Kalman flter. Cambridge University Press, Cambridge, UK.

### 2017-02-24 manuscript/manuscript.tex
finished methods

--- 2017-02-23 manuscript/manuscript.tex
@@
 
-\usepackage{color}
-\usepackage{listings}
-\usepackage{bm}
+\usepackage{color, listings, bm, amsmath, blkarray}
 
 \begin{document}
-
 todo:
-
 include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
-
 introduce terminology:
@@
 \section*{Methods}
+
 \subsection*{Study region}
+
 I'll probably mention anything relevant to this in the intro.
 
-\subsection*{Data collection}
+\subsection*{Water and climate data}
+
 The two response variables used in our analyses were water temperature and discharge.
@@
 
-For post-hoc analyses of potential sub-watershed-scale drivers of response-predictor coupling, we amassed an additional set of landscape predictor data.
-These were collected individually for each of the watersheds that correspond to our 24 river sites, using the EPA's StreamCat (stream-catchment) data collection \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}.
-Landscape predictor categories include lithology, land use, population and road density, and soil type, as well as other categories summarized in \colorbox{red}{\lstinline{Table A1}}.
+% A snowmelt time series was then added to this dataset, using monthly mean SNOTEL records from six sites the USDA's Natural Resources Conservation Service
 
-% A snowmelt time series was then added to this dataset, using monthly mean SNOTEL records from six sites the USDA's Natural Resources Conservation Service
 \subsection*{Time series analysis}
+
 Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al.
 2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
-In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), covariates (which can have unique effects on each response series), and random error.
+In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error.
 We chose DFA over a traditional multivariate state space approach for two reasons.
 First, it provides advantages in computational efficiency, as 1-5 shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost.
-Second, in terms of identifying what drives the shared trends, having fewer trends allows greater inferential parsimony.
+Second, in terms of identifying what drives the shared trends, having fewer of them allows for greater inferential parsimony.
 Being a multivariate technique, DFA also provides an advantage over univariate alternatives in that covariance structure among responses can be specified and compared.
@@
 \begin{equation}
-    \textbf{x}_t = \textbf{x}_{t-1} + \textbf{w}_t\textrm{, where } \textbf{w}_t \sim \textrm{MVN}(0,\textbf{R})
+    \textbf{x}_t = \textbf{x}_{t-1} + \textbf{w}_t\textrm{, where } \textbf{w}_t \sim \textrm{MVN}(0,\textbf{Q})
 \end{equation}
 \begin{equation}
-    \textbf{y}_t = \textbf{Zx}_t + \textbf{Dd}_t + \textbf{v}_t\textrm{, where } \textbf{v}_t \sim \textrm{MVN}(0,\textbf{Q})
+    \textbf{y}_t = \textbf{Zx}_t + \textbf{Dd}_t + \textbf{v}_t\textrm{, where } \textbf{v}_t \sim \textrm{MVN}(0,\textbf{R})
 \end{equation}
 \begin{equation}
-    \textbf{x}_0 \sim \textrm{MVN}(\bm{\pi},\bm{\Lambda})
+    \textbf{x}_0 \sim \textrm{MVN}(0,\bm{\Lambda})
 \end{equation}
 
-At each time step \it{t}, the response variable \textbf{y} is a vector of
+At time step {\it t}, the $m \times 1$ vector of shared trends (\textbf{x}) is a function of \textbf{x} in the previous step, plus normal error (\textbf{w}; $m\times 1$; Eq.
+1).
+This is the definition of a random walk.
+The $n\times 1$ response vector (\textbf{y}) at time {\it t} is a function of the shared trends and their factor loadings (\textbf{Z}; $n\times m$), covariates (\textbf{d}; $q\times 1$) and their river-specific effects (\textbf{D}; $n\times q$), and a second normal error term (\textbf{v}; $n\times 1$; Eq.
+2).
+\textbf{R} and \textbf{Q} are variance-covariance matrices of order m, and \textbf{Q} is set to identity for model identifiability (\colorbox{red}{\lstinline{Harvey 1989}}).
+The initial state of the shared trend vector ($\bm{x}_0$) is multivariate-normally distributed with a mean of zero and a diagonal variance-covariance matrix with large variance (e.g.
+5; Eq.
+3).
+Predictor data were standardized to facilitate comparison of effect sizes.
+Response data were centered on 0, but not scaled, to avoid error inflation \colorbox{red}{\lstinline{details and citation}}.
 
-\section*{Appendix A}
-Table A1:
-...at the scale of individual stream reaches (segments bounded by sources, confluences, or mouths) and their corresponding watersheds.
-Watersheds are calculated as land contributing flow to a reach, and have been determined for 2.6 million reaches within the conterminous United States.
+Because we were interested in isolating the effects of climatic predictors on river temperature and discharge, we used fixed factors to absorb recurring seasonal variation not related to the predictors, with one factor level for each month.
+These factors were incorporated into the covariate matrix (\textbf{d}).
+Thus, the coefficient in \textbf{D} relating, say, air temperature (predictor) and water temperature (response), represents the effect size of the former on the latter.
+In other words, it is the change in water temperature accompanying a unit change in air temperature across the whole time series.
+We call this relationship "coupling." We were also interested in coupling by month for specific predictors, which required that the focal predictor in a particular model be arranged seasonally.
+Concretely,
+
+$$
+\textbf{d} = \begin{blockarray}{cccccc}
+& \textrm{Jan}_{1978} & \textrm{Feb}_{1978} & \textrm{Mar}_{1978} & \cdots & \textrm{Dec}_{2015} \\
+\begin{block}{c(ccccc)}
+    1 & 1 & 0 & 0 & \cdots & 0 \\
+    2 & 0 & 1 & 0 & \cdots & 0 \\
+    3 & 0 & 0 & 1 & \cdots & 0 \\
+      & \vdots & \vdots & \vdots & \ddots & \vdots \\
+    12 & 0 & 0 & 0 & \cdots & 1 \\
+    13 & \theta_{precip}^{(1)} & \theta_{precip}^{(2)} & \theta_{precip}^{(3)} & \cdots & \theta_{precip}^{(T)} \\
+    14 & \theta_{air}^{(1)} & 0 & 0 & \cdots & 0 \\
+    15 & 0 & \theta_{air}^{(2)} & 0 & \cdots & 0 \\
+    16 & 0 & 0 & \theta_{air}^{(3)} & \cdots & 0 \\
+      & \vdots & \vdots & \vdots & \ddots & \vdots \\
+    25 & 0 & 0 & 0 & \cdots & \theta_{air}^{(T)} \\
+\end{block}
+\end{blockarray}
+$$
+
+is the covariate matrix structure necessary to account for exogenous seasonal variation (rows 1-12), and overall effect of precipitation (row 13), while also yielding the by-month effect of air temperature (rows 13-24) on the response (\textbf{y}).
+
+Additional, non seasonal variation due to exogenous effects loads onto the shared trends, and a portion of remaining variation is absorbed by error matrix \textbf{v}.
+We fit models using four unique error structures (\textbf{R}), to allow for different suites of unknown drivers affecting rivers.
+We included shared variance and zero covariance, individual variance and zero covariance, shared variance and shared covariance, and unconstrained error.
+Details on these structures and their implications can be found in colorbox{red}{\lstinline{Holmes et al.
+(2012)}}.
+The best models for water temperature and discharge were determined with AIC.
+
+\subsection*{Landscape predictors and post-hoc regression}
+
+After model selection, climatic predictor effect sizes for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling.
+To achieve this, we amassed an additional dataset of landscape features.
+These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) library \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}.
+Each site was mapped to an individual river reach, defined as a segment bounded on each end by a stream or river source, confluence, or mouth.
+The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database.
+Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), monthly runoff (mm), percent ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
+We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (\colorbox{red}{\lstinline{unit}}) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
 

### 2017-02-28 manuscript/manuscript.tex
WsSlope,snowmelt in methods, final Q run, new plots

@@
 \begin{document}
-todo:
-include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
-introduce terminology:
-responses = water temp, Q
-climatic predictors = air temp, precip, drought
-landscape predictors = many
+% todo:
+% include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
+% introduce terminology:
+% responses = water temp, Q
+% climatic predictors = air temp, precip, drought
+% landscape predictors = many
+%include hydrograph figures from kathy liederman and julian olden to show
+    %rain dom, rain and snow, snow dom
+%include the over-time plots in the supplemental information
 
@@
 
-Potential climatic predictors of water temperature and discharge included mean and max air temperature, precipitation, and hydrologic drought, averaged by month across the response variable time series.
-These data are available through the U.S.
+Potential climatic predictors of water temperature and discharge included mean and max air temperature, total precipitation, snowmelt, and hydrologic drought, averaged by month across the response variable time series.
+All but snowmelt are available through the U.S.
 Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI)\colorbox{red}{\lstinline{cite}}.
@@
 We therefore aggregated these data by monthly mean across the two regions (after verifying their post-standardization similarity), resulting in a single dataset of four regional, climatic predictors.
+A snowmelt time series was then added to this dataset, using monthly mean records from six SNOTEL sites (Bumping Ridge, Elbow Lake, Mount Crag, Park Creek Ridge, Stevens Pass, White Pass) listed by the USDA's Natural Resources Conservation Service \colorbox{red}{\lstinline{cite}}.
+We calculated monthly snowmelt for each site as the absolute value of negative differences in cumulative snow water equivalent from each month to the next.
+The snowmelt time series was assigned zeros for any positive differences (accumulations).
 
-% A snowmelt time series was then added to this dataset, using monthly mean SNOTEL records from six sites the USDA's Natural Resources Conservation Service
+%More information on all datasets can be found in \colorbox{red}{\lstinline{Appendix A}}.
 
@@
 Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), monthly runoff (mm), percent ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
-We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (\colorbox{red}{\lstinline{unit}}) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
+We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
 

### 2017-02-28 02_testing_or_evaluation.R

@@ -73,7 +73,7 @@ y_choice = 'DISCHARGE'
 #still be determined.
 #(snowmelt only available 1978-2015. also I haven't actually used snowmelt
 #in a model yet, so there could be bugs)
-cov_choices = c('meantemp','precip')
+cov_choices = c('meantemp','precip','snowmelt')
 #region choices: '3' (lowland), '4' (upland), '3_4' (average of 3 and 4, or each separately)
 #regions 3 and 4 were discovered to be very similar early on, so most of this script will only work if
 #you choose '3_4' here and 'average_regions=TRUE'.
@@ -88,13 +88,13 @@ method = 'fixed_individual'
 startyr = 1978
 endyr = 2015
 #model params (specific values only relevant for testing, not for parameter optimization loop)
-ntrends = 4
+ntrends = 5
 #error matrix can either hold the MARSS specifications or the TMB ones ('DE', 'DUE', 'EVCV', 'UNC')
 obs_err_var_struc = 'DUE'
 #UPDATE: Mark Schueurell no longer scales his response data. scaling forces the variance of
 #the D matrix to be small, thus artificially diminishing the impact of the covariates.
 scale = FALSE
-na_thresh = 0.3 #exclude sites with >= this proportion of NA values.
+na_thresh = 0.55 #exclude sites with >= this proportion of NA values.
 #be sure to visit section 3.1i, where you can add time interval factors and interaction effects
 #to the covariate matrix
 #transformations are 'log' and 'none' from here. can also explore 'power' and 'boxcox' in section 3.1
@@ -103,14 +103,14 @@ transform = 'log'
 #choose the covariate matrix design here. options are 'just_effect', 'effect_and_seasonality_without_interaction'
 #'effect_byMonth', and 'effect_byMonth_acrossTime'.
 #(see "designer" function in section 3.1 for details)
-design = 'effect_byMonth_acrossTime'
+design = 'effect_byMonth'
 #sections = number of intervals to divide the time series into, if examining change over time
[…]

### 2017-02-28 03_model_fitting.R

@@ -614,7 +614,9 @@ library(viridis)
 #these functions are used later within the loop
 
 process_plotter_TMB <- function(dfa_obj, ntrends){
-  par(mai=c(0.5,0.5,0.5,0.1), omi=c(0,0,0,0), mfrow=c(ntrends, 1))
+    if(ntrends<=4){
+        par(mai=c(0.5,0.5,0.5,0.1), omi=c(0,0,0,0), mfrow=c(ntrends, 1))
+    } else par(mai=c(0,0,0,0), omi=c(0,0,0,0), mfrow=c(ntrends, 1))
   xlbl <- int_dates
   y_ts <- int_dates
   ylm <- c(-1,1)*max(abs(dfa_obj$Estimates$u))
@@ -623,14 +625,16 @@ process_plotter_TMB <- function(dfa_obj, ntrends){
          ylim=ylm, xlab="", ylab="", xaxt="n")
     abline(h=0, col="gray")
     lines(y_ts,dfa_obj$Estimates$u[i,], lwd=2)
-    mtext(paste("Process",i), side=3, line=0.5)
+    mtext(paste("Process",i), side=3, line=-2)
     axis(1, at=xlbl, labels=xlbl, cex.axis=0.8)
   }
 }
 # process_plotter_TMB(dfa, mm)
 
 loading_plotter_TMB <- function(dfa_obj, ntrends){
-  par(mai=c(0.5,0.5,0.5,0.1), omi=c(0,0,0,0), mfrow=c(ntrends, 1))
+    if(ntrends<=4){
+        par(mai=c(0.5,0.5,0.5,0.1), omi=c(0,0,0,0), mfrow=c(ntrends, 1))
+    } else par(mai=c(0,0,0,0), omi=c(0,0,0,0), mfrow=c(ntrends, 1))
   ylbl <- names(obs_ts)
   clr <- viridis(nn) #colors may not line up with series plots in section 2
   ylm <- c(-1,1)*max(abs(dfa_obj$Estimates$u))
@@ -645,7 +649,7 @@ loading_plotter_TMB <- function(dfa_obj, ntrends){
       if(Z_rot[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt=90, adj=0, cex=1.2, col=clr[j])}
       abline(h=0, lwd=1.5, col="gray")
     }
-    mtext(paste("Factor loadings on process",i),side=3,line=0.5)
[…]

### 2017-02-28 data/watershed_data/slope/slope.R

@@ -1,0 +1,11 @@
+rm(list=ls()); cat('/014')
+setwd('C:/Users/Mike/git/stream_nuts_DFA/data/watershed_data/slope')
+library(manipulateR)
+ultimate_reader(dir_args=list(path='raw/',pattern='txt'), merge=FALSE)
+library(data.table)
+lst = mget(ls())
+for(i in 1:length(lst)) lst[[i]][1] = toupper(names(lst[i]))
+slope = rbindlist(lst)
+slope = subset(slope, select=c('Rowid_','MEAN'))
+colnames(slope) = c('siteCode','WsSlope')
+write.csv(slope, 'slope.csv', row.names=FALSE)

### 2017-02-28 manuscript/figures/figscript.R

@@ -40,10 +40,11 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('temp_due_4m_at_byMo_acrossTime_may-aug.rda')
 # load('temp_due_4m_at_byMo_acrossTime_nov-feb.rda')
 # load('temp_due_4m_at_byMo_acrossTime_MASO.rda')
-load('discharge_due_4m_atpc_byMo_allMos.rda')
+# load('discharge_due_4m_atpc_byMo_allMos.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
+load('discharge_due_5m_atpcsn_byMo_allMos.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric
@@ -390,45 +391,57 @@ axis(side=1, at=yy[,1][c(T,F)], labels=yy[,1][c(T,F)])
 legend(x='topleft', legend=c('Precip'), lwd=3, col=c('red'))
 par(defpar)
 
-# 9 - DISCHARGE effect size regression (untouched) ####
+# 9 - DISCHARGE effect size regression ####
 
+pdf('01b_discharge_effect_size_reg.pdf', width=8, height=3.75)
+landvar=land$ElevWs/100
+labels=list(expression(paste(T[air])),'Precip','Snowmelt')
+# covr=1
 land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
 
 #% of watershed area classified as ice/snow land cover (NLCD 2011 class 12)
 #% of watershed area classified as ice/snow land cover (NLCD 2006 class 12)
-# pdf('01_effect_size_reg.pdf', width=7, height=6)
 # png('01_effect_size_reg.png', width=7, height=6, units='in', res=96, type='cairo')
-defpar <- par(mar=c(5,5,2,5))
-pal <- colorRampPalette(c('brown', 'white'))
-cols <- pal(10)[as.numeric(cut(land$ElevWs, breaks=10))]
-plot(land$WsAreaOver1000*100, rescaled_effect_size, type='n',
[…]

### 2017-03-01 manuscript/manuscript.tex
small methods edits and some figure changes

@@
 3).
-Predictor data were standardized to facilitate comparison of effect sizes.
-Response data were centered on 0, but not scaled, to avoid error inflation \colorbox{red}{\lstinline{details and citation}}.
+Response and predictor data were standardized to facilitate comparison of effect sizes and avoid error inflation.
+
+% Response data were centered on 0, but not scaled, to avoid error inflation \colorbox{red}{\lstinline{details and citation}}.
 

### 2017-03-01 02_testing_or_evaluation.R

@@ -776,7 +776,7 @@ process_plotter_TMB <- function(dfa_obj, ntrends){
 }
 # pdf('../manuscript/figures/04_processes_and_loadings.pdf', width=7, height=4)
 # png('../manuscript/figures/04_processes_and_loadings.png', width=7, height=6, units='in', res=96, type='cairo')
-# process_plotter_TMB(dfa, mm)
+process_plotter_TMB(dfa, mm)
 
 loading_plotter_TMB <- function(dfa_obj, ntrends){
     if(ntrends<=4){
@@ -807,7 +807,7 @@ loading_plotter_TMB <- function(dfa_obj, ntrends){
         # mtext("Factor loadings", side=2, line=-21.7, outer=TRUE)
     }
 }
-# loading_plotter_TMB(dfa, mm)
+loading_plotter_TMB(dfa, mm)
 # dev.off()
 
 # full_fit <- dfa$Estimates$Z %*% dfa$Estimates$u + dfa$Estimates$D %*% rbind(cc,covs_z)
@@ -1062,7 +1062,7 @@ load_regress_plotter <- function(mmm, mode, var=NA, col_scale='ElevWs'){
     }
 }
 # load_regress_plotter(mm, 'indiv', 'PctIce2011Ws')
-# load_regress_plotter(mm, 'exploration', , 'ElevWs')
+load_regress_plotter(mm, 'exploration', , 'ElevWs')
 
 # 6 - best TEMP model (abandoned, but there's some useful plotting stuff here) ####
 

### 2017-03-01 manuscript/figures/figscript.R

@@ -395,8 +395,6 @@ par(defpar)
 
 pdf('01b_discharge_effect_size_reg.pdf', width=8, height=3.75)
 landvar=land$ElevWs/100
-labels=list(expression(paste(T[air])),'Precip','Snowmelt')
-# covr=1
 land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
 
 #% of watershed area classified as ice/snow land cover (NLCD 2011 class 12)
@@ -418,9 +416,13 @@ plot(landvar, rescaled_effect_size[,covr], type='n',
      xaxt='n',bty='o')
 mod <- lm(rescaled_effect_size[,covr] ~ I(landvar))
 sig <- ifelse(summary(mod)$coefficients[2,4]<=0.1, '*', '')
-if(covr == 1) mtext(bquote(bold(T[air]) ~.(sig)), 3, cex=1.1, font=2, line=1.2)
-if(covr %in% 2:3){
-    mtext(paste(labels[[covr]],sig), 3, cex=1.1, font=2, line=1.2)
+if(covr == 1) mtext(bquote(plain(theta) == bold(T[air]) ~.(sig)), 3, cex=1.1, font=2, line=1.2)
+if(covr == 2){
+    mtext(bquote(plain(theta) == bold('Precip') ~.(sig)), 3, cex=1.1, font=2, line=1.2)
+    abline(mod, col='gray', lty=2, lwd=3)
+}
+if(covr == 3){
+    mtext(bquote(plain(theta) == bold('Snowmelt') ~.(sig)), 3, cex=1.1, font=2, line=1.2)
     abline(mod, col='gray', lty=2, lwd=3)
 }
 points(landvar, rescaled_effect_size[,covr],

### 2017-03-02 manuscript/manuscript.tex
finished major figures, working on latex text overlay

@@
 \usepackage{color, listings, bm, amsmath, blkarray}
+% \usepackage[percent]{overpic}
+% \usepackage{tikz,graphicx}
+% \usepackage{graphicx}
 
@@
 To achieve this, we amassed an additional dataset of landscape features.
-These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) library \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}.
+These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}.
 Each site was mapped to an individual river reach, defined as a segment bounded on each end by a stream or river source, confluence, or mouth.
 The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database.
-Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), monthly runoff (mm), percent ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
+Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
 We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
+
+\section*{Results}
+
+% \begin{overpic}[grid,scale=0.5,unit=1mm]{figures/16_temp_all_reg.pdf}
+% \begin{overpic}[width=1\textwidth,grid,tics=10]{figures/16_temp_all_reg.pdf}
+% \put (1,55) {\small
+% chili chili chili chili chili chili chili chili chili chili chili chili chili chili
+% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
+% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
+% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
+% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
+% chili chili chili chili chili chili chili chili chili chili chili chili chili chili
+% }
+% \end{overpic}
+
+% \begin{tikzpicture}
+% \node (img) {\includegraphics[width=1\textwidth]{figures/16_temp_all_reg.pdf}};
+% \node [below left,text width=4cm,align=left] at (img.north east){\small Lorem ipsum dolor sit amet consectetur and then a bunch more stuff that no one remembers.};
+% \end{tikzpicture}
+
+\begin{picture}(100,100)
+\put(.1,.1){\includegraphics{figures/16_temp_all_reg.pdf}}
+% \put(.01\textwidth,.3\textwidth){hello}
+\end{picture}
+
+% \begin{figure}
+%   \includegraphics[width=\linewidth]{figures/16_temp_all_reg.pdf}
+%   \caption{A boat.}
+%   \label{fig:boat1}
+% \end{figure}
 

### 2017-03-02 02_testing_or_evaluation.R

@@ -64,7 +64,7 @@ if (is.null(dev.list()) == TRUE){
 
 # response choices: COND FC NH3_N NO2_NO3 OP_DIS OXYGEN PH PRESS SUSSOL TEMP TP_P TURB
 # also DISCHARGE (from USGS)
-y_choice = 'DISCHARGE'
+y_choice = 'TEMP'
 # cov choices: meantemp meantemp_anom precip precip_anom hydroDrought hydroDrought_anom
 # maxtemp maxtemp_anom hdd hdd_anom, snowmelt
 #if you're looking at effects by month, or by month over time (see 'design' option below),
@@ -99,7 +99,7 @@ na_thresh = 0.55 #exclude sites with >= this proportion of NA values.
 #to the covariate matrix
 #transformations are 'log' and 'none' from here. can also explore 'power' and 'boxcox' in section 3.1
 #run function transformables() to see whether your response needs to be transformed.
-transform = 'log'
+transform = 'none'
 #choose the covariate matrix design here. options are 'just_effect', 'effect_and_seasonality_without_interaction'
 #'effect_byMonth', and 'effect_byMonth_acrossTime'.
 #(see "designer" function in section 3.1 for details)
@@ -624,6 +624,7 @@ dfa <- runDFA(obs=dat_z, NumStates=mm, ErrStruc=obs_err_var_struc,
 # save.image('../manuscript/figures/temp_due_4m_at_byMo_acrossTime_MASO.rda')
 
 # save.image('../manuscript/figures/discharge_due_5m_atpcsn_byMo_allMos.rda')
+save.image('../manuscript/figures/temp_due_5m_atpcsn_byMo_allMos.rda')
 
 # 4.2 - or load desired model object ####
 
@@ -639,6 +640,7 @@ dfa <- runDFA(obs=dat_z, NumStates=mm, ErrStruc=obs_err_var_struc,
 # dfa <- readRDS('../round_12_byMoAcrossTime/marAprSepOct/model_objects_discharge/DISCHARGE_DUE_4m_fixed_factors_atpc_1978-2015.rds')
 
 # dfa <- readRDS('../round_13_byMo_allMos_scale_hiM/model_objects_discharge/DISCHARGE_DUE_5m_fixed_factors_atpcsn_1978-2015.rds')
[…]

### 2017-03-02 manuscript/figures/figscript.R

@@ -45,6 +45,7 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
 load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+# load('temp_due_5m_atpcsn_byMo_allMos.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric
@@ -96,6 +97,156 @@ axis(side=1, at=yy[,1][c(T,F)], labels=yy[,1][c(T,F)])
 lines(yy[,1], covs, col='red', lwd=3)
 legend(x='bottomleft', legend=c('Air temp'), lwd=3, col=c('red'))
 par(defpar)
+
+# 1.1 - TEMP effect size regression (linked with loading regression) ####
+
+# system('taskkill /f /im AcroRd32.exe')
+pdf('16_temp_all_reg.pdf', width=7.5, height=7.5)
+layout(matrix(c(1:6,9,7,8),nrow=3,byrow=TRUE))
+defpar = par(oma=c(0,0,1,1), mar=c(4,3.5,0,0))
+landvar=land$ElevWs/100
+land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
+
+#% of watershed area classified as ice/snow land cover (NLCD 2011 class 12)
+#% of watershed area classified as ice/snow land cover (NLCD 2006 class 12)
+# png('01_effect_size_reg.png', width=7, height=6, units='in', res=96, type='cairo')
+# defpar <- par(oma=c(2,4,2,1), mar=c(4,2,2,1))
+# pal <- colorRampPalette(c('brown', 'white'))
+# cols <- pal(10)[as.numeric(cut(land$ElevWs, breaks=10))]
+elev_hi = which(land$Ice06_11 >= 0.7)
+elev_med = which(land$Ice06_11 < 0.7 & land$ElevWs > 600)
+elev_lo = which(land$Ice06_11 < 0.7 & land$ElevWs <= 600)
+cols = rep('black', nrow(land))
+cols[elev_med] = 'gray60'; cols[elev_hi] = 'white'
+for(covr in 1:3){
+    res = rescaled_effect_size[,covr]
[…]

### 2017-03-03 manuscript/manuscript.tex
fixed issue with effect size backtransform. revised plots. abandoned discharge regression combo plot

@@
 % \usepackage[percent]{overpic}
-% \usepackage{tikz,graphicx}
+\usepackage{tikz}
 % \usepackage{graphicx}
+% \usepackage{epsfig}
 
@@
 %include the over-time plots in the supplemental information
+%in regression figure, line up left side of legend box with plot above
+    %make gray horiz line go only to topright corner of box
+    %make axis lines and plot borders cleaner (right now axis segments are darker)
+    %instead of elevation on x, use ice and color by elev (orange to blue) X no story
+    %for loadings, color by ice and make it darkgreen , midgreen, white X not without the above
+    % add % ice as one of the hidden trends
 
@@
 The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database.
-Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
+Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
 We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
@@
 
-\begin{picture}(100,100)
-\put(.1,.1){\includegraphics{figures/16_temp_all_reg.pdf}}
-% \put(.01\textwidth,.3\textwidth){hello}
-\end{picture}
+% \begin{picture}(100,100)
+% \put(.1,.1){\includegraphics{figures/16_temp_all_reg.pdf}}
+% % \put(.01\textwidth,.3\textwidth){hello}
+% \end{picture}
 
@@
 % \end{figure}
+
+% \begin{figure}[h!]
+%    \centering
+%    % \vspace{20pt}%
+%    \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
+%    % \includegraphics[width=20pt]{figures/16_temp_all_reg.pdf}
+%    % \caption{Epipolar Geometry}
+%    % \label{fig:cclogo}
+% \end{figure}
+
+% \begin{figure}[h!]
+% \begin{center}
+% % \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
+% \resizebox{\textwidth}{!}{\includegraphics{figures/16_temp_all_reg.pdf}}
+% \caption{Epipolar Geometry}
+% \label{fig:cclogo}
+% \end{center}
+% \end{figure}
+
+\begin{tikzpicture}
+    \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
+    \draw (-4.1, 0) node[text width=10em] {\footnotesize \textbf{Figure 1} };
+\end{tikzpicture}
 

### 2017-03-03 02_testing_or_evaluation.R

@@ -73,7 +73,7 @@ y_choice = 'TEMP'
 #still be determined.
 #(snowmelt only available 1978-2015. also I haven't actually used snowmelt
 #in a model yet, so there could be bugs)
-cov_choices = c('meantemp','precip','snowmelt')
+cov_choices = c('meantemp')#, 'precip', 'snowmelt')
 #region choices: '3' (lowland), '4' (upland), '3_4' (average of 3 and 4, or each separately)
 #regions 3 and 4 were discovered to be very similar early on, so most of this script will only work if
 #you choose '3_4' here and 'average_regions=TRUE'.
@@ -88,14 +88,15 @@ method = 'fixed_individual'
 startyr = 1978
 endyr = 2015
 #model params (specific values only relevant for testing, not for parameter optimization loop)
-ntrends = 5
+ntrends = 4
 #error matrix can either hold the MARSS specifications or the TMB ones ('DE', 'DUE', 'EVCV', 'UNC')
 obs_err_var_struc = 'DUE'
 #UPDATE: Mark Schueurell no longer scales his response data. scaling forces the variance of
 #the D matrix to be small, thus artificially diminishing the impact of the covariates.
+#Scaling is still recommended by the community at large, and I didn't see a big difference.
 scale = FALSE
-na_thresh = 0.55 #exclude sites with >= this proportion of NA values.
-#be sure to visit section 3.1i, where you can add time interval factors and interaction effects
+na_thresh = 0.3 #exclude sites with >= this proportion of NA values.
+#be sure to visit section 3.1, where you can add time interval factors and interaction effects
 #to the covariate matrix
 #transformations are 'log' and 'none' from here. can also explore 'power' and 'boxcox' in section 3.1
 #run function transformables() to see whether your response needs to be transformed.
@@ -103,14 +104,14 @@ transform = 'none'
 #choose the covariate matrix design here. options are 'just_effect', 'effect_and_seasonality_without_interaction'
 #'effect_byMonth', and 'effect_byMonth_acrossTime'.
 #(see "designer" function in section 3.1 for details)
-design = 'effect_byMonth'
+design = 'effect_byMonth_acrossTime'
[…]

### 2017-03-03 03_model_fitting.R

@@ -83,6 +83,7 @@ ntrends = 1
 obs_err_var_struc = 'diagonal and equal'
 #UPDATE: Mark Schueurell no longer scales his response data. scaling forces the variance of
 #the D matrix to be small, thus artificially diminishing the impact of the covariates.
+#Scaling is still recommended by the community at large, and I didn't see a big difference.
 scale = FALSE
 #exclude sites with >= this proportion of NA values.
 na_thresh = 0.55
@@ -755,33 +756,54 @@ best_landvars <- function(response, top){
 #this converts the scaled effect sizes back to their original scales, based on the original
 #SDs of the response and covariate(s). it cannot presently account for the effects of
 #transformation (and i doubt it can in general). if the model can't converge on untransformed
-#data, our only option is to log transform and report the effect sizes as such. see section
+#data, our only option is to log transform and report the effect sizes as such see section
 #1.3 for more.
+#I tried to make this function account for all the different covariate designs (except ..._acrossTime), but
+#there are tons of different contingencies to take into account, so I may have missed something.
+#if your effect size regressions don't look like what you expected, start here (again, I'm happy
+#to help). If you use design='effect_byMonth', this will
+#assume you have seasonality method='fixed...'. it might also work with fourier. but you'll need
+#seasonality to be included.
 eff_rescaler <- function(all_cov, seas, scaled=scale){
-  #get covariate effect sizes (D coefficients) from model, isolated from seasonal effects
-  if(nrow(all_cov) > 2){
-    # z_effect_size <- as.matrix(dfa$Estimates$D[,(nrow(seas)+1):ncol(dfa$Estimates$D)])
-    z_effect_size <- as.matrix(dfa$Estimates$D[,(nrow(seas)+1):(nrow(seas)+nrow(covs_z))])
-  } else {
-    z_effect_size <- dfa$Estimates$D
-  }
-
-  nstream <- nrow(z_effect_size)
-  ncov <- ncol(z_effect_size)
-
-  #convert effects back to original scale (units response/units covar)
-  #x/sd(response) = 1/sd(covar); x is the effect size scale factor
-  rescaled_effect_size <- matrix(NA, nrow=nstream, ncol=ncov)
-  for(i in 1:nstream){
-    sd_response <- trans$sds[i]
-    for(j in 1:ncov){
[…]

### 2017-03-03 manuscript/figures/figscript.R

@@ -15,6 +15,8 @@
 #the "effect size by month over time" plots are set up to work with 4 focal months.
 #if you specified a different number, you'll have to make changes.
 
+#discharge regression combo plot is scrapped. no relationship with snowmelt?
+#discharge only really relevant in the Q v temp plot.
 
 rm(list=ls()); cat('\014') #clear env and console
 
@@ -98,12 +100,47 @@ lines(yy[,1], covs, col='red', lwd=3)
 legend(x='bottomleft', legend=c('Air temp'), lwd=3, col=c('red'))
 par(defpar)
 
+# 2 - TEMP effect size regression (single, old) ####
+
+# land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
+# 
+# #% of watershed area classified as ice/snow land cover (NLCD 2011 class 12)
+# #% of watershed area classified as ice/snow land cover (NLCD 2006 class 12)
+# # pdf('01_effect_size_reg.pdf', width=7, height=6)
+# # png('01_effect_size_reg.png', width=7, height=6, units='in', res=96, type='cairo')
+# defpar <- par(mar=c(5,5,2,5))
+# pal <- colorRampPalette(c('brown', 'white'))
+# cols <- pal(10)[as.numeric(cut(land$ElevWs, breaks=10))]
+# plot(land$WsAreaOver1000*100, rescaled_effect_size, type='n',
+#      # plot(land$Ice06_11, rescaled_effect_size, type='n',
+#      xlab='Watershed area over 1000m (%)', main='',
+#      # xlab='Watershed % ice cover', main='',
+#      ylab=expression(paste(Delta, ' water', degree, 'C ',
+#                            Delta, ' air', degree, 'C'^-1)),
+#      cex.lab=1.3, cex.axis=1, font=2,
+#      xaxt='n', xlim=c(0,80))
+# mod <- lm(rescaled_effect_size ~ I(land$WsAreaOver1000*100))
+# # mod <- lm(rescaled_effect_size ~ land$Ice06_11)
+# abline(mod, col='gray', lty=2, lwd=3)
+# points(land$WsAreaOver1000*100, rescaled_effect_size,
+#        # points(land$Ice06_11, rescaled_effect_size,
+#        bg=cols, col='black',
[…]

### 2017-03-03 manuscript/figures/figscript.R
whoops. forgot these two things

@@ -15,9 +15,6 @@
 #the "effect size by month over time" plots are set up to work with 4 focal months.
 #if you specified a different number, you'll have to make changes.
 
-#discharge regression combo plot is scrapped. no relationship with snowmelt?
-#discharge only really relevant in the Q v temp plot.
-
 rm(list=ls()); cat('\014') #clear env and console
 
 # 0 - setup (choose temp/discharge here) ####

### 2017-03-04 manuscript/figures/figscript.R
all figures in finished form and added to manuscript

@@ -929,7 +929,8 @@ dev.off()
 # saveRDS(moInts, '../../saved_structures/moInts_temp_due_4m_at.rds')
 
 #save stuff for air water discharge plot
-saveRDS(list(covs, obs_ts), '../../saved_structures/air_water_discharge.rds') # from temp_due_4m_at_byMo_allMos.rda
+# saveRDS(list(covs, obs_ts), '../../saved_structures/air_water_discharge.rds') # from temp_due_4m_at_byMo_allMos.rda
+saveRDS(list(covs[,1,drop=FALSE], obs_ts), '../../saved_structures/air_water_discharge2.rds') # from temp_due_5m_atpcsn_byMo_allMos.rda
 
 
 # 16 - DISCHARGE vs. TEMP ####
@@ -1344,39 +1345,71 @@ shell('C:\\Users\\Mike\\git\\stream_nuts_DFA\\manuscript\\figures\\15_disch_vs_a
 
 # 20 - air water discharge (average) ####
 
-#load temp_due_4m_at_byMo_allMos.rda for this one
-awd <- readRDS('../../saved_structures/air_water_discharge.rds')
+#load discharge_due_5m_atpcsn_byMo_allMos.rda for this one
+awd <- readRDS('../../saved_structures/air_water_discharge2.rds')
 
-png('12_air_water_discharge.png', width=8, height=6, units='in', res=96, type='cairo')
-# pdf('12_air_water_discharge.pdf', width=8, height=6)
+elev_hi = which(land$Ice06_11 >= 0.7)
+elev_med = which(land$Ice06_11 < 0.7 & land$ElevWs > 600)
+elev_lo = which(land$Ice06_11 < 0.7 & land$ElevWs <= 600)
+# cols = rep('black', nrow(land))
+# cols[elev_med] = 'gray60'; cols[elev_hi] = 'white'
+
+# png('12_air_water_discharge.png', width=8, height=6, units='in', res=96, type='cairo')
+pdf('12_air_water_discharge.pdf', width=7, height=6)
 defpar <- par(mar=c(5,4,1,6))
-airtemps = watertemps = discharge = numeric()
+airtemps = watertemps_rain =watertemps_rs = watertemps_snow = discharge = numeric()
 for(i in 1:12){
     airtemps[i] <- mean(awd[[1]][seq(from=i, by=12, to=nrow(awd[[1]]))])
[…]

### 2017-03-06 manuscript/figures/figscript.R
insignif

@@ -43,8 +43,8 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
-load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-# load('temp_due_5m_atpcsn_byMo_allMos.rda')
+# load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+load('temp_due_5m_atpcsn_byMo_allMos.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric
@@ -100,7 +100,7 @@ par(defpar)
 # 2 - TEMP effect size regression (single, old) ####
 
 # land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
-# 
+#
 # #% of watershed area classified as ice/snow land cover (NLCD 2011 class 12)
 # #% of watershed area classified as ice/snow land cover (NLCD 2006 class 12)
 # # pdf('01_effect_size_reg.pdf', width=7, height=6)
@@ -175,8 +175,8 @@ for(covr in 1:3){
                          scriptstyle(plain('(')*plain(degree)*plain('C')~plain(cm^-1)*plain(')'))),
               2, cex=.8, line=1.5)
         axis(2, at=c(-.02,0,.02,.04), labels=c(-0.02,0.00,0.02,0.04), padj=.9, tck=-.02)
-        mtext(bquote(textstyle(bold('Perennial watershed ice/snow coverage'))~scriptstyle(plain('(%)'))), side=1, cex=.8, line=1.7)
-        # mtext(bquote(textstyle(bold('Mean watershed elevation'))~scriptstyle(plain('(100 m)'))), side=1, cex=.8, line=1.7)
+        # mtext(bquote(textstyle(bold('Perennial watershed ice/snow coverage'))~scriptstyle(plain('(%)'))), side=1, cex=.8, line=1.7)
+        mtext(bquote(textstyle(bold('Mean watershed elevation'))~scriptstyle(plain('(100 m)'))), side=1, cex=.8, line=1.7)
[…]

### 2017-03-07 manuscript/manuscript.tex
whoops. forgot these two things
all figures in finished form and added to manuscript
fixed error in temp-prec-disch plot, other plot mods, halfway through results

@@
 
-\usepackage{color, listings, bm, amsmath, blkarray}
+\usepackage{color, listings, bm, amsmath, blkarray, gensymb}
 % \usepackage[percent]{overpic}
@@
 % \usepackage{epsfig}
+% \usepackage{subfig}
+% \captionsetup{belowskip=12pt}
 
@@
     % add % ice as one of the hidden trends 
+%discharge regression combo plot is scrapped. no relationship with snowmelt?
+%discharge only really relevant in the Q v temp plot.
+%increase unit text size in fig 2
+%change area over 1000m to % area over 1000m and replot
+%change "dominated" to "dominant"
+%change "fall" to "autumn"    
+%dont size points in figure 2. include size in legend in fig 3
 
@@
 
-Potential climatic predictors of water temperature and discharge included mean and max air temperature, total precipitation, snowmelt, and hydrologic drought, averaged by month across the response variable time series.
+Potential climatic predictors of water temperature and discharge included mean and max air temperature (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index), averaged by month across the response variable time series.
 All but snowmelt are available through the U.S.
@@
 We included shared variance and zero covariance, individual variance and zero covariance, shared variance and shared covariance, and unconstrained error.
-Details on these structures and their implications can be found in colorbox{red}{\lstinline{Holmes et al.
+Details on these structures and their implications can be found in \colorbox{red}{\lstinline{Holmes et al.
 (2012)}}.
@@
 \subsection*{Landscape predictors and post-hoc regression}
+
+For post-hoc analyses, monitoring sites were separated into three classes based on relative areal coverage of perennial ice and/snow (hereafter ``\% glaciation'') and mean elevation across their watersheds.
+The three classes are loosely based on the classification scheme and language of the Climate Impacts Group at the University of Washington \colorbox{red}{\lstinline{cite}}, and are here delineated according to Table 1.
+
+% \begin{table}
+\begin{center}
+\textbf{Table 1} Watershed classification scheme
+\end{center}
+\begin{center}
+\begin{tabular}{ |c|c|c|c| } 
+ \hline
+ Classification & Abb. & Glaciation (\%) & Mean elev. (m) \\
+ \hline
+ Rain-dominated & RD & $< 0.7$ & $< 600$ \\
+ Rain-and-snow & RS & $< 0.7$ & $\geq 600$ \\
+ Snow-dominated & SD & $\geq 0.7$ & - \\
+ \hline
+\end{tabular}
+\end{center}
+% \end{table}
 
@@
 
-% \begin{overpic}[grid,scale=0.5,unit=1mm]{figures/16_temp_all_reg.pdf}
-% \begin{overpic}[width=1\textwidth,grid,tics=10]{figures/16_temp_all_reg.pdf}
-% \put (1,55) {\small
-% chili chili chili chili chili chili chili chili chili chili chili chili chili chili
-% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
-% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
-% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
-% chili chili chili chili chili chili chili chili chili chili chili chili chili chili\\
-% chili chili chili chili chili chili chili chili chili chili chili chili chili chili
-% }
-% \end{overpic}
+Mean monthly temperature trends for the three river classes, aggregated across all 38 years of data, deviated by a minimum of 1.0\degree C in December, and a maximum of 3.9\degree C in July (Fig.
+1).
+SD rivers remained approximately two degrees colder than their RS counterparts through mid-late summer, and 3-4 degrees colder than RD throughout spring and summer.
+RD rivers were consistently warmest throughout the year.
+In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD \colorbox{red}{\lstinline{stats throughout this section}}.
+RD, meanwhile, only attained a minimum of 5.6\degree C.
+RS reached a peak summer temperature of 16.9\degree C in July, while RS and SD followed in August with peak temperatures of 15.5 and 13.5\degree C, respectively.
 
-% \begin{tikzpicture}
-% \node (img) {\includegraphics[width=1\textwidth]{figures/16_temp_all_reg.pdf}};
-% \node [below left,text width=4cm,align=left] at (img.north east){\small Lorem ipsum dolor sit amet consectetur and then a bunch more stuff that no one remembers.};
-% \end{tikzpicture}
-
-% \begin{picture}(100,100)
-% \put(.1,.1){\includegraphics{figures/16_temp_all_reg.pdf}}
-% % \put(.01\textwidth,.3\textwidth){hello}
-% \end{picture}
-
-% \begin{figure}
-%   \includegraphics[width=\linewidth]{figures/16_temp_all_reg.pdf}
-%   \caption{A boat.}
-%   \label{fig:boat1}
-% \end{figure}
+Meanwhile, the amplitude of T\textsubscript{air} oscillation exceeded that of any river class, dipping below T\textsubscript{water} in autumn to a minimum of 3.2\degree C in December, and rising above RS and SD in March to an August maximum of 17.4\degree C.
+T\textsubscript{air} did not overtake RD T\textsubscript{water} until August, by which time the latter had begun to decline.
 
 % \begin{figure}[h!]
-%    \centering
-%    % \vspace{20pt}%
+\begin{center}
 % \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
-%    % \includegraphics[width=20pt]{figures/16_temp_all_reg.pdf}
-%    % \caption{Epipolar Geometry}
-%    % \label{fig:cclogo}
+\resizebox{\textwidth}{!}{\includegraphics{figures/12_air_water_discharge.pdf}}
+% \caption{Epipolar Geometry}
+% \label{fig:air_water_disch}
+\end{center}
 % \end{figure}
+\textbf{Figure 1} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
+All depicted series represent discrete data.
+\colorbox{red}{\lstinline{see appendix for standard deviations?}} \\[\baselineskip]
+% The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
+Spring peak discharge coincides noticeably with a separation in water temperature between SD and RS, while the summer trough coincides with separation of RD and T\textsubscript{air}.
+November marks both the autumn peak in discharge and the point at which T\textsubscript{air} falls below T\textsubscript{water}.
 
-% \begin{figure}[h!]
-% \begin{center}
-% % \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
-% \resizebox{\textwidth}{!}{\includegraphics{figures/16_temp_all_reg.pdf}}
-% \caption{Epipolar Geometry}
-% \label{fig:cclogo}
-% \end{center}
-% \end{figure}
+DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig.
+2a).
+On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a a $0.45\pm 0.17$\degree C change at SD sites.
+A similar trend was observed with respect to $Precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig.
+2b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
+There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
+2c).
+It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
+Both rivers drain glaciers of the Olympic Mountain Range, and both are SD, though the Elwha's watershed is larger.
 
@@
     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-    \draw (-4.1, 0) node[text width=10em] {\footnotesize \textbf{Figure 1} };
+    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 2} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+Regression lines indicate slopes significant at $\alpha=0.05$.};
 \end{tikzpicture}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
+\textbf{Figure 1} chili chili
+\end{center}
+
+Body text
+
+\section*{Discussion}
+The relationship between river classification and factor loadings is less clear
 
@@
 % Harvey, A. C. 1989. Forecasting, structural time series models and the Kalman flter. Cambridge University Press, Cambridge, UK.
+%CIG puget sound state of knowledge http://cses.washington.edu/picea/mauger/ps-sok/ps-sok_sec03_watercycle_2015.pdf

### 2017-03-07 manuscript/figures/figscript.R

@@ -43,8 +43,8 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
-load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-# load('temp_due_5m_atpcsn_byMo_allMos.rda')
+# load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+load('temp_due_5m_atpcsn_byMo_allMos.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric
@@ -134,10 +134,10 @@ par(defpar)
 # 1.1 - TEMP effect size regression (linked with loading regression) ####
 
 # system('taskkill /f /im AcroRd32.exe')
-pdf('16_temp_all_reg.pdf', width=7.5, height=7.5)
+# pdf('16_temp_all_reg.pdf', width=7.5, height=7.5)
 layout(matrix(c(1:6,9,7,8),nrow=3,byrow=TRUE))
 defpar = par(oma=c(0,0,1,1), mar=c(4,3.5,0,0))
-# landvar=land$Ice06_11
+# landvar=land$WsAreaOver1000/land$WsAreaSqKm #??
 landvar=land$ElevWs/100
 land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
 
@@ -161,8 +161,9 @@ for(covr in 1:3){
     plot(landvar, res, type='n', yaxt='n',
          # xlab='Watershed area over 1000m (%)', main='',
          xlab='', main='', ylab='', cex.lab=1.3, cex.axis=1, font=2, xaxt='n',bty='l')
-    mod <- lm(res ~ I(landvar))
+    mod <- lm(res ~ landvar)
     sig <- ifelse(summary(mod)$coefficients[2,4]<=0.1, '*', '')
+    sig = NULL #turning off sig plotting
     if(covr == 1){
         mtext(bquote(textstyle(bold(Delta)*bold(T[water])~bold(Delta)*bold(T[air]^-1))~
                          scriptstyle(plain('(')*plain(degree)*plain('C')~plain(degree)*plain(C^-1)*plain(')'))),
[…]

### 2017-03-07 manuscript/manuscript.tex
results at 80%

@@
 % \captionsetup{belowskip=12pt}
+% \hfuzz=5.002pt  %suppress overfull warnings
 
@@
 %dont size points in figure 2. include size in legend in fig 3
+%adjust letter position in corners of fig 2 (anchoring)
 
@@
 I'll probably mention anything relevant to this in the intro.
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/map/map.png}}
+\textbf{Figure 1} Site locations (red points) in relation to the region across which climate data were aggregated (colored topography).
+\end{center}
 
@@
 \subsection*{Time series analysis}
-
-Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al.
+% Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al.
 2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
@@
 Mean monthly temperature trends for the three river classes, aggregated across all 38 years of data, deviated by a minimum of 1.0\degree C in December, and a maximum of 3.9\degree C in July (Fig.
-1).
+2).
 SD rivers remained approximately two degrees colder than their RS counterparts through mid-late summer, and 3-4 degrees colder than RD throughout spring and summer.
@@
 
-Meanwhile, the amplitude of T\textsubscript{air} oscillation exceeded that of any river class, dipping below T\textsubscript{water} in autumn to a minimum of 3.2\degree C in December, and rising above RS and SD in March to an August maximum of 17.4\degree C.
+% Meanwhile, the amplitude of T\textsubscript{air} oscillation exceeded that of any river class, dipping below T\textsubscript{water} in autumn to a minimum of 3.2\degree C in December, and rising above RS and SD in March to an August maximum of 17.4\degree C.
 T\textsubscript{air} did not overtake RD T\textsubscript{water} until August, by which time the latter had begun to decline.
@@
 % \end{figure}
-\textbf{Figure 1} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
+
+% \textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
 All depicted series represent discrete data.
-\colorbox{red}{\lstinline{see appendix for standard deviations?}} \\[\baselineskip]
+\colorbox{red}{\lstinline{see appendix for standard deviations?}} \setlength{\parskip}{6pt}%\\[\baselineskip]
+
 % The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
@@
 
-DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig.
-2a).
+% DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig.
+3a).
 On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a a $0.45\pm 0.17$\degree C change at SD sites.
 A similar trend was observed with respect to $Precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig.
-2b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
+3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
 There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
-2c).
+3c).
 It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
@@
     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 2} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+    % \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
 Regression lines indicate slopes significant at $\alpha=0.05$.};
@@
 
+The most parsimonious temperature DFA model selected via AIC included five shared trends.
+Of these, four correlated significantly with at least one known watershed predictor.
+Figure 3 depicts the strongest correlated variables with each trend (insets d-e).
+These are, in arbitrary order of relevance, mean water table depth ($p<0.001, R^2=0.60$; Fig.
+3d), \% glaciation ($p<0.01, R^2=0.30$; Fig.
+3e), BFI ($p=0.01, R^2=0.25$; Fig.
+3f), and mean slope ($p<0.01, R^2=0.29$; Fig.
+3g).
+The fifth shared trend was not correlated with any variables in the watershed predictor dataset.
+
 \begin{center}
 \resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
-\textbf{Figure 1} chili chili
+\textbf{Figure 4} Relationship between $T_{air}\rightarrow T_{water}$ and $T_{air}\rightarrow Q$.
+Both axes are expressed per 1\degree C change in $T_{air}$.
+The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $T_{air}$.
 \end{center}
-
-Body text
 

### 2017-03-07 manuscript/figures/figscript.R

@@ -134,7 +134,7 @@ par(defpar)
 # 1.1 - TEMP effect size regression (linked with loading regression) ####
 
 # system('taskkill /f /im AcroRd32.exe')
-pdf('16_temp_all_reg.pdf', width=7.5, height=7.5)
+# pdf('16_temp_all_reg.pdf', width=7.5, height=7.5)
 layout(matrix(c(1:6,9,7,8),nrow=3,byrow=TRUE))
 defpar = par(oma=c(0,0,1,1), mar=c(4,3.5,0,0))
 # landvar=land$WsAreaOver1000/land$WsAreaSqKm #??
@@ -187,7 +187,7 @@ for(covr in 1:3){
         axis(2, padj=.9, tck=-.02)
     }
     print.letter(paste(letters[covr],sig), c(.9,.9), cex=1.8, font=2, col='steelblue')
-    points(landvar, res, 
+    points(landvar, res,
            bg=cols, col='black', cex=rescale(log(land$WsAreaSqKm),c(1.2,3.2)),
            # pch=21,
            pch=land$siteCode,
@@ -304,6 +304,9 @@ for(trnd in c(1,2,4,5)){
 par(defpar)
 # dev.off()
 # shell('C:\\Users\\Mike\\git\\stream_nuts_DFA\\manuscript\\figures\\16_temp_all_reg.pdf')
+
+#for results (loadings)
+summary(mod)
 
 # 3 - TEMP loading regression ####
 
@@ -1384,9 +1387,9 @@ axis(4, las=1, tck=-.01, hadj=.3, cex.axis=.8, col.axis=adjustcolor('darkslategr
      col.ticks=adjustcolor('darkslategray4', alpha.f=.7))
 # axis(4, tcl=0, col='white', labels='')
 polygon(x=c(0,.57,1:12,12.5,13), y=c(0,mean(c(discharge[1],discharge[12])),discharge,mean(c(discharge[1],discharge[12])),0), col=disch_col, border=NA)
-lines(x=c(12.44,12.44),y=c(0,4018), 
+lines(x=c(12.44,12.44),y=c(0,4018),
                          col='white', xpd=NA, lwd=1)
-lines(x=c(12.44,12.44),y=c(0,4018), 
+lines(x=c(12.44,12.44),y=c(0,4018),
                          col=adjustcolor('darkslategray4', alpha.f=.35), xpd=NA, lwd=1)
[…]

### 2017-03-08 manuscript/manuscript.tex
discussion 70%

--- 2017-03-07 manuscript/manuscript.tex
@@
 %dont size points in figure 2. include size in legend in fig 3
+%adjust letter position in corners of fig 2 (anchoring)
+%see if its possible to directly compare Z and D
+%include the average effect size of tair on twater in 3a
+% can i say how much elevation would inflence temperature independent of snowmelt?
+%examine change over time in just the elwha
 
@@
 \subsection*{Time series analysis}
-
 Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al.
@@
 Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
-We also calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
+Monitoring site elevation was also included.
+Finally, we calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
 
@@
 % \end{figure}
-\textbf{Figure 1} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
+
+\textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
 All depicted series represent discrete data.
-\colorbox{red}{\lstinline{see appendix for standard deviations?}} \\[\baselineskip]
-% The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
+\colorbox{red}{\lstinline{see appendix for standard deviations?}} \setlength{\parskip}{6pt}\\[\baselineskip]
+
+The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
 Spring peak discharge coincides noticeably with a separation in water temperature between SD and RS, while the summer trough coincides with separation of RD and T\textsubscript{air}.
@@
 
+\sloppy
 DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig.
-2a).
-On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a a $0.45\pm 0.17$\degree C change at SD sites.
+3a).
+On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a $0.45\pm 0.17$\degree C change at SD sites.
 A similar trend was observed with respect to $Precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig.
-2b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
+3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
 There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
-2c).
+3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} \colorbox{red}{\lstinline{see appendix}}.
 It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
@@
     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 2} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
 Regression lines indicate slopes significant at $\alpha=0.05$.};
@@
 
+In addition to the three climate predictors above, the best T\textsubscript{water} model also included five shared trends.
+Of these, four correlated significantly with at least one known watershed predictor.
+Figure 3 depicts the strongest correlated variables with each trend (insets d-e).
+These are, in arbitrary order of relevance, mean water table depth ($p<0.001, R^2=0.60$; Fig.
+3d), \% glaciation ($p<0.01, R^2=0.30$; Fig.
+3e), BFI ($p=0.01, R^2=0.25$; Fig.
+3f), and mean slope ($p<0.01, R^2=0.29$; Fig.
+3g).
+The fifth shared trend was not correlated with any variables in the watershed predictor dataset.
+
 \begin{center}
 \resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
-\textbf{Figure 1} chili chili
+\textbf{Figure 4} Relationship between $T_{air}\rightarrow T_{water}$ and $T_{air}\rightarrow Q$.
+Both axes are expressed per 1\degree C change in $T_{air}$.
+The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $T_{air}$.
+Regression lines indicate slopes significant at $\alpha=0.05$.
 \end{center}
 
-Body text
+To examine possible sub-season interactions between T\textsubscript{air}, T\textsubscript{water} and Q, we performed an additional DFA with Q as the response.
+In both models, T\textsubscript{air} was allowed to have unique monthly effects.
+These effects, taken together, can be understood in relation to the four quadrants of the Cartesian coordinate system (increasing clockwise from upper right; Fig.
+4).
+
+In mid-winter (exemplified by February), all river classes primarily occupy the first quadrant, signifying $T\textsubscript{air} \propto T\textsubscript{water}$ and $T\textsubscript{air} \propto Q$, where $\propto$ denotes proportionality.
+RD shows the weakest Q response.
+By spring, many RS and SD sites develop an inverse relationship between T\textsubscript{air} and T\textsubscript{water}, denoted $T\textsubscript{air} \propto \frac{1}{T\textsubscript{water}}$, while RD sites change little from their winter state.
+June and August see a procession of most sites into the near fourth quadrant, with SD trailing.
+This signifies $T\textsubscript{air} \propto \frac{1}{Q}$, though $T\textsubscript{air} \propto T\textsubscript{water}$ remains.
+One stark exception is again the Elwha river, which occupies quadrant three.
+By fall, RS and SD have begun progress back toward their winter states, led by SD.
+RD, meanwhile, remain essentially unmoved from summer.
 
 \section*{Discussion}
+
+The effects of climate on T\textsubscript{water}, determined by dynamic factor analysis, suggest that nearly all rivers included in our dataset were influenced strongly by air temperature, precipitation, and/or snowmelt across 38 years of monthly data (Fig.
+3).
+At most monitoring sites, T\textsubscript{water} closely tracked changes in T\textsubscript{air}, on average responding to increases and decreases with proportional movements of up to 66\% magnitude.
+However, some rivers only weakly track T\textsubscript{air}, and several patterns in the intensity of this coupling correlate strongly with watershed features relating to ice, groundwater, and slope.
+Glaciation and yearly snow burden are prominent among these, and for reasons of ecological and hydrological implication, the primary focus of the following discussion.
+
+Without any analysis, a "buffering" effect (hereafter contrasted with "coupling") of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig.
+2).
+The aggregate hydrograph peaks due to snowmelt from April to June, at the same time that the trajectories of RS and SD (snow-influenced rivers) start to drop off relative to RD.
+After snowmelt begins to subside, RS and SD recover with a noticeable jump.
+For rivers that receive glacial runoff (SD), this effect appears to remain, buffering them from summer temperature rise where RS rivers instead take on the character of RD (Fig.
+4).
+In an extreme case, the Elwha River was actually cooler in August during those years in which air temperature was higher, likely due to increased runoff from Carrie and Eel glaciers.
+The buffering effect of ice on river temperature is therefore two-fold, acting first on all snowmelt-influenced rivers through a cold-water pulse in spring, and then on a subset of those rivers throughout summer and fall, by way of glacial runoff.
+For RD rivers, which receive little to no input from ice, summer temperature is entirely dictated by that of the surrounding air, and whatever rain falls through it.
+Though higher-elevation watersheds will always produce colder water, independent of the influence of ice, it can be expected that RS and SD rivers will grow more similar to RD as regional temperatures warm and glaciers decline.
+That is to say, formerly reliably cold-water streams and associated habitats may see increases in both summer and winter average temperatures, as well as higher variability from year to year.
+The Elwha in particular may slip from its current state of high resistance to seasonal climatic changes.
+We tested for changes in mean and variance of $T\textsubscript{air}\rightarrow T\textsubscript{water}$ and $T\textsubscript{air}\rightarrow Q$ coupling between 1978 and 2015, but did not detect any widespread patterns \colorbox{red}{\lstinline{appendix B}}.
+
+In addition to the three climate predictors, five shared trends were fit by the most parsimonious DFA model.
+These represent additional drivers responsible for structuring water temperature across some or all of the 24 sites included in the analysis.
+Each monitoring sites' factor loading on a particular shared trend indicates the degree to which the trend accounted for variance in T\textsubscript{water} at that site.
+While the precise identities of these drivers cannot be attained with certainty, they can be inferred through correlation with predictor variables.
+In this way, we determined the most likely landscape drivers of T\textsubscript{water} to be perennial ice and snow cover, mean watershed slope, and groundwater influx.
+Slope's likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \colorbox{red}{\lstinline{cite}}.
+
+%experiment with vim marks
+
+% single out elwha as mostly unaffected by climate
+% explain that the effect has not changed over time (nor the variance)
+
+% fig1: buffering effect is visible as damping followed by jump in RS SD
+    SD RS colder throughout the year, less influenced by changes in air.
+    % spring snowmelt (hydrograph peak) is the most significant item
+    RD actually remains warmer than air throughout the spring, and matches it through July.
+% fig2: snow/glac streams are buffered, overall, from temp changes due to air and precip.
+    snow isn't signif because of the arrangement of points along the x axis. however, there's a negative
+        snowmelt->Twater for the glac streams and a positive one for the RS streams
+    Why might each of the landscape predictors be such?
+        water table depth: ground water farther from surface, less responsive ground water (to whatever it is)
+        BFI: also ground water influx
+        ice snow: separation of RD RS?: this could be the rain-on-snow effect
+        slope: faster and more turbulent water, greater mixing; also correlated with snow/glac
+    what could the other factor be (trend 3)? random error, or a geological or morphological or microclimatic feature not captured by our watershed predictors 
+fog3: step through
+
 The relationship between river classification and factor loadings is less clear
+
+\section*{Appendix A}
+
+DFA plots here
+
+\section*{Appendix B}
+
+over-time plots here
 

### 2017-03-08 manuscript/figures/figscript.R

@@ -1265,6 +1265,7 @@ for(i in c(0,2,4,6,8,11)){
         ymax=max(temp_moInts[-c(6,17,18,20,22)])
         mod <- lm(temp_moInts[,i][-c(6,17,18,20,22)] ~ disch_moInts[,i])
         sig <- ifelse(summary(mod)$coefficients[2,4]<=0.1, '*', '')
+        sig = NULL #suppressing sig plotting
         plot(disch_moInts[,i], temp_moInts[,i][-c(6,17,18,20,22)],
              yaxt='n', xaxt='n', bty='o', type='n', fg='gray60',
              xlim=c(-.15, max(disch_moInts)),
@@ -1287,8 +1288,9 @@ for(i in c(0,2,4,6,8,11)){
         #                  pch=24, cex=3, lwd=1, col='black', bg='white')
         points(disch_moInts[,i], temp_moInts[,i][-c(6,17,18,20,22)],# cex=1.5,
                pch=21, bg=cols, col='black',
-               # pch=land$siteCode, col=cols,
-               cex=rescale(land$WsAreaSqKm,c(1.2,3.2)))
+               # pch=land$siteCode, col=cols)
+               cex=1.3)
+               # cex=rescale(log(land$elevation_),c(1.2,3.2))) #WtDepWs WsAreaSqKm BFIWs WsSlope elevation_
         print.letter(paste0(month.abb[i], sig), c(.9,.9), cex=1.2, font=2)
         if(i %in% c(2,4,8)){
             # axis(4, las=2, line=.6, at=c(-.5,0,.5,1),

### 2017-03-08 manuscript/manuscript.tex
discussion 80%

@@
 Each monitoring sites' factor loading on a particular shared trend indicates the degree to which the trend accounted for variance in T\textsubscript{water} at that site.
-While the precise identities of these drivers cannot be attained with certainty, they can be inferred through correlation with predictor variables.
+While the precise identities of these drivers cannot be obtained with certainty, they can be inferred through correlation with predictor variables.
 In this way, we determined the most likely landscape drivers of T\textsubscript{water} to be perennial ice and snow cover, mean watershed slope, and groundwater influx.
-Slope's likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \colorbox{red}{\lstinline{cite}}.
+In the case of slope, the likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \colorbox{red}{\lstinline{(cite}}; Fig.
+3g).
+As for groundwater, greater influx (represented by baseflow index, or BFI; fig.
+3f) corresponds to greater {\it de}-coupling of climatic effects and river temperature, as groundwater should be insulated relative to surface water.
+For the same reason, greater depth of groundwater should be associated with better insulation and thus further decoupling (Fig.
+3d).
+The buffering effect of perennial ice and snow for SD rivers has already been discussed, but the separation of RD and RS rivers by factor loadings on the associated trend is worth noting (Fig.
+3e)
 
@@
         snowmelt->Twater for the glac streams and a positive one for the RS streams
-    Why might each of the landscape predictors be such?
-        water table depth: ground water farther from surface, less responsive ground water (to whatever it is)
-        BFI: also ground water influx
+    % Why might each of the landscape predictors be such?
+        % water table depth: ground water farther from surface, less responsive ground water (to whatever it is)
+        % BFI: also ground water influx
         ice snow: separation of RD RS?: this could be the rain-on-snow effect
-        slope: faster and more turbulent water, greater mixing; also correlated with snow/glac
+        % slope: faster and more turbulent water, greater mixing; also correlated with snow/glac
     what could the other factor be (trend 3)? random error, or a geological or morphological or microclimatic feature not captured by our watershed predictors
@@
 
-DFA plots here
+DFA plots here blargenshnargen
 

### 2017-03-09 manuscript/manuscript.tex
discussion complete

@@
 %examine change over time in just the elwha
+%%%%why am i removing additional seasonal variation (fixed factor)???
+%probably get snowmelt out of the picture? it may be misleading and a source of criticism
+%explain why some SD rivers might have a larger effect than RD (related to slope, groundwater?)
 
@@
 
-For post-hoc analyses, monitoring sites were separated into three classes based on relative areal coverage of perennial ice and/snow (hereafter ``\% glaciation'') and mean elevation across their watersheds.
+For post-hoc analyses, monitoring sites were separated into three classes based on relative areal coverage of perennial ice and/snow (hereinafter ``\% glaciation'') and mean elevation across their watersheds.
 The three classes are loosely based on the classification scheme and language of the Climate Impacts Group at the University of Washington \colorbox{red}{\lstinline{cite}}, and are here delineated according to Table 1.
@@
 
-Without any analysis, a "buffering" effect (hereafter contrasted with "coupling") of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig.
+Without any analysis, a "buffering" effect (hereinafter contrasted with "coupling") of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig.
 2).
@@
 3d).
-The buffering effect of perennial ice and snow for SD rivers has already been discussed, but the separation of RD and RS rivers by factor loadings on the associated trend is worth noting (Fig.
-3e)
+The buffering effect of perennial ice and snow on SD rivers has already been discussed, but the uniquely high factor loadings of RS rivers in relation to the associated trend are worth noting (Fig.
+3e).
+This trend may account for variation in RS due to traits shared by RD and SD, or to a "rain-on-snow" effect that may yield additional cold water in early spring.
+The fifth trend did not correlate strongly with any of the landscape predictors in our dataset.
+It may therefore represent additional, unknown drivers like marine or microclimatic effects, or it may simply account for random noise.
+
+The relationship between climate and river temperature is further influenced by the interaction of discharge, and the fates of rivers in the Puget Sound watershed can be best understood by examining these factors in combination (Fig.
+4).
+Whether rain-, both-, or snow-dominated, all rivers took on RD characteristics in winter, when the effects of ice lay latent.
+As a result, warmer Februaries on average yielded warmer rivers and higher flow (less precipitation bound in ice).
+The critical differences between river classes played out in spring and summer, and it's during these months that future purturbations due to changing climate may be felt most acutely.
+For example, warmer Aprils on average produced colder water at 9 out of 15 RS and SD sites.
+Though we determined discharge, groundwater, and slope to be likely components of this relationship, only melting ice could be credited with actually reversing it.
+Projected reductions in snowpack for the Pacific Northwest can therefore be expected to fundamentally alter the responses of currently snow-influenced rivers to yearly variation in spring temperature.
+In the longer term, changes can be expected for rivers that now receive the temperature-buffering effect of glacial runoff.
+Glaciers continue to decline across North America, with the disappearance of all glacial ice in \colorbox{red}{\lstinline{(extent}} projected for the year \colorbox{red}{\lstinline{(year}}.
+
+\section*{Conclusion}
+
+Temperature regimes across the rivers of the Puget Sound watershed are structured by a combination of climatic drivers at the regional scale, and geophysical drivers at watershed scales.
+In the absence of snow and ice, river temperature is closely coupled to that of the surrounding air, while contributions of snowmelt and glacial runoff can dampen or even reverse this coupling in spring and summer.
+In some cases, icemelt-influenced rivers exhibit stronger positive responses to climate patterns than their rain-driven counterparts.
+Our results suggest elevational variations in groundwater influx, total discharge, and watershed slope account for these patterns.
+However, while these factors may influence the degree of coupling between climatic drivers and water temperature, only snow and ice can reverse it.
+Since 1978, such reversals have been widespread, particularly during spring melt.
+Though we did not detect changes in this effect across historical observations, future reductions in snowpack and glacial mass are projected.
+Consequently, many rivers that now undergo the mildest seasonal temperature changes may be impacted most strongly.
+
+The effects of melting snow and ice characterize
+
+degree to which any given river's temperature responds to climatic fluctuations can be described
 
 %experiment with vim marks
-
-% single out elwha as mostly unaffected by climate
-% explain that the effect has not changed over time (nor the variance)
-
-% fig1: buffering effect is visible as damping followed by jump in RS SD
-    SD RS colder throughout the year, less influenced by changes in air.
-    % spring snowmelt (hydrograph peak) is the most significant item
-    RD actually remains warmer than air throughout the spring, and matches it through July.
-% fig2: snow/glac streams are buffered, overall, from temp changes due to air and precip.
-    snow isn't signif because of the arrangement of points along the x axis. however, there's a negative
-        snowmelt->Twater for the glac streams and a positive one for the RS streams
-    % Why might each of the landscape predictors be such?
-        % water table depth: ground water farther from surface, less responsive ground water (to whatever it is)
-        % BFI: also ground water influx
-        ice snow: separation of RD RS?: this could be the rain-on-snow effect
-        % slope: faster and more turbulent water, greater mixing; also correlated with snow/glac
-    what could the other factor be (trend 3)? random error, or a geological or morphological or microclimatic feature not captured by our watershed predictors
-fog3: step through
-
-The relationship between river classification and factor loadings is less clear
 

### 2017-03-10 manuscript/manuscript.tex
insignif
started intro

@@
 \section*{Introduction}
+
+The ecological condition of a stream or river, the life it supports, and the goods and services it provides, are influenced by the timing and magnitude of seasonal changes in water temperature.
+Temperature is a chief consideration in the management of fisheries, as it affects species distribution (Boisneau et al.
+2008), life history (Elliott 1994), and migration timing (Boscarino et al.
+2007).
+In the Puget Sound watershed of the American Pacific Northwest, [many species depend on temperature, and recreational and commercial fisheries depend on rivers maintaining temperautre within certain ranges].
+Physically, temperature determines vapor pressure, surface tension, density (Stevens et al.
+1975), gas solubility, reaction rates (Brezonik 1972) and more.
+These properties also affect the efficacy of electrical power generation (Foerster & Lillestam 2010), and drinking water production (Ramaker et al.
+2005).
+
+These in turn are governed by climate at regional scales, and by physical attributes of the landscape at watershed or sub-watershed scales.
+
+River networks, being fractal in structure, are naturally governed by environmental processes at multiple scales.
+Water temperature is controlled at regional scales by climatic drivers, including air temperature and precipitation.
+At watershed or sub-watershed scales, physical attributes of the landscape, including geomorphology and geology, further structure temperature regimes, either directly or through interaction with climate drivers.
+
+water temperature by air temperature, sediment influx by precipitation (Garner et al.
+2013)—and at local scales by geographic, geologic, and anthropogenic effects.
+Taken together, this hierarchical system is difficult to manage, as the physical characteristics of any one river may be largely the product of climatic drivers, while those of another may result primarily from within-watershed conditions.
+We argue that these characteristics, in particular water temperature and transported load, should be managed with specific attention to the relative influence of regional- and local-scale conditions for any given stream.
+The first aim of this chapter is to identify streams in the Puget Sound region whose temperature (°C), suspended solids (mg/L), and turbidity (NTU) are primarily predicted by regional temporal trends in air temperature and precipitation, and those that depart from regional patterns.
+The second aim is to identify watershed features (land-use, geology, mean elevation, etc., specifics below) that correlate with such departures.
+
+We hypothesize that:
+1.
+Air temperature will correlate strongly with water temperature for most streams (Ward 1985; Garner et al.
+2013).
+Deviations from this relationship will be strongest in those streams with glacially influenced hydrology (Lisi et al.
+2015).
+2.
+Patterns in suspended load and turbidity, being highly dependent on soil type, lithology, and human activity, will be largely determined by within-watershed variables.
+Of these, we expect proportional agricultural cover to be most predictive (Waters 1995).
 

### 2017-03-11 manuscript/manuscript.tex
finished intro and appendices, working on bibliography

--- 2017-03-09 manuscript/manuscript.tex
@@
-\documentclass{article}
+\documentclass[notitlepage]{article}
 
@@
 % \usepackage[percent]{overpic}
-\usepackage{tikz}
+\usepackage{tikz, lineno, apalike}
+\linenumbers
 % \usepackage{graphicx}
@@
 
+\author{Michael Vlah}
+\title{Determining multi-scale controls on river temperature: a time series approach}
+
 \begin{document}
+\pagenumbering{gobble}
 % todo:
@@
 %explain why some SD rivers might have a larger effect than RD (related to slope, groundwater?)
+%get snowmelt out of there. it's a watershed trait. not a regional one
+%add discharge locations to the map?
+
+\maketitle
+\vspace{10pc}
+\noindent
+\textbf{Summary:}
+
+\noindent
+5 data sources; watershed delineation; 2 DFAs: m=1:15, 7 covariate combinations, 4 err. struc., 2 seasonality models; TMB; parallel computing; Bayesian change/time
+\clearpage
+\pagenumbering{arabic}
+
+\section*{Abstract}
+Temperature is among the most important determinants of riverine biodiversity and health.
+It is therefore a primary freshwater management concern, particularly where cold-water fish are of high ecological, recreational, and commercial value.
+However, river temperature in the Puget Sound watershed of the Northwestern U.S.A.
+is affected by a great diversity of drivers at multiple spatial and temporal scales, and little is known of their interactions.
+We used dynamic factor analysis, a multivariate time-series technique, to examine relationships among these drivers, synthesizing long-term climate and fine-scale landcover data.
+We found that primarily rain-fed rivers experience large seasonal temperature fluctuations, which closely track atmospheric temperature, while snow-fed rivers tend to be weakly, and in some cases inversely, coupled with such fluctuations.
+Among watersheds, groundwater influx, land slope, and discharge further augment or dampen these relationships.
+Our results suggest the temperature of high-elevation rivers, absent the influence of ice, should be highly variable, and that glacially fed streams stand to see the largest changes in temperature regime under proposed climate scenarios.
+\clearpage
 
@@
 
+The ecological condition of a stream or river, the life it supports, and the goods and services it provides, are influenced by the timing and magnitude of seasonal changes in water temperature.
+Temperature is a chief consideration in the management of fisheries, as it affects species distribution \cite{Boisneau2008}, growth and reproduction \cite{mccullough1999review}, and migration timing \cite{boscarino2007effects}.
+In particular, In the Puget Sound watershed of the American Pacific Northwest, several salmonid species spawn, migrate, and emerge only within the bounds of a few degrees Celsius, and thrive under even greater temperature constraints \cite{carter2005effects}.
+As a result, the success of commercial and recreational fisheries that depend on the region's riverine habitat rests on many precarious factors.
+
+River networks, being fractal in structure, are naturally governed by environmental processes at multiple scales.
+Seasonal variation in water temperature in rivers of the Pacific Northwest is a function of the surrounding air, as well as precipitation and snowmelt \cite{eldridge1967water}.
+These drivers may in turn be mediated or supplemented by several aspects of watershed morphology at smaller scales, including slope, elevation, and geology \cite{poole2001ecological,lisi2013association}.
+Taken together, this hierarchical system complicates fishery management, as the temperature regime of one river may be the direct product of climate, while that of another may depend more on within-watershed conditions.
+
+Adding to this picture, flow regimes across rivers of the Puget Sound watershed vary with latitude and elevation \cite{reidy2012hydrogeomorphic,mauger2015CIG}, and can be classified broadly into three categories by flow source and hydrograph shape.
+Rain-dominated (RD) rivers receive little or no input from snowmelt, and thus peak in discharge during the rainy season, usually between October and February.
+Snow-dominated (SD) rivers instead see peak flow during spring snowmelt, often in April, May, or June.
+Between these extremes lies a third class of rain-and-snow-driven (RS) rivers, which have appreciable peaks at both times.
+
+Effective management plans must therefore integrate a diversity of factors across space and time in order to determine which rivers and watersheds are likely to see consequential changes under projected climate and land use scenarios.
+However, the understanding required to do so is limited by knowledge of relationships among temperature drivers at scale.
+
+% The physical properties of water also change with temperature, which measurably affects the efficacy of electrical power generation (Foerster & Lillestam 2010), and drinking water production (Ramaker et al.
+2005).
+% Physically, temperature determines vapor pressure, surface tension, density (Stevens et al. 1975), gas solubility, reaction rates (Brezonik 1972) and more.
+
+We sought to identify streams in the Puget Sound region whose temperatures fluctuate closely with regional trends in air temperature, precipitation, and snowmelt, and those that depart from regional patterns.
+Our second aim was to identify watershed features that correlate with such departures, and thus provide a nuanced basis for predicting impacts of water temperature on aquatic biodiversity and fishery health.
+We hypothesized that water temperature would track air temperature most closely in RD rivers (Ward 1985; Garner et al.
+2013).
+We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt (Lisi et al.
+2015) and with factors affecting heat capacity of water, including discharge (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing).
+
+%carter 2005 http://www.swrcb.ca.gov/northcoast/water_issues/programs/tmdls/shasta_river/060707/28appendixaetheeffectsoftemperatureonsteelheadtroutcohosalmonandchinooksalmonbiologyandfunction.pdf:
+    %usepa 1999
+
 \section*{Methods}
 
-\subsection*{Study region}
+% \subsection*{Study region and climate data}
 
-I'll probably mention anything relevant to this in the intro.
+% Monthly time series of mean and max air temperature (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index) were acquired for the time-span of 1978 to 2015.
+
+\subsection*{Water and climate data}
+
+We investigated climate and landscape controls on water temperature and discharge, as separate response variables, from 1978 to 2015.
+Monthly time series of water temperature were obtained for 24 river sites via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \colorbox{red}{\lstinline{cite}}.
+These sites represent 19 nonnested watersheds across 9 counties, and range from 4 to 775 m in elevation.
+For at least one site at each river, monthly discharge time series were also available, either from the same location as one of the temperature monitoring sites, or from within 30 km on the same major reach.
+Discharge data were aggregated by monthly mean from the USGS Washington Water Science Center (collected daily 1978-2007) and the USGS National Water Information System (collected at 15-minute intervals 2008-2015).
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/map/map.png}}
-\textbf{Figure 1} Site locations (red points) in relation to the region across which climate data were aggregated (colored topography).
+\textbf{Figure 1} Site locations (red points) in relation to combined Washington State Climate Divisions 3 and 4 (colored topography), the region across which climate data were aggregated.
 \end{center}
 
-\subsection*{Water and climate data}
-
-The two response variables used in our analyses were water temperature and discharge.
-We obtained monthly water temperature readings from 1978 through 2015 via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \colorbox{red}{\lstinline{cite}}.
-In all, 24 monitoring sites within the Puget Sound region (Fig.
-\colorbox{red}{\lstinline{1}}) were included, representing 19 rivers across 9 counties, and ranging from 4 to 775 m in elevation.
-For one site at each river, monthly discharge time series were available, either for the same location as one of the temperature monitoring sites, or for another location within 30 km on the same major reach.
-Discharge data were aggregated by monthly mean from the USGS Washington Water Science Center (collected daily 1978-2007) and the USGS National Water Information System (collected at 15-minute intervals 2008-2015).
-At least one discharge monitoring site was available for every river represented in the temperature dataset.
-
 Potential climatic predictors of water temperature and discharge included mean and max air temperature (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index), averaged by month across the response variable time series.
-All but snowmelt are available through the U.S.
+All but snowmelt were available through the U.S.
 Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI)\colorbox{red}{\lstinline{cite}}.
-We acquired climatic predictor data grouped by Washington State climate division, and all but two of our sites fell within divisions 3 (Puget Sound Lowland) and 4 (East Olympic/Cascade Foothills).
-We therefore aggregated these data by monthly mean across the two regions (after verifying their post-standardization similarity), resulting in a single dataset of four regional, climatic predictors.
+We acquired climatic predictor data grouped by Washington State climate division, and all but two of our sites fell within divisions 3 (Puget Sound Lowland) and 4 (East Olympic/Cascade Foothills; see Fig.
+1).
+We therefore aggregated these data by monthly mean across the two regions (after verifying their post-standardization similarity), resulting in a single dataset of four climatic predictor variables.
 A snowmelt time series was then added to this dataset, using monthly mean records from six SNOTEL sites (Bumping Ridge, Elbow Lake, Mount Crag, Park Creek Ridge, Stevens Pass, White Pass) listed by the USDA's Natural Resources Conservation Service \colorbox{red}{\lstinline{cite}}.
@@
 We chose DFA over a traditional multivariate state space approach for two reasons.
-First, it provides advantages in computational efficiency, as 1-5 shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost.
+First, it provides advantages in computational efficiency, as a small number of shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost.
 Second, in terms of identifying what drives the shared trends, having fewer of them allows for greater inferential parsimony.
@@
 Consequently, many rivers that now undergo the mildest seasonal temperature changes may be impacted most strongly.
+\clearpage
 
-The effects of melting snow and ice characterize
-
-degree to which any given river's temperature responds to climatic fluctuations can be described
-
-%experiment with vim marks
+\bibliography{manuscript}
+\bibliographystyle{apalike}
+\clearpage
 
@@
 
-DFA plots here blargenshnargen
+\subsection*{Temperature DFA output and diagnostics}
+Model selection included four climate covariates (air temperature, precipitation, snowmelt, and hydrological drought), between 1 and 15 shared trends, four within-and-among-site error structures (see methods), and two models of unknown seasonal variation (fixed monthly factors and Fourier series).
+The most parsimonious model of river temperature was selected using the Akaike Information Criterion (AIC), and included air temperature, precipitation, and snowmelt as covariates.
+This model also included five shared trends and an independent and unequally distributed error structure among streams (i.e.
+diagonal and unequal variance-covariance matrix).
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/01_trends.pdf}}
+\textbf{Figure A1} Shared trends.
+\end{center}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/02_loadings.pdf}}
+\textbf{Figure A2} Factor loadings on shared trends.
+\end{center}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/03_fits.pdf}}
+\textbf{Figure A3} Model fits (gray line = overall; black line = trends-only).
+\end{center}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/04_residuals.pdf}}
+\textbf{Figure A4} Residuals.
+\end{center}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/05_acf.pdf}}
+\textbf{Figure A5} Autocovariance function (ACF).
+\end{center}
+
+Output from the discharge model looks very similar to this, and is omitted here for the sake of retaining a reasonable page count.
+\clearpage
 
@@
 
-over-time plots here
+\subsection*{Testing for change in coupling over time}
+
+We used an additional DFA model to test for changes in $T_{air}\rightarrow T_{water}$ coupling over time, by dividing the 1978-2015 time series into 5 intervals and comparing central tendency and variance of effect sizes for each interval.
+Figures B1-B3 show mean effect size for each stream.
+
+To approximate estimates of variability over time, we performed the same analysis within a Bayesian framework, and obtained uncertainty estimates from the credible intervals of the effect size posteriors.
+This approach yielded no trends in variation over time, and is not visualized here (This will be included in the final version of this paper).
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/05a_temp_effSize_byMonth_acrossTime_may-aug.pdf}}
+\textbf{Figure B1} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
+\end{center}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/05b_temp_effSize_byMonth_acrossTime_nov-feb.pdf}}
+\textbf{Figure B2} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
+\end{center}
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/05c_temp_effSize_byMonth_acrossTime_MASO.pdf}}
+\textbf{Figure B3} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
+\end{center}
 

### 2017-03-11 02_testing_or_evaluation.R

@@ -73,7 +73,7 @@ y_choice = 'TEMP'
 #still be determined.
 #(snowmelt only available 1978-2015. also I haven't actually used snowmelt
 #in a model yet, so there could be bugs)
-cov_choices = c('meantemp')#, 'precip', 'snowmelt')
+cov_choices = c('meantemp', 'precip', 'snowmelt')
 #region choices: '3' (lowland), '4' (upland), '3_4' (average of 3 and 4, or each separately)
 #regions 3 and 4 were discovered to be very similar early on, so most of this script will only work if
 #you choose '3_4' here and 'average_regions=TRUE'.
@@ -88,14 +88,14 @@ method = 'fixed_individual'
 startyr = 1978
 endyr = 2015
 #model params (specific values only relevant for testing, not for parameter optimization loop)
-ntrends = 4
+ntrends = 5
 #error matrix can either hold the MARSS specifications or the TMB ones ('DE', 'DUE', 'EVCV', 'UNC')
 obs_err_var_struc = 'DUE'
 #UPDATE: Mark Schueurell no longer scales his response data. scaling forces the variance of
 #the D matrix to be small, thus artificially diminishing the impact of the covariates.
 #Scaling is still recommended by the community at large, and I didn't see a big difference.
-scale = FALSE
-na_thresh = 0.3 #exclude sites with >= this proportion of NA values.
+scale = TRUE
+na_thresh = 0.55 #exclude sites with >= this proportion of NA values.
 #be sure to visit section 3.1, where you can add time interval factors and interaction effects
 #to the covariate matrix
 #transformations are 'log' and 'none' from here. can also explore 'power' and 'boxcox' in section 3.1
@@ -104,14 +104,14 @@ transform = 'none'
 #choose the covariate matrix design here. options are 'just_effect', 'effect_and_seasonality_without_interaction'
 #'effect_byMonth', and 'effect_byMonth_acrossTime'.
 #(see "designer" function in section 3.1 for details)
-design = 'effect_byMonth_acrossTime'
[…]

### 2017-03-11 manuscript/manuscript.tex
completed first draft

@@
 % \usepackage[percent]{overpic}
-\usepackage{tikz, lineno, apalike}
+\usepackage{tikz, lineno, natbib, url}
 \linenumbers
@@
 %add discharge locations to the map?
+%probably should include a table of data sources and station ids
 
@@
 The ecological condition of a stream or river, the life it supports, and the goods and services it provides, are influenced by the timing and magnitude of seasonal changes in water temperature.
-Temperature is a chief consideration in the management of fisheries, as it affects species distribution \cite{Boisneau2008}, growth and reproduction \cite{mccullough1999review}, and migration timing \cite{boscarino2007effects}.
-In particular, In the Puget Sound watershed of the American Pacific Northwest, several salmonid species spawn, migrate, and emerge only within the bounds of a few degrees Celsius, and thrive under even greater temperature constraints \cite{carter2005effects}.
+Temperature is a chief consideration in the management of fisheries, as it affects species distribution \citep{Boisneau2008}, growth and reproduction \citep{mccullough1999review}, and migration timing \citep{boscarino2007effects}.
+In particular, In the Puget Sound watershed of the American Pacific Northwest, several salmonid species spawn, migrate, and emerge only within the bounds of a few degrees Celsius, and thrive under even greater temperature constraints \citep{carter2005effects}.
 As a result, the success of commercial and recreational fisheries that depend on the region's riverine habitat rests on many precarious factors.
@@
 River networks, being fractal in structure, are naturally governed by environmental processes at multiple scales.
-Seasonal variation in water temperature in rivers of the Pacific Northwest is a function of the surrounding air, as well as precipitation and snowmelt \cite{eldridge1967water}.
-These drivers may in turn be mediated or supplemented by several aspects of watershed morphology at smaller scales, including slope, elevation, and geology \cite{poole2001ecological,lisi2013association}.
+Seasonal variation in water temperature in rivers of the Pacific Northwest is a function of the surrounding air, as well as precipitation and snowmelt \citep{eldridge1967water}.
+These drivers may in turn be mediated or supplemented by several aspects of watershed morphology at smaller scales, including slope, elevation, and geology \citep{poole2001ecological,lisi2013association}.
 Taken together, this hierarchical system complicates fishery management, as the temperature regime of one river may be the direct product of climate, while that of another may depend more on within-watershed conditions.
 
-Adding to this picture, flow regimes across rivers of the Puget Sound watershed vary with latitude and elevation \cite{reidy2012hydrogeomorphic,mauger2015CIG}, and can be classified broadly into three categories by flow source and hydrograph shape.
+Adding to this picture, flow regimes across rivers of the Puget Sound watershed vary with latitude and elevation \citep{reidy2012hydrogeomorphic,mauger2015CIG}, and can be classified broadly into three categories by flow source and hydrograph shape.
 Rain-dominated (RD) rivers receive little or no input from snowmelt, and thus peak in discharge during the rainy season, usually between October and February.
@@
 
-Effective management plans must therefore integrate a diversity of factors across space and time in order to determine which rivers and watersheds are likely to see consequential changes under projected climate and land use scenarios.
+Effective management plans must therefore integrate a diversity of factors across space and time in order to determine which rivers and watersheds are likely to see consequential changes under projected climate and land use scenarios for the Pacific Northwest \citep{mote2010future,radeloff2012economic}.
 However, the understanding required to do so is limited by knowledge of relationships among temperature drivers at scale.
@@
 Our second aim was to identify watershed features that correlate with such departures, and thus provide a nuanced basis for predicting impacts of water temperature on aquatic biodiversity and fishery health.
-We hypothesized that water temperature would track air temperature most closely in RD rivers (Ward 1985; Garner et al.
-2013).
-We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt (Lisi et al.
-2015) and with factors affecting heat capacity of water, including discharge (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing).
+We hypothesized that water temperature would track air temperature most closely in RD rivers \citep{ward1985thermal,garner2014river}.
+We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt \citep{lisi2015watershed} and with factors affecting heat capacity of water, including discharge (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing; \citealt{van2013global}).
 
@@
 
+% https://wa.water.usgs.gov/data/realtime/adr/interactive/ #discharge daily
+% https://waecy.maps.arcgis.com/apps/Viewer/index.html?appid=832e254169e640fba6e117780e137e7b #Q 15min
+% http://www.horizon-systems.com/NHDPlus/NHDPlusV2_17.php #nhdplus
+% ftp://newftp.epa.gov/EPADataCommons/ORD/NHDPlusLandscapeAttributes/StreamCat/HydroRegions/ #streamca
+% ncdc.noaa.gov/cag/time-series/us #climate data
+% https://www.nrcs.usda.gov/wps/portal/nrcs/detail/or/snow/?cid=nrcs142p2_046350 #snow
+% https://arxiv.org/pdf/1509.00660.pdf #TMB
+% Harvey, A. C. 1989. Forecasting, structural time series models and the Kalman flter. Cambridge University Press, Cambridge, UK.
+%CIG puget sound state of knowledge http://cses.washington.edu/picea/mauger/ps-sok/ps-sok_sec03_watercycle_2015.pdf
+
 \subsection*{Water and climate data}
@@
 We investigated climate and landscape controls on water temperature and discharge, as separate response variables, from 1978 to 2015.
-Monthly time series of water temperature were obtained for 24 river sites via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \colorbox{red}{\lstinline{cite}}.
+Monthly time series of water temperature were obtained for 24 river sites via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \citep{DoEwaterData}.
 These sites represent 19 nonnested watersheds across 9 counties, and range from 4 to 775 m in elevation.
 For at least one site at each river, monthly discharge time series were also available, either from the same location as one of the temperature monitoring sites, or from within 30 km on the same major reach.
-Discharge data were aggregated by monthly mean from the USGS Washington Water Science Center (collected daily 1978-2007) and the USGS National Water Information System (collected at 15-minute intervals 2008-2015).
+Discharge data were aggregated by monthly mean from the USGS National Water Information System database \citep{USGSdischarge}.
 
@@
 All but snowmelt were available through the U.S.
-Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI)\colorbox{red}{\lstinline{cite}}.
+Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI; \citealt{climateData}).
 We acquired climatic predictor data grouped by Washington State climate division, and all but two of our sites fell within divisions 3 (Puget Sound Lowland) and 4 (East Olympic/Cascade Foothills; see Fig.
@@
 We therefore aggregated these data by monthly mean across the two regions (after verifying their post-standardization similarity), resulting in a single dataset of four climatic predictor variables.
-A snowmelt time series was then added to this dataset, using monthly mean records from six SNOTEL sites (Bumping Ridge, Elbow Lake, Mount Crag, Park Creek Ridge, Stevens Pass, White Pass) listed by the USDA's Natural Resources Conservation Service \colorbox{red}{\lstinline{cite}}.
+A snowmelt time series was then added to this dataset, using monthly mean records from six SNOTEL sites (Bumping Ridge, Elbow Lake, Mount Crag, Park Creek Ridge, Stevens Pass, White Pass) listed by the USDA's Natural Resources Conservation Service; \citealt{snowData}.
 We calculated monthly snowmelt for each site as the absolute value of negative differences in cumulative snow water equivalent from each month to the next.
@@
 \subsection*{Time series analysis}
-Response time series were modeled using dynamic factor analysis (DFA; \colorbox{red}{\lstinline{Zuur et al.
-2003}}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
+Response time series were modeled using dynamic factor analysis (DFA; \citealt{zuur2003estimating}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
 In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error.
 We chose DFA over a traditional multivariate state space approach for two reasons.
-First, it provides advantages in computational efficiency, as a small number of shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost.
+First, it provides advantages in computational efficiency, as a small number of shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost \citep{zuur2003dynamic}.
 Second, in terms of identifying what drives the shared trends, having fewer of them allows for greater inferential parsimony.
 Being a multivariate technique, DFA also provides an advantage over univariate alternatives in that covariance structure among responses can be specified and compared.
-All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software \colorbox{red}{\lstinline{Kristensen et al.
-2015}}, which we called using package TMB in R \colorbox{red}{\lstinline{R Core team 2016...}}.
+All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software \citep{kristensen2015tmb}, which we called using package TMB in R \citep{Rmanual,tmbPackage}.
 
@@
 2).
-\textbf{R} and \textbf{Q} are variance-covariance matrices of order m, and \textbf{Q} is set to identity for model identifiability (\colorbox{red}{\lstinline{Harvey 1989}}).
+\textbf{R} and \textbf{Q} are variance-covariance matrices of order m, and \textbf{Q} is set to identity for model identifiability \citep{harvey1990forecasting}.
 The initial state of the shared trend vector ($\bm{x}_0$) is multivariate-normally distributed with a mean of zero and a diagonal variance-covariance matrix with large variance (e.g.
@@
 We included shared variance and zero covariance, individual variance and zero covariance, shared variance and shared covariance, and unconstrained error.
-Details on these structures and their implications can be found in \colorbox{red}{\lstinline{Holmes et al.
-(2012)}}.
+Details on these structures and their implications can be found in \citep{holmes2012marss}.
 The best models for water temperature and discharge were determined with AIC.
@@
 For post-hoc analyses, monitoring sites were separated into three classes based on relative areal coverage of perennial ice and/snow (hereinafter ``\% glaciation'') and mean elevation across their watersheds.
-The three classes are loosely based on the classification scheme and language of the Climate Impacts Group at the University of Washington \colorbox{red}{\lstinline{cite}}, and are here delineated according to Table 1.
+The three classes are loosely based on the classification scheme and language of the Climate Impacts Group at the University of Washington \citep{mauger2015CIG}, and are here delineated according to Table 1.
 
@@
 To achieve this, we amassed an additional dataset of landscape features.
-These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \colorbox{red}{\lstinline{cite}} and the National Hydrography Dataset (NHDPlusV2)\colorbox{red}{\lstinline{cite}}.
+These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \citep{hill2016stream} and the National Hydrography Dataset (NHDPlusV2; \citealt{mckay2012nhdplus}).
 Each site was mapped to an individual river reach, defined as a segment bounded on each end by a stream or river source, confluence, or mouth.
@@
 Monitoring site elevation was also included.
-Finally, we calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap \colorbox{red}{\lstinline{cite}}.
+Finally, we calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap version 10.4 \citep{arcviewenvironmental}.
 
@@
 RD rivers were consistently warmest throughout the year.
-In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD \colorbox{red}{\lstinline{stats throughout this section}}.
+In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD (Student's t: $p<0.01, F=11.9$).
 RD, meanwhile, only attained a minimum of 5.6\degree C.
@@
 All depicted series represent discrete data.
-\colorbox{red}{\lstinline{see appendix for standard deviations?}} \setlength{\parskip}{6pt}\\[\baselineskip]
+\setlength{\parskip}{6pt}\\[\baselineskip]
+% see appendix for standard deviations?
 
@@
 There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
-3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} \colorbox{red}{\lstinline{see appendix}}.
+3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} (See Appendix A.).
 It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
@@
     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-    \draw (-4.1, -1.5) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+    \draw (-4.1, -2.1) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+Points are sized by watershed area.
 Regression lines indicate slopes significant at $\alpha=0.05$.};
@@
 The Elwha in particular may slip from its current state of high resistance to seasonal climatic changes.
-We tested for changes in mean and variance of $T\textsubscript{air}\rightarrow T\textsubscript{water}$ and $T\textsubscript{air}\rightarrow Q$ coupling between 1978 and 2015, but did not detect any widespread patterns \colorbox{red}{\lstinline{appendix B}}.
+We tested for changes in mean and variance of $T\textsubscript{air}\rightarrow T\textsubscript{water}$ and $T\textsubscript{air}\rightarrow Q$ coupling between 1978 and 2015, but did not detect any regular patterns (Appendix B).
 
@@
 In this way, we determined the most likely landscape drivers of T\textsubscript{water} to be perennial ice and snow cover, mean watershed slope, and groundwater influx.
-In the case of slope, the likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \colorbox{red}{\lstinline{(cite}}; Fig.
+In the case of slope, the likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \citealt{brutsaert1975theory}; Fig.
 3g).
@@
 In the longer term, changes can be expected for rivers that now receive the temperature-buffering effect of glacial runoff.
-Glaciers continue to decline across North America, with the disappearance of all glacial ice in \colorbox{red}{\lstinline{(extent}} projected for the year \colorbox{red}{\lstinline{(year}}.
+Glaciers continue to decline across North America, with glacial ice across Western Canada projected to decline by 70\% from 2005 to 2100 \citep{clarke2015projected}.
 
@@
 \end{document}
-
-% https://wa.water.usgs.gov/data/realtime/adr/interactive/ #discharge daily
-% https://waecy.maps.arcgis.com/apps/Viewer/index.html?appid=832e254169e640fba6e117780e137e7b #Q 15min
-% http://www.horizon-systems.com/NHDPlus/NHDPlusV2_17.php #nhdplus
-% ftp://newftp.epa.gov/EPADataCommons/ORD/NHDPlusLandscapeAttributes/StreamCat/HydroRegions/ #streamca
-% ncdc.noaa.gov/cag/time-series/us #climate data
-% https://www.nrcs.usda.gov/wps/portal/nrcs/detail/or/snow/?cid=nrcs142p2_046350 #snow
-% https://arxiv.org/pdf/1509.00660.pdf #TMB
-% Harvey, A. C. 1989. Forecasting, structural time series models and the Kalman flter. Cambridge University Press, Cambridge, UK.
-%CIG puget sound state of knowledge http://cses.washington.edu/picea/mauger/ps-sok/ps-sok_sec03_watercycle_2015.pdf

### 2017-03-22 02_testing_or_evaluation.R
PCoA, plot with dams

@@ -928,7 +928,10 @@ WsSlope <- read.csv('watershed_data/slope/slope.csv')
 land <- merge(land, WsSlope, by='siteCode', all.x=TRUE)
 
 pcascores <- read.csv('watershed_data/pca_scores.csv')
-land <- merge(land,pcascores, by.x='siteCode', by.y='X')
+land <- merge(land, pcascores, by.x='siteCode', by.y='X')
+
+PCoAscores <- read.csv('watershed_data/pcoa_scores.csv')
+land <- merge(land, PCoAscores, by.x='siteCode', by.y='X')
 
 #choose landscape variables of interest
 landvars <- c('BFIWs','ElevWs','PctImp2006WsRp100',
@@ -937,7 +940,7 @@ landvars <- c('BFIWs','ElevWs','PctImp2006WsRp100',
               'PctUrbMd2011WsRp100','PctUrbHi2011WsRp100',
               'RdDensWsRp100','RunoffWs','OmWs',
               'RckDepWs','WtDepWs','PermWs','PopDen2010Ws',
-              'WsAreaSqKm','WsAreaOver1000','WsSlope','PC1','PC2')
+              'WsAreaSqKm','WsAreaOver1000','WsSlope','PC1','PC2','PCo1','PCo2')
 # landvars <- c('ElevWs','PctImp2006WsRp100',
 #               'PctGlacLakeFineWs','PctAlluvCoastWs','PctIce2011Ws',
 #               'PctCrop2011Ws', 'PctUrbOp2011WsRp100','PctUrbLo2011WsRp100',

### 2017-03-22 06_landvar_pcoa.R

@@ -1,0 +1,80 @@
+#Puget Sound rivers DFA
+#Mike Vlah (vlahm13@gmail.com)
+#created: 5/17/16
+rm(list=ls()); cat('\014') #clear env and console
+
+# 0 - setup and preprocessing ####
+setwd('C:/Users/Mike/git/stream_nuts_DFA')
+
+# load packages and some functions
+library(vegan)
+library(cluster)
+source('http://www.umass.edu/landeco/teaching/ecodata/labs/biostats.R')
+
+#load desired model output and accessories
+# load('1trend.rda')
+# load('2trend.rda')
+# load('1trendNoSeas.rda')
+load('single_trend_exploration/2trendNoSeas.rda')
+
+#add % ice and dams to watershed data
+ice2006 = read.csv('manuscript/figures/PctIce2006Ws.csv', stringsAsFactors=FALSE)
+land = merge(land,ice2006[,2:3], by.x='siteCode', by.y='site')
+land$Ice06_11 = rowMeans(cbind(land$PctIce2006Ws, land$PctIce2011Ws))
+
+dams = read.csv('data/watershed_data/watershed_data_simp.csv', stringsAsFactors=FALSE)[,c('siteCode','dam_upstream')]
+dams$siteCode[dams$siteCode=='AA'] <- 'ZA'
+land = merge(land, dams, by='siteCode')
+
+#compile watershed variables for pca
+ordnames = c('BFIWs','ElevWs','PctImp2006WsRp100','RunoffWs',
+             'WtDepWs','PermWs', 'WsAreaSqKm','WsAreaOver1000','WsSlope','Ice06_11','dam_upstream')
+
+ordlandcols = rep(NA, length(ordnames))
+for(i in 1:length(ordnames)){
+    ordlandcols[i] = which(colnames(land) == ordnames[i])
+}
+
+ordvars = land[,ordlandcols]
+row.names(ordvars) = land$siteCode
+
+#transform proportion and percent data
+ordvars[,'PctImp2006WsRp100'] = asin(sqrt(ordvars[,'PctImp2006WsRp100']/100))
+ordvars[,'WsAreaOver1000'] = asin(sqrt(ordvars[,'WsAreaOver1000']))
+ordvars[,'Ice06_11'] = asin(sqrt(ordvars[,'Ice06_11']/100))
+
+#get Gower's distance (standardization is done internally)
+ordvars$dam_upstream = factor(ordvars$dam_upstream)
+orddist = daisy(ordvars, metric='gower')
+
[…]

### 2017-03-22 manuscript/figures/figscript.R

@@ -43,8 +43,8 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
-# load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-load('temp_due_5m_atpcsn_byMo_allMos.rda')
+load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+# load('temp_due_5m_atpcsn_byMo_allMos.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric
@@ -1226,6 +1226,11 @@ dev.off()
 
 #load discharge_due_5m_atpcsn_byMo_allMos.rda in the setup section, or else names
 #will be screwed up.
+dams = read.csv('../../data/watershed_data/watershed_data_simp.csv', stringsAsFactors=FALSE)[,c('siteCode','dam_upstream')]
+dams$siteCode[dams$siteCode=='AA'] <- 'ZA'
+land = merge(land, dams, by='siteCode')
+
+
 library(plotrix)
 #these are generated in the effect size v month plots
 disch_moInts <- readRDS('../../saved_structures/moInts_discharge_due_5m_atpcsn.rds')
@@ -1253,6 +1258,8 @@ elev_med = which(land$Ice06_11 < 0.7 & land$ElevWs > 600)
 elev_lo = which(land$Ice06_11 < 0.7 & land$ElevWs <= 600)
 cols = rep('black', nrow(land))
 cols[elev_med] = 'gray75'; cols[elev_hi] = 'white'
+dam_col = rep('black',nrow(land))
+dam_col[land$dam_upstream != 0] <- 'darkgoldenrod2'
 for(i in c(0,2,4,6,8,11)){
     if(i == 0) {
     plot(1,1,axes=FALSE,,xlim=c(0,1),ylim=c(0,1),ann=FALSE,type='n')
@@ -1287,7 +1294,7 @@ for(i in c(0,2,4,6,8,11)){
         #                  # pch=4, cex=4, lwd=3, col='steelblue3')
         #                  pch=24, cex=3, lwd=1, col='black', bg='white')
[…]

### 2017-03-23 02_testing_or_evaluation.R
added aspect data

@@ -940,7 +940,7 @@ landvars <- c('BFIWs','ElevWs','PctImp2006WsRp100',
               'PctUrbMd2011WsRp100','PctUrbHi2011WsRp100',
               'RdDensWsRp100','RunoffWs','OmWs',
               'RckDepWs','WtDepWs','PermWs','PopDen2010Ws',
-              'WsAreaSqKm','WsAreaOver1000','WsSlope','PC1','PC2','PCo1','PCo2')
+              'WsAreaSqKm','WsAreaOver1000','WsSlope','PC1','PC2','PCo1','PCo2','aspect')
 # landvars <- c('ElevWs','PctImp2006WsRp100',
 #               'PctGlacLakeFineWs','PctAlluvCoastWs','PctIce2011Ws',
 #               'PctCrop2011Ws', 'PctUrbOp2011WsRp100','PctUrbLo2011WsRp100',

### 2017-03-24 manuscript/manuscript.tex
incorporated aspect and dams and redid plots; prettified PCoA plot

@@
 The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database.
-Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
-Monitoring site elevation was also included.
-Finally, we calculated area over 1000 m (km\textsuperscript{2}) and mean slope (percent rise) by delineating watersheds from a digital elevation model in ArcMap version 10.4 \citep{arcviewenvironmental}.
+Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow coverage (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
+Monitoring site elevation (m) and presence of upstream dams (as full/partial/no damming of upstream mainstem and major tributaries) were also included.
+Finally, we calculated \% area over 1000 m (km\textsuperscript{2}), mean slope (percent rise), and mean aspect (degree) by delineating watersheds from a digital elevation model in ArcMap version 10.4 \citep{arcviewenvironmental}.
 

### 2017-03-24 02_testing_or_evaluation.R

@@ -64,7 +64,7 @@ if (is.null(dev.list()) == TRUE){
 
 # response choices: COND FC NH3_N NO2_NO3 OP_DIS OXYGEN PH PRESS SUSSOL TEMP TP_P TURB
 # also DISCHARGE (from USGS)
-y_choice = 'TEMP'
+y_choice = 'DISCHARGE'
 # cov choices: meantemp meantemp_anom precip precip_anom hydroDrought hydroDrought_anom
 # maxtemp maxtemp_anom hdd hdd_anom, snowmelt
 #if you're looking at effects by month, or by month over time (see 'design' option below),
@@ -88,7 +88,7 @@ method = 'fixed_individual'
 startyr = 1978
 endyr = 2015
 #model params (specific values only relevant for testing, not for parameter optimization loop)
-ntrends = 1
+ntrends = 2
 #error matrix can either hold the MARSS specifications or the TMB ones ('DE', 'DUE', 'EVCV', 'UNC')
 obs_err_var_struc = 'DUE'
 #UPDATE: Mark Schueurell no longer scales his response data. scaling forces the variance of
@@ -1041,7 +1041,7 @@ eff_rescaler <- function(all_cov, seas, scaled=scale){
 rescaled_effect_size <- eff_rescaler(cov_and_seas, cc)
 
 #grab top landscape vars that correlate with effect size, plot and get stats
-# eff_best <- best_landvars(rescaled_effect_size, 6)
+eff_best <- best_landvars(rescaled_effect_size, 6)
 
 #look inside function for details (be sure to change the y axis label to 'D log(resp)/D cov'
 #if you log transformed the response. (note that this may have been done by default in section 1.3
@@ -1102,7 +1102,7 @@ eff_regress_plotter <- function(mode, var=NA, col_scale='ElevWs'){
         }
     }
 }
-eff_regress_plotter('indiv', 'PctIce2011Ws', 'ElevWs')
+eff_regress_plotter('indiv', 'ElevWs', 'ElevWs')
 # eff_regress_plotter('exploration', , 'ElevWs')
 
 # 5.4 - process loading regressions (look inside functions for details) ####

### 2017-03-24 06_landvar_pcoa.R

@@ -12,23 +12,24 @@ library(cluster)
 source('http://www.umass.edu/landeco/teaching/ecodata/labs/biostats.R')
 
 #load desired model output and accessories
-# load('1trend.rda')
-# load('2trend.rda')
-# load('1trendNoSeas.rda')
-load('single_trend_exploration/2trendNoSeas.rda')
+# load('1trend.rda') #doesnt have aspect added to landscape vars
+# load('2trend.rda') #doesnt have aspect added to landscape vars
+# load('1trendNoSeas.rda') #doesnt have aspect added to landscape vars
+load('single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add % ice and dams to watershed data
 ice2006 = read.csv('manuscript/figures/PctIce2006Ws.csv', stringsAsFactors=FALSE)
 land = merge(land,ice2006[,2:3], by.x='siteCode', by.y='site')
 land$Ice06_11 = rowMeans(cbind(land$PctIce2006Ws, land$PctIce2011Ws))
 
-dams = read.csv('data/watershed_data/watershed_data_simp.csv', stringsAsFactors=FALSE)[,c('siteCode','dam_upstream')]
-dams$siteCode[dams$siteCode=='AA'] <- 'ZA'
-land = merge(land, dams, by='siteCode')
+# dams = read.csv('data/watershed_data/watershed_data_simp.csv', stringsAsFactors=FALSE)[,c('siteCode','dam_upstream')]
+# dams$siteCode[dams$siteCode=='AA'] <- 'ZA'
+# land = merge(land, dams, by='siteCode')
 
 #compile watershed variables for pca
 ordnames = c('BFIWs','ElevWs','PctImp2006WsRp100','RunoffWs',
-             'WtDepWs','PermWs', 'WsAreaSqKm','WsAreaOver1000','WsSlope','Ice06_11','dam_upstream')
+             'WtDepWs','PermWs', 'WsAreaSqKm','WsAreaOver1000','WsSlope',
+             'Ice06_11','dam_upstream','aspect')
 
 ordlandcols = rep(NA, length(ordnames))
 for(i in 1:length(ordnames)){
@@ -38,13 +39,15 @@ for(i in 1:length(ordnames)){
 ordvars = land[,ordlandcols]
 row.names(ordvars) = land$siteCode
 
-#transform proportion and percent data
[…]

### 2017-03-24 manuscript/figures/figscript.R

@@ -45,6 +45,7 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
 load('discharge_due_5m_atpcsn_byMo_allMos.rda')
 # load('temp_due_5m_atpcsn_byMo_allMos.rda')
+# load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric
@@ -189,8 +190,8 @@ for(covr in 1:3){
     print.letter(paste(letters[covr],sig), c(.9,.9), cex=1.8, font=2, col='steelblue')
     points(landvar, res,
            bg=cols, col='black', cex=rescale(log(land$WsAreaSqKm),c(1.2,3.2)),
-           # pch=21,
-           pch=land$siteCode,
+           pch=21,
+           # pch=land$siteCode,
            lwd=1)
     # color.legend(xl=4,xr=4.4,yb=0.5, yt=0.6, legend=c('147', '1349'),
     #              rect.col=colorRampPalette(c('brown', 'white'))(10),
@@ -307,6 +308,185 @@ par(defpar)
 
 #for results (loadings)
 summary(mod)
+
+# 1.3 - TEMP eff size and loadings with PCs ####
+land = readRDS('../../saved_structures/2trendNoSeasNoSnow_land.rds')
+dam_col = rep('black',nrow(land))
+dam_col[land$dam_upstream != 0] <- 'chocolate2'
+
+dfa = readRDS('../../saved_structures/full_dfaOut.rds')
+land = readRDS('../../saved_structures/full_land.rds')
+
+# system('taskkill /f /im AcroRd32.exe')
+pdf('19_temp_simp_reg.pdf', width=7.5, height=5)
+# layout(matrix(c(1:6,9,7,8),nrow=3,byrow=TRUE))
+defpar = par(mfrow=c(2,3), oma=c(0,0,1,1), mar=c(4,3.5,0,0))
+landvar=land$ElevWs
+# landvar=land$ElevWs/100
+land_sub <- land[,landcols] #subset landscape variables by those used in the analysis
+#% of watershed area classified as ice/snow land cover (NLCD 2011 class 12)
[…]

### 2017-03-27 manuscript/manuscript.tex
updating methods and results

@@
 % todo:
-% include snowmelt in model run. see if it's worth including with precip and temp as a climate predictor
-% introduce terminology:
-% responses = water temp, Q
-% climatic predictors = air temp, precip, drought
-% landscape predictors = many
 %include hydrograph figures from kathy liederman and julian olden to show
     %rain dom, rain and snow, snow dom
-%include the over-time plots in the supplemental information
-%in regression figure, line up left side of legend box with plot above
-    %make gray horiz line go only to topright corner of box
-    %make axis lines and plot borders cleaner (right now axis segments are darker)
-    %instead of elevation on x, use ice and color by elev (orange to blue) X no story
-    %for loadings, color by ice and make it darkgreen , midgreen, white X not without the above
-    % add % ice as one of the hidden trends
-%discharge regression combo plot is scrapped. no relationship with snowmelt?
-%discharge only really relevant in the Q v temp plot.
-%increase unit text size in fig 2
-%change area over 1000m to % area over 1000m and replot
 %change "dominated" to "dominant"
 %change "fall" to "autumn"
-%dont size points in figure 2. include size in legend in fig 3
-%adjust letter position in corners of fig 2 (anchoring)
 %see if its possible to directly compare Z and D
@@
 % can i say how much elevation would inflence temperature independent of snowmelt?
-%examine change over time in just the elwha
-%%%%why am i removing additional seasonal variation (fixed factor)???
-%probably get snowmelt out of the picture? it may be misleading and a source of criticism
-%explain why some SD rivers might have a larger effect than RD (related to slope, groundwater?)
-%get snowmelt out of there. it's a watershed trait. not a regional one
+%examine change over time in just the elwha (mike will ask about this)
+%explain why some SD rivers might have a larger effect than RD (related to slope, groundwater, dams?)
 %add discharge locations to the map?
 %probably should include a table of data sources and station ids
+% make sure figure captions are all formatted appropriately
+% look at differences in elwha pre and post dam
 
@@
 
-is the covariate matrix structure necessary to account for exogenous seasonal variation (rows 1-12), and overall effect of precipitation (row 13), while also yielding the by-month effect of air temperature (rows 13-24) on the response (\textbf{y}).
+\noindent
+is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and overall effect of precipitation (row 13), while also yielding the by-month effect of air temperature (rows 13-24) on the response (\textbf{y}; Eq.
+2).
 
-Additional, non seasonal variation due to exogenous effects loads onto the shared trends, and a portion of remaining variation is absorbed by error matrix \textbf{v}.
-We fit models using four unique error structures (\textbf{R}), to allow for different suites of unknown drivers affecting rivers.
-We included shared variance and zero covariance, individual variance and zero covariance, shared variance and shared covariance, and unconstrained error.
+Additional, non seasonal variation due to unknown factors manifests in the shared trends, and a portion of any residual variation is absorbed by error matrix \textbf{v}.
+We fit models using four unique error structures (\textbf{R}), to allow for multiple suites of unknown drivers affecting rivers.
+We included shared variance with zero covariance, individual variance with zero covariance, shared variance with shared covariance, and individual variance with individual covariance.
 Details on these structures and their implications can be found in \citep{holmes2012marss}.
-The best models for water temperature and discharge were determined with AIC.
+The best models for water temperature and discharge were determined via AIC.
+However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, rendering untrustworthy common rules of thumb for accepting additional parameters under AIC \citep{burnham2003model}.
+Thus, we had reason to doubt that the "most parsimonious" model according to AIC alone was any better than a much simpler alternative.
+To manage this, we required that each additional trend, covariate, or seasonal structure improve the median coefficient of determination (R\textsuperscript{2}) by at least 1\% in order to justify accepting the added complexity.
 
 \subsection*{Landscape predictors and post-hoc regression}
-
 For post-hoc analyses, monitoring sites were separated into three classes based on relative areal coverage of perennial ice and/snow (hereinafter ``\% glaciation'') and mean elevation across their watersheds.
@@
 
-After model selection, climatic predictor effect sizes for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling.
+After model selection, climatic predictor effect sizes (\textbf{D}; Eq.
+2) for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling.
 To achieve this, we amassed an additional dataset of landscape features.
@@
 Monitoring site elevation (m) and presence of upstream dams (as full/partial/no damming of upstream mainstem and major tributaries) were also included.
-Finally, we calculated \% area over 1000 m (km\textsuperscript{2}), mean slope (percent rise), and mean aspect (degree) by delineating watersheds from a digital elevation model in ArcMap version 10.4 \citep{arcviewenvironmental}.
+Finally, we calculated area above 1000 m (as \% watershed area), mean slope (\% rise), and mean aspect (degree) by delineating and summarizing watersheds from a digital elevation model in ArcMap v.
+10.4 \citep{arcviewenvironmental}.
 
-\section*{Results}
+An additional set of post-hoc regressions was performed using factor loadings on shared trends (\textbf{Z}; Eq.2) as dependent variables, with landscape predictors again as independent variables.
+Here, the goal was to %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+To facilitate this, we used a reduced version of the full model.
+
+\section*{Results} %%
 
@@
 In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD (Student's t: $p<0.01, F=11.9$).
-RD, meanwhile, only attained a minimum of 5.6\degree C.
+RD only attained a minimum of 5.6\degree C.
 RS reached a peak summer temperature of 16.9\degree C in July, while RS and SD followed in August with peak temperatures of 15.5 and 13.5\degree C, respectively.
@@
 
-The combined hydrograph of all rivers reveals two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
-Spring peak discharge coincides noticeably with a separation in water temperature between SD and RS, while the summer trough coincides with separation of RD and T\textsubscript{air}.
-November marks both the autumn peak in discharge and the point at which T\textsubscript{air} falls below T\textsubscript{water}.
+The combined hydrograph of all rivers revealed two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
+Spring peak discharge coincided noticeably with a separation in water temperature between SD and RS, while the summer trough coincided with separation of RD and T\textsubscript{air}.
+On average, November marked both the autumn peak in discharge and the point at which T\textsubscript{air} fell below T\textsubscript{water}.
 
-\sloppy
-DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with greater watershed elevation ($p=0.04, R^2=0.18$; Fig.
+There was also an apparent divergence in slope between RD and all snow-influenced rivers, beginning in early spring and culminating in June.
+Between June and July, RS and SD saw a large jump in temperature, which coincided with the decline in snowmelt.
+
+DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with increasing watershed elevation ($p=0.04, R^2=0.18$; Fig.
 3a).
@@
 
-\begin{tikzpicture}
-    \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-    \draw (-4.1, -2.1) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+% \begin{tikzpicture}
+%     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
+%     \draw (-4.1, -2.1) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
 Points are sized by watershed area.
 Regression lines indicate slopes significant at $\alpha=0.05$.};
-\end{tikzpicture}
+% \end{tikzpicture}
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/19_temp_simp_reg.pdf}}
+\textbf{Figure 4} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and between watershed features and factor loadings on shared trends \textbf{(d-e)}.
+Regression lines indicate slopes significant at $\alpha=0.1$.
+\end{center}
 

### 2017-03-27 manuscript/figures/figscript.R

@@ -43,8 +43,8 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
-load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-# load('temp_due_5m_atpcsn_byMo_allMos.rda')
+# load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+load('temp_due_5m_atpcsn_byMo_allMos.rda')
 # load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
@@ -310,9 +310,14 @@ par(defpar)
 summary(mod)
 
 # 1.3 - TEMP eff size and loadings with PCs ####
+
+#load the full atpcsn model?
+
 land = readRDS('../../saved_structures/2trendNoSeasNoSnow_land.rds')
-dam_col = rep('black',nrow(land))
-dam_col[land$dam_upstream != 0] <- 'chocolate2'
+# dam_col = rep('black',nrow(land))
+# dam_col[land$dam_upstream != 0] <- 'chocolate2'
+dam_pch = rep(FALSE,nrow(land))
+dam_pch[land$dam_upstream != 0] <- TRUE
 
 dfa = readRDS('../../saved_structures/full_dfaOut.rds')
 land = readRDS('../../saved_structures/full_land.rds')
@@ -372,11 +377,12 @@ for(covr in 1:3){
     }
     print.letter(paste(letters[covr],sig), c(.9,.9), cex=1.8, font=2, col='steelblue')
     points(landvar, res,
-           bg=cols, col=dam_col, cex=2,
+           bg=cols, col='black', cex=2,
            # bg=cols, col='black', cex=rescale(log(land$WsAreaSqKm),c(1.2,3.2)),
            pch=21,
            # pch=land$siteCode,
            lwd=1)
+    points(landvar[dam_pch], res[dam_pch], col='chocolate2', cex=2, lwd=1, pch=124)
     # color.legend(xl=4,xr=4.4,yb=0.5, yt=0.6, legend=c('147', '1349'),
[…]

### 2017-03-30 manuscript/manuscript.tex
figure fixes, some writing, accidentally remade a bunch of old figures

@@
 
-Because we were interested in isolating the effects of climatic predictors on river temperature and discharge, we used fixed factors to absorb recurring seasonal variation not related to the predictors, with one factor level for each month.
-These factors were incorporated into the covariate matrix (\textbf{d}).
-Thus, the coefficient in \textbf{D} relating, say, air temperature (predictor) and water temperature (response), represents the effect size of the former on the latter.
-In other words, it is the change in water temperature accompanying a unit change in air temperature across the whole time series.
-We call this relationship "coupling." We were also interested in coupling by month for specific predictors, which required that the focal predictor in a particular model be arranged seasonally.
+Because we were interested in isolating the effects of climatic predictors on river temperature and discharge, we used a fixed factor to account for recurring seasonal variation not related to the predictors, with one factor level for each month.
+This factor was incorporated into the covariate matrix (\textbf{d}).
+Thus, the coefficient in \textbf{D} relating, say, precipitation (predictor) and water temperature (response), represents the effect size of the former on the latter.
+In other words, it is the change in water temperature accompanying a unit change in precipitation across the whole time series.
+We call this relationship "coupling." We were also interested in coupling by month for T\textsubscript{air}, which required that it be arranged as twelve separate, monthly time series.
 Concretely,
@@
     12 & 0 & 0 & 0 & \cdots & 1 \\
-    13 & \theta_{precip}^{(1)} & \theta_{precip}^{(2)} & \theta_{precip}^{(3)} & \cdots & \theta_{precip}^{(T)} \\
-    14 & \theta_{air}^{(1)} & 0 & 0 & \cdots & 0 \\
-    15 & 0 & \theta_{air}^{(2)} & 0 & \cdots & 0 \\
-    16 & 0 & 0 & \theta_{air}^{(3)} & \cdots & 0 \\
+    13 & precip_1 & precip_2 & precip_3 & \cdots & precip_T \\
+    14 & snowmelt_1 & snowmelt_2 & snowmelt_3 & \cdots & snowmelt_T \\
+    15 & air_1 & 0 & 0 & \cdots & 0 \\
+    16 & 0 & air_2 & 0 & \cdots & 0 \\
+    17 & 0 & 0 & air_3 & \cdots & 0 \\
       & \vdots & \vdots & \vdots & \ddots & \vdots \\
-    25 & 0 & 0 & 0 & \cdots & \theta_{air}^{(T)} \\
+    26 & 0 & 0 & 0 & \cdots & air_T \\
 \end{block}
@@
 \noindent
-is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and overall effect of precipitation (row 13), while also yielding the by-month effect of air temperature (rows 13-24) on the response (\textbf{y}; Eq.
+is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and the effects of precipitation (row 13) and snowmelt (row 14), while also yielding the effect of air temperature by month (rows 15-26) on the response (\textbf{y}; Eq.
 2).
+This is the covariate structure of the model we used for subsequent analyses, not including those described in Figure 5d-e, and Appendix B.
 
@@
 The best models for water temperature and discharge were determined via AIC.
-However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, rendering untrustworthy common rules of thumb for accepting additional parameters under AIC \citep{burnham2003model}.
+However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, undermining common rules of thumb for admitting additional parameters under AIC \citep{burnham2003model}.
 Thus, we had reason to doubt that the "most parsimonious" model according to AIC alone was any better than a much simpler alternative.
-To manage this, we required that each additional trend, covariate, or seasonal structure improve the median coefficient of determination (R\textsuperscript{2}) by at least 1\% in order to justify accepting the added complexity.
+To manage this, we required that each additional trend, covariate, or seasonal structure improve the median coefficient of determination (R\textsuperscript{2}) by at least 1\% in order to justify accepting its attendant complexity.
 
@@
 An additional set of post-hoc regressions was performed using factor loadings on shared trends (\textbf{Z}; Eq.2) as dependent variables, with landscape predictors again as independent variables.
-Here, the goal was to %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To facilitate this, we used a reduced version of the full model.
+Loadings represent the degree to which each river's temperature fluctuates with the anonymous force driving the corresponding shared trend.
+A landscape feature that varies in proportion to these loadings is therefore likely to be a mediator of the anonymous force, if not the force itself.
+To facilitate inference by way of the shared trends, we made three simplifications to the model.
+We removed the monthly factor and the snowmelt predictor from the covariate matrix (\textbf{d}, rows 1-12 and 14), so that the trends would be free to express seasonal and elevational variation.
+Then, we  limited the number of trends to between one and three, to avoid "trend specialization." in other words, we optimized the trends for flexibility while concentrating their explanatory power.
+Additionally, we summarized the landscape predictors with principal coordinates analysis, as a way to conceptually "group" them by correlation.
+We used the Gower dissimilarity coefficient (Gower's distance) to account for association among both continuous and nominal variables \citep{gowerDist}.
 
@@
 
+\noindent
 \textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
 All depicted series represent discrete data.
-\setlength{\parskip}{6pt}\\[\baselineskip]
+%\setlength{\parskip}{6pt}
+\\[\baselineskip]
 % see appendix for standard deviations?
@@
 
-DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with increasing watershed elevation ($p=0.04, R^2=0.18$; Fig.
-3a).
-On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a $0.45\pm 0.17$\degree C change at SD sites.
-A similar trend was observed with respect to $Precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig.
-3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
-There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
-3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} (See Appendix A.).
-It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $Precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
-Both rivers drain glaciers of the Olympic Mountain Range, and both are SD, though the Elwha's watershed is larger.
-
 % \begin{tikzpicture}
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/19_temp_simp_reg.pdf}}
-\textbf{Figure 4} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and between watershed features and factor loadings on shared trends \textbf{(d-e)}.
+% \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and between watershed features and factor loadings on shared trends \textbf{(d-e)}.
 Regression lines indicate slopes significant at $\alpha=0.1$.
+\resizebox{\textwidth}{!}{\includegraphics{figures/18_PCoA.pdf}}
 \end{center}
+\textbf{Figure 3} (a-c) Relationships between watershed elevation and climatic effects on T\textsubscript{water}, and (d-e) between watershed features and factor loadings on shared trends.
+Regression lines indicate slopes significant at $\alpha=0.1$.
+(f) Ordination of landscape predictors by principal coordinates analysis.
+Length and direction of arrows are proportional to loading of landscape predictors onto each principal axis of their variation.
+% Dotted arrow indicates lack of association with either axis shown
+\\[\baselineskip]
+
+DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with increasing watershed elevation ($p=0.04, R^2=0.18$; Fig.
+3a).
+On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a $0.45\pm 0.17$\degree C change at SD sites.
+A similar trend was observed with respect to $precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig.
+3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
+There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
+3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} (See Appendix A.).
+It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
+Both rivers drain glaciers of the Olympic Mountain Range, and both are SD, though the Elwha's watershed is larger.
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
+\end{center}
 \textbf{Figure 4} Relationship between $T_{air}\rightarrow T_{water}$ and $T_{air}\rightarrow Q$.
@@
 Regression lines indicate slopes significant at $\alpha=0.05$.
-\end{center}
+\\[\baselineskip]
 

### 2017-03-30 06_landvar_pcoa.R

@@ -16,6 +16,12 @@ source('http://www.umass.edu/landeco/teaching/ecodata/labs/biostats.R')
 # load('2trend.rda') #doesnt have aspect added to landscape vars
 # load('1trendNoSeas.rda') #doesnt have aspect added to landscape vars
 load('single_trend_exploration/2trendNoSeasNoSnow.rda')
+print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
+    tmp <- par("usr")
+    text.x <- tmp[1]+xy[1]*diff(tmp[1:2])   #x position, diff=difference
+    text.y <- tmp[3]+xy[2]*diff(tmp[3:4])   #y position
+    text(x=text.x, y=text.y, labels=label, ...)
+}
 
 #add % ice and dams to watershed data
 ice2006 = read.csv('manuscript/figures/PctIce2006Ws.csv', stringsAsFactors=FALSE)
@@ -53,7 +59,7 @@ orddist = daisy(ordvars, metric='gower')
 # 1 - run PCoA ####
 
 #perform pcoa
-ord_out = cmdscale(orddist, eig=TRUE, add=FALSE)
+ord_out = cmdscale(orddist, k=2, eig=TRUE, add=FALSE)
 #get scores (PCs) and eigenvalues
 scores = ord_out$points
 colnames(scores) = c('PCo1','PCo2')
@@ -78,23 +84,56 @@ row.names(ord_load$vectors$arrows) <- c('BFI','elev.','imperv.','runoff','water 
 row.names(ord_load$factors$centroids) <- c('','','(dam-influenced)')
 # row.names(ord_load$factors$centroids) <- c('','','')
 
-#plot pcoa
+#plot pcoa (2 axes)
 pdf('manuscript/figures/18_PCoA.pdf', width=5, height=5)
+
+dam_pch = rep(FALSE,nrow(land))
+dam_pch[land$dam_upstream != 0] <- TRUE
+
 elev_hi = which(land$Ice06_11 >= 0.7)
 elev_med = which(land$Ice06_11 < 0.7 & land$ElevWs > 600)
 elev_lo = which(land$Ice06_11 < 0.7 & land$ElevWs <= 600)
 cols = rep('black', nrow(land))
 cols[elev_med] = 'gray60'; cols[elev_hi] = 'white'
-dam_col = rep('black',nrow(land))
[…]

### 2017-03-30 manuscript/figures/figscript.R

@@ -311,7 +311,7 @@ summary(mod)
 
 # 1.3 - TEMP eff size and loadings with PCs ####
 
-#load the full atpcsn model?
+#load the full atpcsn model
 
 land = readRDS('../../saved_structures/2trendNoSeasNoSnow_land.rds')
 # dam_col = rep('black',nrow(land))
@@ -324,6 +324,7 @@ land = readRDS('../../saved_structures/full_land.rds')
 
 # system('taskkill /f /im AcroRd32.exe')
 pdf('19_temp_simp_reg.pdf', width=7.5, height=5)
+# pdf('20_temp_simp_reg_NAMES.pdf', width=7.5, height=5)
 # layout(matrix(c(1:6,9,7,8),nrow=3,byrow=TRUE))
 defpar = par(mfrow=c(2,3), oma=c(0,0,1,1), mar=c(4,3.5,0,0))
 landvar=land$ElevWs
@@ -454,23 +455,23 @@ land_sub <- land[,landcols] #subset landscape variables by those used in the ana
 # elev_lo = which(land$Ice06_11 < 0.7 & land$ElevWs <= 600)
 # cols = rep('black', nrow(land))
 # cols[elev_med] = 'gray60'; cols[elev_hi] = 'white'
-for(trnd in c(1,2)){
+for(trnd in c(2,1)){
     plot(landvar[[trnd]], dfa$Estimates$Z[,trnd], type='n', yaxt='n',
          xlab='', main='', ylab='', cex.lab=1, font=2, xaxt='n',bty='l')
     mod <- lm(dfa$Estimates$Z[,trnd] ~ landvar[[trnd]])
     sig <- ifelse(summary(mod)$coefficients[2,4]<=0.1, '*', '')
     sig <- NULL #turning off sig plotting
-    if(trnd == 1){
-        mtext(bquote(textstyle(bold('PC 1 '))*textstyle(plain('(50.8%)'))), 1, cex=.8, font=2, line=2)
-        print.letter(paste('d',sig),c(.9,.9), cex=1.8, font=2, col='springgreen4')
+    if(trnd == 2){
+        mtext(bquote(textstyle(bold('Principal axis 1 '))*textstyle(plain('(50.8%)'))), 1, cex=.8, font=2, line=2)
+        print.letter(paste('d',sig),c(.1,.9), cex=1.8, font=2, col='springgreen4')
         mtext(bquote(bold('Shared trend loadings')~
[…]

### 2017-03-31 manuscript/manuscript.tex
couple more changes to figs and ms. editing discussion now

@@
 % look at differences in elwha pre and post dam
+%mention palmer drought index?
+% find and replace "water temperature" with T_water, same for air
 
 \maketitle
-\vspace{10pc}
-\noindent
-\textbf{Summary:}
+\clearpage
 
-\noindent
-5 data sources; watershed delineation; 2 DFAs: m=1:15, 7 covariate combinations, 4 err. struc., 2 seasonality models; TMB; parallel computing; Bayesian change/time
-\clearpage
 \pagenumbering{arabic}
@@
 We found that primarily rain-fed rivers experience large seasonal temperature fluctuations, which closely track atmospheric temperature, while snow-fed rivers tend to be weakly, and in some cases inversely, coupled with such fluctuations.
-Among watersheds, groundwater influx, land slope, and discharge further augment or dampen these relationships.
+Among watersheds, groundwater influx, land slope, and Q further augment or dampen these relationships.
 Our results suggest the temperature of high-elevation rivers, absent the influence of ice, should be highly variable, and that glacially fed streams stand to see the largest changes in temperature regime under proposed climate scenarios.
@@
 Adding to this picture, flow regimes across rivers of the Puget Sound watershed vary with latitude and elevation \citep{reidy2012hydrogeomorphic,mauger2015CIG}, and can be classified broadly into three categories by flow source and hydrograph shape.
-Rain-dominated (RD) rivers receive little or no input from snowmelt, and thus peak in discharge during the rainy season, usually between October and February.
+Rain-dominated (RD) rivers receive little or no input from snowmelt, and thus peak in discharge (Q) during the rainy season, usually between October and February.
 Snow-dominated (SD) rivers instead see peak flow during spring snowmelt, often in April, May, or June.
@@
 Our second aim was to identify watershed features that correlate with such departures, and thus provide a nuanced basis for predicting impacts of water temperature on aquatic biodiversity and fishery health.
-We hypothesized that water temperature would track air temperature most closely in RD rivers \citep{ward1985thermal,garner2014river}.
-We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt \citep{lisi2015watershed} and with factors affecting heat capacity of water, including discharge (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing; \citealt{van2013global}).
+We hypothesized that water temperature (T\textsubscript{water} would track air temperature T\textsubscript{air} most closely in RD rivers \citep{ward1985thermal,garner2014river}.
+We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt \citep{lisi2015watershed} and with factors affecting heat capacity of water, including Q (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing; \citealt{van2013global}).
 
@@
 
-We investigated climate and landscape controls on water temperature and discharge, as separate response variables, from 1978 to 2015.
+We investigated climate and landscape controls on T\textsubscript{water} and Q, as separate response variables, from 1978 to 2015.
 Monthly time series of water temperature were obtained for 24 river sites via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \citep{DoEwaterData}.
 These sites represent 19 nonnested watersheds across 9 counties, and range from 4 to 775 m in elevation.
-For at least one site at each river, monthly discharge time series were also available, either from the same location as one of the temperature monitoring sites, or from within 30 km on the same major reach.
-Discharge data were aggregated by monthly mean from the USGS National Water Information System database \citep{USGSdischarge}.
+For at least one site at each river, monthly Q time series were also available, either from the same location as one of the temperature monitoring sites, or from within 30 km on the same major reach.
+Q data were aggregated by monthly mean from the USGS National Water Information System database \citep{USGSdischarge}.
 
@@
 
-Potential climatic predictors of water temperature and discharge included mean and max air temperature (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index), averaged by month across the response variable time series.
+Potential climatic predictors of T\textsubscript{water} and Q included mean and max T\textsubscript{air} (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index), averaged by month across the response variable time series.
 All but snowmelt were available through the U.S.
@@
 \subsection*{Time series analysis}
-Response time series were modeled using dynamic factor analysis (DFA; \citealt{zuur2003estimating}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
+Response time series (T\textsubscript{water} and Q) were modeled using dynamic factor analysis (DFA; \citealt{zuur2003estimating}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
 In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error.
@@
 
-Because we were interested in isolating the effects of climatic predictors on river temperature and discharge, we used a fixed factor to account for recurring seasonal variation not related to the predictors, with one factor level for each month.
+Because we were interested in isolating the effects of climatic predictors on T\textsubscript{water} and Q, we used a fixed factor to account for recurring seasonal variation not related to the predictors, with one factor level for each month.
 This factor was incorporated into the covariate matrix (\textbf{d}).
-Thus, the coefficient in \textbf{D} relating, say, precipitation (predictor) and water temperature (response), represents the effect size of the former on the latter.
+Thus, the coefficient in \textbf{D} relating, say, precipitation (predictor) and T\textsubscript{water} (response), represents the effect size of the former on the latter.
 In other words, it is the change in water temperature accompanying a unit change in precipitation across the whole time series.
@@
 \noindent
-is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and the effects of precipitation (row 13) and snowmelt (row 14), while also yielding the effect of air temperature by month (rows 15-26) on the response (\textbf{y}; Eq.
+is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and the effects of precipitation (row 13) and snowmelt (row 14), while also yielding the effect of T\textsubscript{air} by month (rows 15-26) on the response (\textbf{y}; Eq.
 2).
-This is the covariate structure of the model we used for subsequent analyses, not including those described in Figure 5d-e, and Appendix B.
+This is the covariate structure of the T\textsubscript{water} model we used for subsequent analyses, not including those described in Figure 5d-e, and Appendix B.
+The same form was used for the Q model.
 
@@
 Details on these structures and their implications can be found in \citep{holmes2012marss}.
-The best models for water temperature and discharge were determined via AIC.
+The best models for T\textsubscript{water} and Q were determined via AIC.
 However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, undermining common rules of thumb for admitting additional parameters under AIC \citep{burnham2003model}.
@@
 Then, we  limited the number of trends to between one and three, to avoid "trend specialization." in other words, we optimized the trends for flexibility while concentrating their explanatory power.
-Additionally, we summarized the landscape predictors with principal coordinates analysis, as a way to conceptually "group" them by correlation.
+Additionally, we summarized the landscape predictors with principal coordinates analysis (PCoA), as a way to conceptually "group" them by correlation.
 We used the Gower dissimilarity coefficient (Gower's distance) to account for association among both continuous and nominal variables \citep{gowerDist}.
@@
 The combined hydrograph of all rivers revealed two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
-Spring peak discharge coincided noticeably with a separation in water temperature between SD and RS, while the summer trough coincided with separation of RD and T\textsubscript{air}.
-On average, November marked both the autumn peak in discharge and the point at which T\textsubscript{air} fell below T\textsubscript{water}.
+Spring peak Q coincided noticeably with a separation in water temperature between SD and RS, while the summer trough coincided with separation of RD and T\textsubscript{air}.
+On average, November marked both the autumn peak in Q and the point at which T\textsubscript{air} fell below T\textsubscript{water}.
 
@@
 \end{center}
-\textbf{Figure 3} (a-c) Relationships between watershed elevation and climatic effects on T\textsubscript{water}, and (d-e) between watershed features and factor loadings on shared trends.
+\textbf{Figure 3} (a-c) Relationships between watershed elevation and climatic effects on T\textsubscript{water}, obtained from full model fit.
+(d-e) Relationships between watershed features and factor loadings on shared trends, from constrained model fit.
 Regression lines indicate slopes significant at $\alpha=0.1$.
@@
 
-DFA results, aggregated across months and years for each site, revealed a trend toward reduced $T_{air}\rightarrow T_{water}$ coupling with increasing watershed elevation ($p=0.04, R^2=0.18$; Fig.
+DFA results, aggregated across months and years for each site, revealed a trend toward reduced $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ coupling with increasing watershed elevation ($p=0.04, \textrm{mult.} R^2=0.18$; Fig.
 3a).
 On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a $0.45\pm 0.17$\degree C change at SD sites.
-A similar trend was observed with respect to $precip\rightarrow T_{water}$ coupling ($p=0.03, R^2=0.21$; Fig.
+A similar trend was observed with respect to $\textrm{precip}\rightarrow \textrm{T}_{water}$ coupling ($p=0.03, \textrm{mult.} R^2=0.21$; Fig.
 3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
-There was no evidence of coupling between snowmelt and T\textsubscript{water} (Fig.
+There was no evidence of coupling overall between snowmelt and T\textsubscript{water} (Fig.
 3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} (See Appendix A.).
-It is of note that the strongest examples of $T_{air}\rightarrow T_{water}$ and $precip\rightarrow T_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
-Both rivers drain glaciers of the Olympic Mountain Range, and both are SD, though the Elwha's watershed is larger.
+The strongest examples of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip}\rightarrow \textrm{T}_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
+Both rivers drain glaciers of the Olympic Mountain Range, and both are SD.
+Among SD rivers, those influenced by dams appear to couple less strongly with T\textsubscript{air} and precip, but more so with snowmelt.
 
-In addition to the three climate predictors above, the best T\textsubscript{water} model also included five shared trends.
+% In addition to the three climate predictors above, the best T\textsubscript{water} model also included five shared trends.
 Of these, four correlated significantly with at least one known watershed predictor.
 Figure 3 depicts the strongest correlated variables with each trend (insets d-e).
-These are, in arbitrary order of relevance, mean water table depth ($p<0.001, R^2=0.60$; Fig.
-3d), \% glaciation ($p<0.01, R^2=0.30$; Fig.
-3e), BFI ($p=0.01, R^2=0.25$; Fig.
-3f), and mean slope ($p<0.01, R^2=0.29$; Fig.
+These are, in arbitrary order, mean water table depth ($p<0.001, \textrm{mult.} R^2=0.60$; Fig.
+3d), \% glaciation ($p<0.01, \textrm{mult.} R^2=0.30$; Fig.
+3e), BFI ($p=0.01, \textrm{mult.} R^2=0.25$; Fig.
+3f), and mean slope ($p<0.01, \textrm{mult.} R^2=0.29$; Fig.
 3g).
 The fifth shared trend was not correlated with any variables in the watershed predictor dataset.
+Factor loadings from a constrained, two-trend model each correlated with one of the two dominant, principal axes of variation across landscape predictors, determined by PCoA (Fig.
+3f).
+The first principal axis was driven by mean watershed slope, snow (\% area > 1000 m) and ice, soil permeability, and other features that vary along elevational gradients, as well as mean elevation itself.
+Watershed's scores along this axis correlated with loadings from one trend, with marginal significance ($p=0.07, \textrm{mult.} R^2=0.14$; Fig.
+3d).
+The second principal axis was driven by runoff, base flow, and upstream dams, and correlated with the other trend's loadings ($p<0.01, \textrm{mult.} R^2=0.35$; Fig.
+3e).
+Combined, the first two principal axes accounted for 79.4\% of variation across landscape predictors.
 
@@
 \end{center}
-\textbf{Figure 4} Relationship between $T_{air}\rightarrow T_{water}$ and $T_{air}\rightarrow Q$.
-Both axes are expressed per 1\degree C change in $T_{air}$.
-The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $T_{air}$.
+\textbf{Figure 4} Relationship between $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{T}_{air}\rightarrow \textrm{Q}$.
+Both axes are expressed per 1\degree C change in $\textrm{T}_{air}$.
+The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $\textrm{T}_{air}$.
 Regression lines indicate slopes significant at $\alpha=0.05$.
@@
 In both models, T\textsubscript{air} was allowed to have unique monthly effects.
-These effects, taken together, can be understood in relation to the four quadrants of the Cartesian coordinate system (increasing clockwise from upper right; Fig.
+These effects, taken together, can be conceptualized in relation to the four quadrants of the Cartesian coordinate system (increasing clockwise from upper right; Fig.
 4).
 
-In mid-winter (exemplified by February), all river classes primarily occupy the first quadrant, signifying $T\textsubscript{air} \propto T\textsubscript{water}$ and $T\textsubscript{air} \propto Q$, where $\propto$ denotes proportionality.
+In mid-winter (exemplified by February), all river classes primarily occupy the first quadrant, signifying $\textrm{T}\textsubscript{air} \propto \textrm{T}\textsubscript{water}$ and $\textrm{T}\textsubscript{air} \propto \textrm{Q}$, where $\propto$ denotes proportionality.
 RD shows the weakest Q response.
-By spring, many RS and SD sites develop an inverse relationship between T\textsubscript{air} and T\textsubscript{water}, denoted $T\textsubscript{air} \propto \frac{1}{T\textsubscript{water}}$, while RD sites change little from their winter state.
+By spring, many RS and SD sites develop an inverse relationship between T\textsubscript{air} and T\textsubscript{water}, denoted $\textrm{T}\textsubscript{air} \propto \frac{1}{\textrm{T}\textsubscript{water}}$, while RD sites change little from their winter state.
 June and August see a procession of most sites into the near fourth quadrant, with SD trailing.
-This signifies $T\textsubscript{air} \propto \frac{1}{Q}$, though $T\textsubscript{air} \propto T\textsubscript{water}$ remains.
+This signifies $\textrm{T}\textsubscript{air} \propto \frac{1}{\textrm{Q}}$, though $\textrm{T}\textsubscript{air} \propto \textrm{T}\textsubscript{water}$.
 One stark exception is again the Elwha river, which occupies quadrant three.
@@
 
+Rivers influenced by dams do not appear to deviate appreciably from the rest in February, August, or November.
+However, SD rivers in April divide across the x-axis according to whether they are dammed.
+Those with dams exhibit $\textrm{T}\textsubscript{air} \propto \frac{1}{\textrm{T}\textsubscript{water}}$, while $\textrm{T}\textsubscript{air} \propto \textrm{T}\textsubscript{water}$ for those without.
+Similarly, in June, dammed SD rivers display stronger coupling between T\textsubscript{air} and Q than those without dams.
+
 \section*{Discussion}
 
-The effects of climate on T\textsubscript{water}, determined by dynamic factor analysis, suggest that nearly all rivers included in our dataset were influenced strongly by air temperature, precipitation, and/or snowmelt across 38 years of monthly data (Fig.
-3).
+The effects of climate on T\textsubscript{water}, inferred through dynamic factor analysis, suggest that nearly all rivers included in our dataset were influenced strongly by air temperature, precipitation, and/or snowmelt across 38 years of monthly data (Fig.
+3a-c).
 At most monitoring sites, T\textsubscript{water} closely tracked changes in T\textsubscript{air}, on average responding to increases and decreases with proportional movements of up to 66\% magnitude.
-However, some rivers only weakly track T\textsubscript{air}, and several patterns in the intensity of this coupling correlate strongly with watershed features relating to ice, groundwater, and slope.
-Glaciation and yearly snow burden are prominent among these, and for reasons of ecological and hydrological implication, the primary focus of the following discussion.
+However, some rivers only weakly track T\textsubscript{air}, and patterns in the intensity of this coupling relate primarily to changing landscape features along an elevational gradient.
+Glaciation and yearly snow burden are prominent among these, and for reasons of ecological and hydrological implication, the primary focus of the following discussion, along with the interacting role of dams.
 
-Without any analysis, a "buffering" effect (hereinafter contrasted with "coupling") of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig.
+Before any analysis, a "buffering" effect (the inverse of coupling) of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig.
 2).
@@
 After snowmelt begins to subside, RS and SD recover with a noticeable jump.
-For rivers that receive glacial runoff (SD), this effect appears to remain, buffering them from summer temperature rise where RS rivers instead take on the character of RD (Fig.
+For rivers that receive glacial runoff (SD), this buffering effect appears to remain into the summer months, guarding them from temperature rise when RS rivers instead approach the temperature of RD (Fig.
 4).
-In an extreme case, the Elwha River was actually cooler in August during those years in which air temperature was higher, likely due to increased runoff from Carrie and Eel glaciers.
+In an extreme case, the Elwha River was actually cooler in August during those years in which air temperature was higher, probably due to increased runoff from Carrie and Eel glaciers.
 The buffering effect of ice on river temperature is therefore two-fold, acting first on all snowmelt-influenced rivers through a cold-water pulse in spring, and then on a subset of those rivers throughout summer and fall, by way of glacial runoff.
 For RD rivers, which receive little to no input from ice, summer temperature is entirely dictated by that of the surrounding air, and whatever rain falls through it.
-Though higher-elevation watersheds will always produce colder water, independent of the influence of ice, it can be expected that RS and SD rivers will grow more similar to RD as regional temperatures warm and glaciers decline.
+
+Temperature buffering appears to be enhanced by the presence of artificial impoundments.
+Six rivers included in this study are interrupted by dams or embankments that release stored water from the base.
+The net result of this...
+
+link to next paragraph...
+
+Though higher-elevation watersheds will always produce colder water, independent of the influence of ice and snow, it can be expected that RS and SD rivers will grow more similar to RD as regional temperatures warm and glaciers decline.
 That is to say, formerly reliably cold-water streams and associated habitats may see increases in both summer and winter average temperatures, as well as higher variability from year to year.
@@
 
-The relationship between climate and river temperature is further influenced by the interaction of discharge, and the fates of rivers in the Puget Sound watershed can be best understood by examining these factors in combination (Fig.
+The relationship between climate and river temperature is further influenced by the interaction of Q, and the fates of rivers in the Puget Sound watershed can be best understood by examining these factors in combination (Fig.
 4).
@@
 For example, warmer Aprils on average produced colder water at 9 out of 15 RS and SD sites.
-Though we determined discharge, groundwater, and slope to be likely components of this relationship, only melting ice could be credited with actually reversing it.
+Though we determined Q, groundwater, and slope to be likely components of this relationship, only melting ice could be credited with actually reversing it.
 Projected reductions in snowpack for the Pacific Northwest can therefore be expected to fundamentally alter the responses of currently snow-influenced rivers to yearly variation in spring temperature.
@@
 In some cases, icemelt-influenced rivers exhibit stronger positive responses to climate patterns than their rain-driven counterparts.
-Our results suggest elevational variations in groundwater influx, total discharge, and watershed slope account for these patterns.
+Our results suggest elevational variations in groundwater influx, total Q, and watershed slope account for these patterns.
 However, while these factors may influence the degree of coupling between climatic drivers and water temperature, only snow and ice can reverse it.

### 2017-03-31 manuscript/figures/figscript.R

@@ -43,8 +43,8 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
-# load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-load('temp_due_5m_atpcsn_byMo_allMos.rda')
+load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+# load('temp_due_5m_atpcsn_byMo_allMos.rda')
 # load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
@@ -366,7 +366,7 @@ for(covr in 1:3){
               2, cex=.8, line=1.5)
         axis(2, at=c(-.02,0,.02,.04), labels=c(-0.02,0.00,0.02,0.04), padj=.9, tck=-.02)
         # mtext(bquote(textstyle(bold('Perennial watershed ice/snow coverage'))~textstyle(plain('(%)'))), side=1, cex=.8, line=1.7)
-        mtext(bquote(textstyle(bold('Mean watershed elevation'))~textstyle(plain('(100 m)'))), 
+        mtext(bquote(textstyle(bold('Mean watershed elevation'))~textstyle(plain('(100 m)'))),
               side=1, cex=.8, line=2)
         abline(mod, col='steelblue', lty=2, lwd=3)
     }
@@ -378,8 +378,8 @@ for(covr in 1:3){
     }
     print.letter(paste(letters[covr],sig), c(.9,.9), cex=1.8, font=2, col='steelblue')
     points(landvar, res,
-           bg=cols, col='black', cex=2,
-           # bg=cols, col='black', cex=rescale(log(land$WsAreaSqKm),c(1.2,3.2)),
+           # bg=cols, col='black', cex=2,
+           bg=cols, col='black', cex=rescale(log(land$WsAreaSqKm),c(1.2,3.2)),
            pch=21,
            # pch=land$siteCode,
            lwd=1)
@@ -475,12 +475,12 @@ for(trnd in c(2,1)){
     }
     abline(mod, col='springgreen4', lty=2, lwd=3)
[…]

### 2017-04-01 manuscript/manuscript.tex
figured out my pitch, made table C1

@@
 % \usepackage[percent]{overpic}
-\usepackage{tikz, lineno, natbib, url}
+\usepackage{tikz, lineno, natbib, url, csvsimple, lscape, longtable}
 \linenumbers
@@
 The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database.
-Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), base flow index, soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow coverage (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
+Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow coverage (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
 Monitoring site elevation (m) and presence of upstream dams (as full/partial/no damming of upstream mainstem and major tributaries) were also included.
-Finally, we calculated area above 1000 m (as \% watershed area), mean slope (\% rise), and mean aspect (degree) by delineating and summarizing watersheds from a digital elevation model in ArcMap v.
+Finally, we calculated area above 1000 m (as \% watershed area), mean slope (\% rise), and mean aspect (degree from true north) by delineating and summarizing watersheds from a digital elevation model in ArcMap v.
 10.4 \citep{arcviewenvironmental}.
@@
 Then, we  limited the number of trends to between one and three, to avoid "trend specialization." in other words, we optimized the trends for flexibility while concentrating their explanatory power.
-Additionally, we summarized the landscape predictors with principal coordinates analysis (PCoA), as a way to conceptually "group" them by correlation.
+Additionally, we ordinated the landscape predictors with principal coordinates analysis (PCoA), as a way to conceptually "group" them by correlation.
+Data constrained to irregular, restricted ranges were scaled to [0-1] and arcsine-square-root transformed, along with all proportional data (The logit transform was avoided to prevent generation of infinite values.).
+All continuous data were then centered and scaled to unit variance before PCoA was performed.
 We used the Gower dissimilarity coefficient (Gower's distance) to account for association among both continuous and nominal variables \citep{gowerDist}.
@@
 The buffering effect of ice on river temperature is therefore two-fold, acting first on all snowmelt-influenced rivers through a cold-water pulse in spring, and then on a subset of those rivers throughout summer and fall, by way of glacial runoff.
-For RD rivers, which receive little to no input from ice, summer temperature is entirely dictated by that of the surrounding air, and whatever rain falls through it.
+For RD rivers, which receive little to no input from ice, summer temperature is entirely dictated by that of the surrounding air, and any rain falling through it.
 
-Temperature buffering appears to be enhanced by the presence of artificial impoundments.
-Six rivers included in this study are interrupted by dams or embankments that release stored water from the base.
-The net result of this...
+Temperature buffering by snow and ice appears to be enhanced by the action of artificial impoundments.
+Eight sites on five rivers included in this study are (or were until 2014, in the case of the Elwha RIver) interrupted by dams or embankments that release stored water from the bases of their reservoirs.
+At 33 m, even the shallowest of these reservoirs is deep enough to stratify in summer, meaning released water would come primarily from the cold hypolimnion \citep{olden_hypoCold}.
+This certainly would have affected temperature readings for the Green, Elwha, Cedar and upper Skagit River sites, whose mainstems are or were dammed upstream of the sample location.
+The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam.
+These sites are RS and SD, respectively, and both fall almost exactly on the regression line in Figure 4a.
+The upper Skagit site therefore occupies a middling space of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ coupling between "fully" obstructed and unobstructed SD sites.
 
-link to next paragraph...
+As for the unobstructed SD sites, they appear to oppose the trend exemplified overall.
+In particular, the Duckabush and Puyallup Rivers (upper white circles in Figs 3a, 3b, and 4-Apr.) noticeably break suit with the rest in terms of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$, showing stronger relationships even than many of the RD rivers.
+Compared to all RS and RD rivers, and many SD, these stand out in terms of mean water table depth (Appendix C), suggesting they receive little influence from groundwater influx, which would otherwise serve to decouple $\textrm{T}_{air}$ and $\textrm{T}_{water}$.
+They also occupy smaller watersheds than most of the other SD rivers, which yield lower overall discharge and heat capacity, and thus greater susceptibility to temperature change \citep{caissie2006thermal}.
+There may be additional factors at work in the SD rivers that account for the surprisingly high coupling seen in some unobstructed SD rivers.
+Another potential candidate is watershed slope, which increases with elevation and affects T\textsubscript{water} by influencing residence time and evaporative cooling (via turbulence).
+High slope and elevation are also associated with lower-order tributaries, and thus lower heat capacity.
+
+The role of reservoirs in restructuring natural coupling relationships is complex \citep{damsComplex,WarmingPostDam}, and here confounded with many additional variables.
+Omitting all obstructed sites from Figure 3a, it would appear that no trend exists, yet we believe such omission is unwarranted.
+If the presence of reservoirs negated the influence of other factors, there would be no separation between obstructed sites of different river classes.
+Furthermore, though cold, hypolimnetic outflow sould be expected to buffer T\textsubscript{water} in summer, it alone cannot explain an inverse relationship between T\textsubscript{air} and T\textsubscript{water}.
+Instead, reservoirs may serve to enhance the decoupling of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$ brought on by snowmelt and glacial runoff, by selectively withholding warm water in their epilimnia and admitting cold water through their hypolimnia.
+Evidence for this phenomenon can be seen in the coupling of snowmelt and T\textsubscript{water}, which is generally greater in RS and SD sites downstream of obstructions (Fig.
+3c).
+The Elwha River, which was cleared of its two dams between 2011 and 2014, will provide an excellent opportunity to compare each form of coupling with and without reservoirs, once enough time has passed for signals to overcome interannual variability.
+%ask julian if the selective passage thing has been demonstrated
 
@@
 The Elwha in particular may slip from its current state of high resistance to seasonal climatic changes.
-We tested for changes in mean and variance of $T\textsubscript{air}\rightarrow T\textsubscript{water}$ and $T\textsubscript{air}\rightarrow Q$ coupling between 1978 and 2015, but did not detect any regular patterns (Appendix B).
+We tested for changes in mean and variance of $\textrm{T}\textsubscript{air}\rightarrow \textrm{T}\textsubscript{water}$ and $\textrm{T}\textsubscript{air}\rightarrow \textrm{Q}$ coupling between 1978 and 2015, but did not detect any regular patterns (Appendix B).
 
-In addition to the three climate predictors, five shared trends were fit by the most parsimonious DFA model.
-These represent additional drivers responsible for structuring water temperature across some or all of the 24 sites included in the analysis.
-Each monitoring sites' factor loading on a particular shared trend indicates the degree to which the trend accounted for variance in T\textsubscript{water} at that site.
+In addition to the most parsimonious DFA, we fit a simplified model, designed to focus on what variation in T\textsubscript{water} could be explained by landscape predictors.
+The two trends of this model represent additional drivers responsible for structuring water temperature across some or all of the 24 sites included in the analysis.
 While the precise identities of these drivers cannot be obtained with certainty, they can be inferred through correlation with predictor variables.
+In this way, we determined elevation to be the dominant determinant of $\textrm{T}\textsubscript{air}\rightarrow \textrm{T}\textsubscript{water}$ and $\textrm{precip.}\rightarrow \textrm{T}\textsubscript{water}$, and to %%%%%%
+
 In this way, we determined the most likely landscape drivers of T\textsubscript{water} to be perennial ice and snow cover, mean watershed slope, and groundwater influx.
@@
 
+\section*{Appendix C}
+
+\textbf{Table C1}.
+\begin{landscape}
+\csvautolongtable[respect all]{table_C1a.csv}
+\clearpage
+\csvautolongtable[respect all]{table_C1b.csv}
+% \csvautolongtable[respect all, tabular=|l|c|c|c|c|c|c|c|c|c|c|c|]{table_C1b.csv}
+\end{landscape}
+
 \end{document}

### 2017-04-01 manuscript/figures/figscript.R

@@ -43,8 +43,8 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
-load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-# load('temp_due_5m_atpcsn_byMo_allMos.rda')
+# load('discharge_due_5m_atpcsn_byMo_allMos.rda')
+load('temp_due_5m_atpcsn_byMo_allMos.rda')
 # load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
@@ -323,8 +323,8 @@ dfa = readRDS('../../saved_structures/full_dfaOut.rds')
 land = readRDS('../../saved_structures/full_land.rds')
 
 # system('taskkill /f /im AcroRd32.exe')
-pdf('19_temp_simp_reg.pdf', width=7.5, height=5)
-# pdf('20_temp_simp_reg_NAMES.pdf', width=7.5, height=5)
+# pdf('19_temp_simp_reg.pdf', width=7.5, height=5)
+pdf('20_temp_simp_reg_NAMES.pdf', width=7.5, height=5)
 # layout(matrix(c(1:6,9,7,8),nrow=3,byrow=TRUE))
 defpar = par(mfrow=c(2,3), oma=c(0,0,1,1), mar=c(4,3.5,0,0))
 landvar=land$ElevWs
@@ -380,8 +380,8 @@ for(covr in 1:3){
     points(landvar, res,
            # bg=cols, col='black', cex=2,
            bg=cols, col='black', cex=rescale(log(land$WsAreaSqKm),c(1.2,3.2)),
-           pch=21,
-           # pch=land$siteCode,
+           # pch=21,
+           pch=land$siteCode,
            lwd=1)
     points(landvar[dam_pch], res[dam_pch], col='chocolate2', cex=2, lwd=1, pch=124)
     # color.legend(xl=4,xr=4.4,yb=0.5, yt=0.6, legend=c('147', '1349'),
@@ -492,8 +492,8 @@ for(trnd in c(2,1)){
 
 par(defpar)
 dev.off()
[…]

### 2017-04-04 manuscript/manuscript.tex
finished second draft (not including over-time plots), updated all diagnostic plots

@@
 
-\usepackage{color, listings, bm, amsmath, blkarray, gensymb}
+\usepackage{color, listings, bm, amsmath, blkarray, gensymb, pdfpages}
 % \usepackage[percent]{overpic}
@@
 \pagenumbering{gobble}
-% todo:
-%include hydrograph figures from kathy liederman and julian olden to show
-    %rain dom, rain and snow, snow dom
-%change "dominated" to "dominant"
-%change "fall" to "autumn"
-%see if its possible to directly compare Z and D
-%include the average effect size of tair on twater in 3a
-% can i say how much elevation would inflence temperature independent of snowmelt?
-%examine change over time in just the elwha (mike will ask about this)
-%explain why some SD rivers might have a larger effect than RD (related to slope, groundwater, dams?)
-%add discharge locations to the map?
-%probably should include a table of data sources and station ids
-% make sure figure captions are all formatted appropriately
-% look at differences in elwha pre and post dam
-%mention palmer drought index?
-% find and replace "water temperature" with T_water, same for air
 
@@
 Temperature is among the most important determinants of riverine biodiversity and health.
-It is therefore a primary freshwater management concern, particularly where cold-water fish are of high ecological, recreational, and commercial value.
-However, river temperature in the Puget Sound watershed of the Northwestern U.S.A.
-is affected by a great diversity of drivers at multiple spatial and temporal scales, and little is known of their interactions.
-We used dynamic factor analysis, a multivariate time-series technique, to examine relationships among these drivers, synthesizing long-term climate and fine-scale landcover data.
-We found that primarily rain-fed rivers experience large seasonal temperature fluctuations, which closely track atmospheric temperature, while snow-fed rivers tend to be weakly, and in some cases inversely, coupled with such fluctuations.
-Among watersheds, groundwater influx, land slope, and Q further augment or dampen these relationships.
-Our results suggest the temperature of high-elevation rivers, absent the influence of ice, should be highly variable, and that glacially fed streams stand to see the largest changes in temperature regime under proposed climate scenarios.
+It is therefore a primary freshwater management concern, particularly where temperature-sensitive fish are of high ecological, recreational, and commercial value.
+River temperature in the Puget Sound watershed of the Northwestern U.S.A.
+is affected by a great diversity of drivers at multiple spatial and temporal scales, but little is known of their interactions.
+We used dynamic factor analysis, a multivariate time-series technique for dimension reduction, to examine relationships among these drivers, synthesizing long-term climate and fine-scale land cover data.
+We found that primarily rain-fed rivers undergo large seasonal temperature fluctuations, which closely track air temperature, while snow-fed rivers tend to be more weakly, and in some cases inversely, coupled with air trends.
+However, variation in coupling among snow-fed rivers is high, and disproportionately influenced by artificial reservoirs, which appear to augment the decoupling effect of melting snow and glacial ice in summer.
+Still, our results suggest snow-influenced rivers stand to see the largest changes in temperature regime under projected climate scenarios.
+
 \clearpage
@@
 
-We sought to identify streams in the Puget Sound region whose temperatures fluctuate closely with regional trends in air temperature, precipitation, and snowmelt, and those that depart from regional patterns.
+We sought to identify rivers in the Puget Sound region whose temperatures fluctuate closely with regional trends in air temperature, precipitation, and snowmelt, and those that depart from regional patterns.
 Our second aim was to identify watershed features that correlate with such departures, and thus provide a nuanced basis for predicting impacts of water temperature on aquatic biodiversity and fishery health.
@@
 
-%carter 2005 http://www.swrcb.ca.gov/northcoast/water_issues/programs/tmdls/shasta_river/060707/28appendixaetheeffectsoftemperatureonsteelheadtroutcohosalmonandchinooksalmonbiologyandfunction.pdf:
-    %usepa 1999
-
 \section*{Methods}
-
-% \subsection*{Study region and climate data}
-
-% Monthly time series of mean and max air temperature (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index) were acquired for the time-span of 1978 to 2015.
-
-% https://wa.water.usgs.gov/data/realtime/adr/interactive/ #discharge daily
-% https://waecy.maps.arcgis.com/apps/Viewer/index.html?appid=832e254169e640fba6e117780e137e7b #Q 15min
-% http://www.horizon-systems.com/NHDPlus/NHDPlusV2_17.php #nhdplus
-% ftp://newftp.epa.gov/EPADataCommons/ORD/NHDPlusLandscapeAttributes/StreamCat/HydroRegions/ #streamca
-% ncdc.noaa.gov/cag/time-series/us #climate data
-% https://www.nrcs.usda.gov/wps/portal/nrcs/detail/or/snow/?cid=nrcs142p2_046350 #snow
-% https://arxiv.org/pdf/1509.00660.pdf #TMB
-% Harvey, A. C. 1989. Forecasting, structural time series models and the Kalman flter. Cambridge University Press, Cambridge, UK.
-%CIG puget sound state of knowledge http://cses.washington.edu/picea/mauger/ps-sok/ps-sok_sec03_watercycle_2015.pdf
 
@@
 \begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/map/map.png}}
+\resizebox{\textwidth}{!}{\includegraphics{figures/map/puget_sound_labeled.png}}
 \textbf{Figure 1} Site locations (red points) in relation to combined Washington State Climate Divisions 3 and 4 (colored topography), the region across which climate data were aggregated.
+See Appendix C for site information.
 \end{center}
@@
 The snowmelt time series was assigned zeros for any positive differences (accumulations).
-
-%More information on all datasets can be found in \colorbox{red}{\lstinline{Appendix A}}.
 
@@
 
-% Response data were centered on 0, but not scaled, to avoid error inflation \colorbox{red}{\lstinline{details and citation}}.
-
 Because we were interested in isolating the effects of climatic predictors on T\textsubscript{water} and Q, we used a fixed factor to account for recurring seasonal variation not related to the predictors, with one factor level for each month.
@@
 In other words, it is the change in water temperature accompanying a unit change in precipitation across the whole time series.
-We call this relationship "coupling." We were also interested in coupling by month for T\textsubscript{air}, which required that it be arranged as twelve separate, monthly time series.
+We call this relationship ``coupling.'' We were also interested in coupling by month for T\textsubscript{air}, which required that it be arranged as twelve separate, monthly time series.
 Concretely,
@@
 
-Additional, non seasonal variation due to unknown factors manifests in the shared trends, and a portion of any residual variation is absorbed by error matrix \textbf{v}.
+Additional, non-seasonal variation due to unknown factors manifests in the shared trends, and a portion of any residual variation is absorbed by error matrix \textbf{v}.
 We fit models using four unique error structures (\textbf{R}), to allow for multiple suites of unknown drivers affecting rivers.
@@
 However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, undermining common rules of thumb for admitting additional parameters under AIC \citep{burnham2003model}.
-Thus, we had reason to doubt that the "most parsimonious" model according to AIC alone was any better than a much simpler alternative.
+Thus, we had reason to doubt that the ``most parsimonious'' model according to AIC alone was any better than a much simpler alternative.
 To manage this, we required that each additional trend, covariate, or seasonal structure improve the median coefficient of determination (R\textsuperscript{2}) by at least 1\% in order to justify accepting its attendant complexity.
@@
 2) for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling.
-To achieve this, we amassed an additional dataset of landscape features.
+To achieve this, we amassed an additional dataset of landscape features (Appendix C).
 These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \citep{hill2016stream} and the National Hydrography Dataset (NHDPlusV2; \citealt{mckay2012nhdplus}).
@@
 We removed the monthly factor and the snowmelt predictor from the covariate matrix (\textbf{d}, rows 1-12 and 14), so that the trends would be free to express seasonal and elevational variation.
-Then, we  limited the number of trends to between one and three, to avoid "trend specialization." in other words, we optimized the trends for flexibility while concentrating their explanatory power.
-Additionally, we ordinated the landscape predictors with principal coordinates analysis (PCoA), as a way to conceptually "group" them by correlation.
+Then, we limited the number of trends to between one and three, to avoid ``trend specialization.'' in other words, we optimized the trends for flexibility while concentrating their explanatory power.
+Additionally, we ordinated the landscape predictors with principal coordinates analysis (PCoA), as a way to conceptually ``group'' them by correlation.
 Data constrained to irregular, restricted ranges were scaled to [0-1] and arcsine-square-root transformed, along with all proportional data (The logit transform was avoided to prevent generation of infinite values.).
@@
 
-\section*{Results} %%
+\section*{Results}
 
@@
 \\[\baselineskip]
-% see appendix for standard deviations?
 
-The combined hydrograph of all rivers revealed two primary peaks, one beginning in late spring and the other extending from late fall to early winter, with a prominent trough in late summer.
+The combined hydrograph of all rivers revealed two primary peaks, one beginning in late spring and the other extending from late autumn to early winter, with a prominent trough in late summer.
 Spring peak Q coincided noticeably with a separation in water temperature between SD and RS, while the summer trough coincided with separation of RD and T\textsubscript{air}.
@@
 Length and direction of arrows are proportional to loading of landscape predictors onto each principal axis of their variation.
-% Dotted arrow indicates lack of association with either axis shown
 \\[\baselineskip]
@@
 One stark exception is again the Elwha river, which occupies quadrant three.
-By fall, RS and SD have begun progress back toward their winter states, led by SD.
+By autumn, RS and SD have begun progress back toward their winter states, led by SD.
 RD, meanwhile, remain essentially unmoved from summer.
@@
 At most monitoring sites, T\textsubscript{water} closely tracked changes in T\textsubscript{air}, on average responding to increases and decreases with proportional movements of up to 66\% magnitude.
-However, some rivers only weakly track T\textsubscript{air}, and patterns in the intensity of this coupling relate primarily to changing landscape features along an elevational gradient.
+However, some rivers only weakly tracked T\textsubscript{air}, and patterns in the intensity of this coupling relate primarily to changing landscape features along an elevational gradient (Fig.
+3f).
 Glaciation and yearly snow burden are prominent among these, and for reasons of ecological and hydrological implication, the primary focus of the following discussion, along with the interacting role of dams.
 
-Before any analysis, a "buffering" effect (the inverse of coupling) of ice on river temperature can be seen in the yearly patterns of \textsubscript{water} relative to T\textsubscript{air} (Fig.
+Before any analysis, a ``buffering'' effect (the inverse of coupling) of ice on river temperature can be seen in the yearly patterns of T\textsubscript{water} relative to T\textsubscript{air} (Fig.
 2).
@@
 In an extreme case, the Elwha River was actually cooler in August during those years in which air temperature was higher, probably due to increased runoff from Carrie and Eel glaciers.
-The buffering effect of ice on river temperature is therefore two-fold, acting first on all snowmelt-influenced rivers through a cold-water pulse in spring, and then on a subset of those rivers throughout summer and fall, by way of glacial runoff.
+The buffering effect of ice on river temperature is therefore two-fold, acting first on all snowmelt-influenced rivers through a cold-water pulse in spring, and then on a subset of those rivers throughout summer and autumn, by way of glacial runoff.
 For RD rivers, which receive little to no input from ice, summer temperature is entirely dictated by that of the surrounding air, and any rain falling through it.
@@
 Temperature buffering by snow and ice appears to be enhanced by the action of artificial impoundments.
-Eight sites on five rivers included in this study are (or were until 2014, in the case of the Elwha RIver) interrupted by dams or embankments that release stored water from the bases of their reservoirs.
-At 33 m, even the shallowest of these reservoirs is deep enough to stratify in summer, meaning released water would come primarily from the cold hypolimnion \citep{olden_hypoCold}.
+Eight sites on five rivers included in this study are (or were until 2014, in the case of the Elwha River) interrupted by dams or embankments that release stored water from the bases of their reservoirs.
+At 33 m, even the shallowest of these reservoirs is deep enough to stratify in summer, meaning released water would be delivered from the cold hypolimnion \citep{olden_hypoCold}.
 This certainly would have affected temperature readings for the Green, Elwha, Cedar and upper Skagit River sites, whose mainstems are or were dammed upstream of the sample location.
-The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam.
+The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam, resetting or partially resetting natural conditions \citep{ward1983serial}.
 These sites are RS and SD, respectively, and both fall almost exactly on the regression line in Figure 4a.
-The upper Skagit site therefore occupies a middling space of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ coupling between "fully" obstructed and unobstructed SD sites.
+The upper Skagit site therefore occupies a middling space of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ coupling between ``fully'' obstructed and unobstructed SD sites.
 
 As for the unobstructed SD sites, they appear to oppose the trend exemplified overall.
-In particular, the Duckabush and Puyallup Rivers (upper white circles in Figs 3a, 3b, and 4-Apr.) noticeably break suit with the rest in terms of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$, showing stronger relationships even than many of the RD rivers.
+In particular, the Duckabush and Puyallup Rivers (upper white circles in Figs 3a, 3b, and 4-Apr.) noticeably break suit with the other SD sites in terms of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$, showing stronger relationships even than many of the RD rivers.
 Compared to all RS and RD rivers, and many SD, these stand out in terms of mean water table depth (Appendix C), suggesting they receive little influence from groundwater influx, which would otherwise serve to decouple $\textrm{T}_{air}$ and $\textrm{T}_{water}$.
@@
 
-The role of reservoirs in restructuring natural coupling relationships is complex \citep{damsComplex,WarmingPostDam}, and here confounded with many additional variables.
+The role of reservoirs in restructuring natural temperature coupling relationships is complex \citep{damsComplex,WarmingPostDam}, and here confounded with many additional variables.
 Omitting all obstructed sites from Figure 3a, it would appear that no trend exists, yet we believe such omission is unwarranted.
 If the presence of reservoirs negated the influence of other factors, there would be no separation between obstructed sites of different river classes.
-Furthermore, though cold, hypolimnetic outflow sould be expected to buffer T\textsubscript{water} in summer, it alone cannot explain an inverse relationship between T\textsubscript{air} and T\textsubscript{water}.
+Furthermore, though cold, hypolimnetic outflow should be expected to buffer T\textsubscript{water} in summer, it alone cannot explain an {\it inverse} relationship between T\textsubscript{air} and T\textsubscript{water}.
 Instead, reservoirs may serve to enhance the decoupling of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$ brought on by snowmelt and glacial runoff, by selectively withholding warm water in their epilimnia and admitting cold water through their hypolimnia.
@@
 3c).
-The Elwha River, which was cleared of its two dams between 2011 and 2014, will provide an excellent opportunity to compare each form of coupling with and without reservoirs, once enough time has passed for signals to overcome interannual variability.
-%ask julian if the selective passage thing has been demonstrated
+The Elwha River, which was cleared of its two dams between 2011 and 2014, will provide an excellent opportunity to compare each form of coupling with and without reservoirs, using the same dataset, once enough time has passed for signals to overcome inter-annual variability.
 
 Though higher-elevation watersheds will always produce colder water, independent of the influence of ice and snow, it can be expected that RS and SD rivers will grow more similar to RD as regional temperatures warm and glaciers decline.
-That is to say, formerly reliably cold-water streams and associated habitats may see increases in both summer and winter average temperatures, as well as higher variability from year to year.
+That is to say, formerly reliably cold-water rivers and associated habitats may see increases in both summer and winter average temperatures, as well as higher variability from year to year.
 The Elwha in particular may slip from its current state of high resistance to seasonal climatic changes.
@@
 The two trends of this model represent additional drivers responsible for structuring water temperature across some or all of the 24 sites included in the analysis.
-While the precise identities of these drivers cannot be obtained with certainty, they can be inferred through correlation with predictor variables.
-In this way, we determined elevation to be the dominant determinant of $\textrm{T}\textsubscript{air}\rightarrow \textrm{T}\textsubscript{water}$ and $\textrm{precip.}\rightarrow \textrm{T}\textsubscript{water}$, and to %%%%%%
+While the precise identities of these drivers cannot be obtained with certainty, they can be inferred through their relationships with predictor variables.
+In this way, we found elevation to be one of the dominant determinants of $\textrm{T}\textsubscript{air}\rightarrow \textrm{T}\textsubscript{water}$ and $\textrm{precip.}\rightarrow \textrm{T}\textsubscript{water}$, by driving variation in snow- and icemelt, soil permeability, and slope (Fig.
+3f).
+Dams (reservoirs) and BFI (essentially groundwater contribution) were also major components of variation in temperature coupling, along with water table depth.
+Groundwater, being insulated from the air, maintains relatively constant temperatures throughout the year, particular if it is deep underground.
 
-In this way, we determined the most likely landscape drivers of T\textsubscript{water} to be perennial ice and snow cover, mean watershed slope, and groundwater influx.
-In the case of slope, the likely mechanism of influence is increased turbulence and mixing of water and air in steep, headwater streams, which allows convective warming and cooling to occur more rapidly \citealt{brutsaert1975theory}; Fig.
-3g).
-As for groundwater, greater influx (represented by baseflow index, or BFI; fig.
-3f) corresponds to greater {\it de}-coupling of climatic effects and river temperature, as groundwater should be insulated relative to surface water.
-For the same reason, greater depth of groundwater should be associated with better insulation and thus further decoupling (Fig.
-3d).
-The buffering effect of perennial ice and snow on SD rivers has already been discussed, but the uniquely high factor loadings of RS rivers in relation to the associated trend are worth noting (Fig.
-3e).
-This trend may account for variation in RS due to traits shared by RD and SD, or to a "rain-on-snow" effect that may yield additional cold water in early spring.
-The fifth trend did not correlate strongly with any of the landscape predictors in our dataset.
-It may therefore represent additional, unknown drivers like marine or microclimatic effects, or it may simply account for random noise.
-
-The relationship between climate and river temperature is further influenced by the interaction of Q, and the fates of rivers in the Puget Sound watershed can be best understood by examining these factors in combination (Fig.
+The relationship between climate and river temperature is further influenced by the interaction of discharge, and the fates of rivers in the Puget Sound watershed can be best understood by examining these factors in combination (Fig.
 4).
-Whether rain-, both-, or snow-dominated, all rivers took on RD characteristics in winter, when the effects of ice lay latent.
-As a result, warmer Februaries on average yielded warmer rivers and higher flow (less precipitation bound in ice).
-The critical differences between river classes played out in spring and summer, and it's during these months that future purturbations due to changing climate may be felt most acutely.
+Whether rain-, both-, or snow-dominated, all rivers appear to take on RD characteristics in winter, when the effects of ice lay latent.
+As a result, warmer winters should on average yield warmer rivers and higher flow (less precipitation bound in ice).
+The critical differences between river classes play out in spring and summer, and it's during these months that future perturbations due to changing climate may be felt most acutely.
 For example, warmer Aprils on average produced colder water at 9 out of 15 RS and SD sites.
-Though we determined Q, groundwater, and slope to be likely components of this relationship, only melting ice could be credited with actually reversing it.
 Projected reductions in snowpack for the Pacific Northwest can therefore be expected to fundamentally alter the responses of currently snow-influenced rivers to yearly variation in spring temperature.
@@
 Temperature regimes across the rivers of the Puget Sound watershed are structured by a combination of climatic drivers at the regional scale, and geophysical drivers at watershed scales.
-In the absence of snow and ice, river temperature is closely coupled to that of the surrounding air, while contributions of snowmelt and glacial runoff can dampen or even reverse this coupling in spring and summer.
+In the absence of snow and ice, river temperature is closely coupled to that of the surrounding air, while discharge contributions from snowmelt and glacial runoff can dampen or even reverse this coupling in spring and summer, particularly where hypolimnetic-release reservoirs augment downstream cooling.
 In some cases, icemelt-influenced rivers exhibit stronger positive responses to climate patterns than their rain-driven counterparts.
-Our results suggest elevational variations in groundwater influx, total Q, and watershed slope account for these patterns.
-However, while these factors may influence the degree of coupling between climatic drivers and water temperature, only snow and ice can reverse it.
-Since 1978, such reversals have been widespread, particularly during spring melt.
+Our results suggest elevational variations in groundwater influx and total discharge may account for these patterns.
+However, while these factors and artificial reservoirs may influence the degree of coupling between climatic drivers and water temperature, only snow and ice can reverse it.
+Since 1978, such reversals have been widespread and commonplace, particularly during spring melt.
 Though we did not detect changes in this effect across historical observations, future reductions in snowpack and glacial mass are projected.
@@
 \clearpage
+
+\section*{Acknowledgements}
+We thank Timothy Cline for the use of his TMB script, and Drs.
+Mark Scheuerell, Eric Ward, Eli Holmes, and Adrianne Smits for technical advice.
+Drs.
+Daniel Schindler and Michael Brett provided additional suggestions and guidance.
 
@@
 \subsection*{Temperature DFA output and diagnostics}
-Model selection included four climate covariates (air temperature, precipitation, snowmelt, and hydrological drought), between 1 and 15 shared trends, four within-and-among-site error structures (see methods), and two models of unknown seasonal variation (fixed monthly factors and Fourier series).
-The most parsimonious model of river temperature was selected using the Akaike Information Criterion (AIC), and included air temperature, precipitation, and snowmelt as covariates.
-This model also included five shared trends and an independent and unequally distributed error structure among streams (i.e.
+The process of selecting the best T\textsubscript{water} and discharge models involved four climate covariates (air temperature, precipitation, snowmelt, and hydrological drought), between 1 and 15 shared trends, four within-and-among-site error structures (see methods), and two expressions of unknown seasonal variation (fixed monthly factors and Fourier series).
+The most parsimonious models were selected using the Akaike Information Criterion (AIC) in tandem with R\textsuperscript{2} (required increase of 1\% for each additional parameter), and in each case included air temperature, precipitation, and snowmelt as covariates.
+The T\textsubscript{water} model also included five shared trends and an independent and unequally distributed error structure among rivers (i.e.
 diagonal and unequal variance-covariance matrix).
+The discharge model (not shown) included six trends.
+All subsequent plots relate to the T\textsubscript{water} model, and alphabetic names correspond to sampling sites (Fig 1).
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/01_trends.pdf}}
+\end{center}
 \textbf{Figure A1} Shared trends.
-\end{center}
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/02_loadings.pdf}}
+\end{center}
 \textbf{Figure A2} Factor loadings on shared trends.
-\end{center}
 
 \begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/03_fits.pdf}}
-\textbf{Figure A3} Model fits (gray line = overall; black line = trends-only).
+% \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/03_fits.pdf}}
+\includepdf[pages={1-},scale=0.75]{figures/diagnostic_plots/03_fits.pdf}
 \end{center}
+\textbf{Figure A3} Model fits (green line = overall fit; blue line = trends-only fit, points = data).
 
 \begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/04_residuals.pdf}}
+% \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/04_residuals.pdf}}
+\includepdf[pages={1-},scale=0.75]{figures/diagnostic_plots/04_residuals.pdf}
+\end{center}
 \textbf{Figure A4} Residuals.
-\end{center}
 
 \begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/05_acf.pdf}}
+% \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/05_acf.pdf}}
+\includepdf[pages={1-},scale=0.75]{figures/diagnostic_plots/05_acf.pdf}
+\end{center}
 \textbf{Figure A5} Autocovariance function (ACF).
-\end{center}
 
-Output from the discharge model looks very similar to this, and is omitted here for the sake of retaining a reasonable page count.
 \clearpage
@@
 We used an additional DFA model to test for changes in $T_{air}\rightarrow T_{water}$ coupling over time, by dividing the 1978-2015 time series into 5 intervals and comparing central tendency and variance of effect sizes for each interval.
-Figures B1-B3 show mean effect size for each stream.
+Figures B1-B3 show mean effect size for each river.
 
 To approximate estimates of variability over time, we performed the same analysis within a Bayesian framework, and obtained uncertainty estimates from the credible intervals of the effect size posteriors.
-This approach yielded no trends in variation over time, and is not visualized here (This will be included in the final version of this paper).
+This approach yielded no trends in variation over time, and is not visualized here.
+For Bayesian analyses, we used R package ``statss'' \citep{eric_ward_2017_375646}.
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/05a_temp_effSize_byMonth_acrossTime_may-aug.pdf}}
+\end{center}
 \textbf{Figure B1} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
-\end{center}
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/05b_temp_effSize_byMonth_acrossTime_nov-feb.pdf}}
+\end{center}
 \textbf{Figure B2} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
-\end{center}
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/05c_temp_effSize_byMonth_acrossTime_MASO.pdf}}
+\end{center}
 \textbf{Figure B3} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
-\end{center}
 

### 2017-04-04 02_testing_or_evaluation.R

@@ -64,7 +64,7 @@ if (is.null(dev.list()) == TRUE){
 
 # response choices: COND FC NH3_N NO2_NO3 OP_DIS OXYGEN PH PRESS SUSSOL TEMP TP_P TURB
 # also DISCHARGE (from USGS)
-y_choice = 'DISCHARGE'
+y_choice = 'TEMP'
 # cov choices: meantemp meantemp_anom precip precip_anom hydroDrought hydroDrought_anom
 # maxtemp maxtemp_anom hdd hdd_anom, snowmelt
 #if you're looking at effects by month, or by month over time (see 'design' option below),
@@ -73,7 +73,7 @@ y_choice = 'DISCHARGE'
 #still be determined.
 #(snowmelt only available 1978-2015. also I haven't actually used snowmelt
 #in a model yet, so there could be bugs)
-cov_choices = c('meantemp', 'precip')
+cov_choices = c('meantemp', 'precip','snowmelt')
 #region choices: '3' (lowland), '4' (upland), '3_4' (average of 3 and 4, or each separately)
 #regions 3 and 4 were discovered to be very similar early on, so most of this script will only work if
 #you choose '3_4' here and 'average_regions=TRUE'.
@@ -88,7 +88,7 @@ method = 'fixed_individual'
 startyr = 1978
 endyr = 2015
 #model params (specific values only relevant for testing, not for parameter optimization loop)
-ntrends = 2
+ntrends = 5
 #error matrix can either hold the MARSS specifications or the TMB ones ('DE', 'DUE', 'EVCV', 'UNC')
 obs_err_var_struc = 'DUE'
 #UPDATE: Mark Schueurell no longer scales his response data. scaling forces the variance of
@@ -104,7 +104,7 @@ transform = 'none'
 #choose the covariate matrix design here. options are 'just_effect', 'effect_and_seasonality_without_interaction'
 #'effect_byMonth', 'effect_byMonth_noSeas', and 'effect_byMonth_acrossTime'.
 #(see "designer" function in section 3.1 for details)
-design = 'effect_byMonth_noSeas'
+design = 'effect_byMonth'
[…]

### 2017-04-04 manuscript/figures/figscript.R

@@ -36,7 +36,7 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 
 #load all objects generated during the creation and postprocessing of the best model
 # load('temp_due_4m_at_byMo_allMos.rda')
-# load('temp_due_4m_at_byMo_acrossTime_may-aug.rda')
+load('temp_due_4m_at_byMo_acrossTime_may-aug.rda')
 # load('temp_due_4m_at_byMo_acrossTime_nov-feb.rda')
 # load('temp_due_4m_at_byMo_acrossTime_MASO.rda')
 # load('discharge_due_4m_atpc_byMo_allMos.rda')
@@ -44,7 +44,7 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 # load('discharge_due_4m_atpc_byMo_acrossTime_nov-feb.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
 # load('discharge_due_5m_atpcsn_byMo_allMos.rda')
-load('temp_due_5m_atpcsn_byMo_allMos.rda')
+# load('temp_due_5m_atpcsn_byMo_allMos.rda')
 # load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.

### 2017-04-06 manuscript/manuscript.tex
insignif

@@
     12 & 0 & 0 & 0 & \cdots & 1 \\
-    13 & precip_1 & precip_2 & precip_3 & \cdots & precip_T \\
-    14 & snowmelt_1 & snowmelt_2 & snowmelt_3 & \cdots & snowmelt_T \\
+    13 & precip_1 & precip_2 & precip_3 & \cdots & precip_{456} \\
+    14 & snowmelt_1 & snowmelt_2 & snowmelt_3 & \cdots & snowmelt_{456} \\
     15 & air_1 & 0 & 0 & \cdots & 0 \\
@@
       & \vdots & \vdots & \vdots & \ddots & \vdots \\
-    26 & 0 & 0 & 0 & \cdots & air_T \\
+    26 & 0 & 0 & 0 & \cdots & air_{456} \\
 \end{block}

### 2017-04-07 manuscript/manuscript.tex
changed SDC citation

@@
 This certainly would have affected temperature readings for the Green, Elwha, Cedar and upper Skagit River sites, whose mainstems are or were dammed upstream of the sample location.
-The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam, resetting or partially resetting natural conditions \citep{ward1983serial}.
+The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam, resetting or partially resetting natural conditions \citep{stanford2001revisiting}.
 These sites are RS and SD, respectively, and both fall almost exactly on the regression line in Figure 4a.

### 2017-04-10 manuscript/manuscript.tex
made some plot adjustments

@@
 \documentclass[notitlepage]{article}
+
+%add to appendix A: trends and loadings from 2-trend model
 
@@
 \end{center}
-\textbf{Figure A3} Model fits (green line = overall fit; blue line = trends-only fit, points = data).
+\textbf{Figure A3} Model fits (gray line = overall fit; green line = trends-only fit, points = data).
 
@@
 \textbf{Figure A5} Autocovariance function (ACF).
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/01_trends_simple.pdf}}
+\end{center}
+\textbf{Figure A6} Shared trends from simplified model (no seasonal fixed factor, no snowmelt predictor)
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/02_loadings_simple.pdf}}
+\end{center}
+\textbf{Figure A7} Factor loadings from simplified model (no seasonal fixed factor, no snowmelt predictor)
 

### 2017-04-10 02_testing_or_evaluation.R

@@ -776,9 +776,7 @@ process_plotter_TMB <- function(dfa_obj, ntrends, chunk=NULL, mark_jans=TRUE){
     #mark_jans controls plotting of vertical lines on every january.
     xlm = c(1,int_dates[length(int_dates)])
     if(!is.null(chunk)) xlm=c(chunk*48-47,chunk*48)
-    if(ntrends<=4){
-        par(mai=c(0.5,0.5,0.5,0.1), oma=c(4,4,1,1), mfrow=c(ntrends, 1))
-    } else par(mai=c(0,0,0,0), oma=c(4,4,1,1), mfrow=c(ntrends, 1))
+    par(mai=c(0,0,0,0), oma=c(4,4,1,1), mfrow=c(ntrends, 1))
     xlbl <- int_dates
     y_ts <- int_dates
     ylm <- c(-1,1)*max(abs(dfa_obj$Estimates$u))
@@ -794,7 +792,7 @@ process_plotter_TMB <- function(dfa_obj, ntrends, chunk=NULL, mark_jans=TRUE){
         }
         if(mark_jans) abline(v=seq(1,xlm[2],12), col='gray', lty=2)
         # mtext(paste("Process",i), side=3, line=-2)
-        mtext(paste("Process",i), side=3, line=-2, cex=.8)
+        mtext(paste("Process",i), side=3, line=-1, cex=.8)
         mtext(expression(paste('Standardized ',T[water])), side=2, line=2, outer=TRUE)
         mtext('Year', side=1, line=2.5, outer=TRUE)
         xlbl = xlbl*c(rep(0,11),1)
@@ -806,15 +804,13 @@ process_plotter_TMB <- function(dfa_obj, ntrends, chunk=NULL, mark_jans=TRUE){
     # mtext(expression(paste('Water ', degree, 'C (de-meaned)')),
     # side=2, line=-1.5, outer=TRUE)
 }
-# pdf('../manuscript/figures/diagnostic_plots/01_trends.pdf', width=7, height=5)
+# pdf('../manuscript/figures/diagnostic_plots/01_trends_simp.pdf', width=7, height=5)
 # png('../manuscript/figures/04_processes_and_loadings.png', width=7, height=6, units='in', res=96, type='cairo')
-process_plotter_TMB(dfa, mm, chunk=NULL, mark_jan=TRUE)
+process_plotter_TMB(dfa, mm, chunk=NULL, mark_jan=FALSE)
[…]

### 2017-04-10 manuscript/figures/figscript.R

@@ -36,7 +36,7 @@ print.letter <- function(label="(a)",xy=c(0.1,0.925),...) {
 
 #load all objects generated during the creation and postprocessing of the best model
 # load('temp_due_4m_at_byMo_allMos.rda')
-load('temp_due_4m_at_byMo_acrossTime_may-aug.rda')
+# load('temp_due_4m_at_byMo_acrossTime_may-aug.rda')
 # load('temp_due_4m_at_byMo_acrossTime_nov-feb.rda')
 # load('temp_due_4m_at_byMo_acrossTime_MASO.rda')
 # load('discharge_due_4m_atpc_byMo_allMos.rda')
@@ -45,7 +45,7 @@ load('temp_due_4m_at_byMo_acrossTime_may-aug.rda')
 # load('discharge_due_4m_atpc_byMo_acrossTime_MASO.rda')
 # load('discharge_due_5m_atpcsn_byMo_allMos.rda')
 # load('temp_due_5m_atpcsn_byMo_allMos.rda')
-# load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
+load('../../single_trend_exploration/2trendNoSeasNoSnow.rda')
 
 #add percent watershed ice cover data from 2006, average with those from 2011.
 #NOTE: WsAreaOver1000 is probably a better metric

### 2017-04-13 02_testing_or_evaluation.R
fixed loadings figure and plotter

@@ -829,7 +829,7 @@ loading_plotter_TMB <- function(dfa_obj, ntrends){
         #     if(Z_rot[j,i] > minZ) {text(j, -0.03, ylbl[j], srt=90, adj=1, cex=lab_size)}
         #     if(Z_rot[j,i] < -minZ) {text(j, 0.03, ylbl[j], srt=90, adj=0, cex=lab_size)}
         # }
-        axis(1, 1:nn, labels=colnames(obs_ts), outer=TRUE)
+        axis(1, 1:nn, labels=colnames(obs_ts), outer=TRUE, cex.axis=.8)
         abline(h=0, lwd=1.5, col="gray60", lty=2)
         mtext(paste("Process",i),side=3,line=-1, col='gray30', cex=.8)
         mtext('Factor loading', side=2, line=3, outer=TRUE)
@@ -841,9 +841,9 @@ loading_plotter_TMB <- function(dfa_obj, ntrends){
     }
 }
 # pdf('../manuscript/figures/diagnostic_plots/02_loadings.pdf', width=7, height=4)
-# pdf('../manuscript/figures/diagnostic_plots/02_loadings_simp.pdf', width=7, height=4)
+pdf('../manuscript/figures/diagnostic_plots/02_loadings_simp.pdf', width=7, height=4)
 loading_plotter_TMB(dfa, mm)
-# dev.off()
+dev.off()
 
 # full_fit <- dfa$Estimates$Z %*% dfa$Estimates$u + dfa$Estimates$D %*% rbind(cc,covs_z)
 # identical(full_fit, dfa$Fits)

### 2017-04-18 manuscript/manuscript.tex
added some notes to the manuscript
converted tex to docx, fixed a few small things

@@
 %add to appendix A: trends and loadings from 2-trend model
+%do a comparison of loadings and effect sizes. find the major players
+%make sure x axes of regression plots arent missing numeric labels
+%talk about comparisons between effect sizes: i.e. air temperature is the dominant climatic factor
+%incorporate ashley steele paper on relationship between sun, water, aid
 
@@
 \begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/01_trends_simple.pdf}}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/01_trends_simp.pdf}}
 \end{center}
@@
 \begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/02_loadings_simple.pdf}}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/02_loadings_simp.pdf}}
 \end{center}

### 2017-04-18 manuscript/manuscript_temp.tex
converted tex to docx, fixed a few small things

--- 2017-04-07 manuscript/manuscript.tex
@@
-\documentclass[notitlepage]{article}
-
-\usepackage{color, listings, bm, amsmath, blkarray, gensymb, pdfpages}
-% \usepackage[percent]{overpic}
-\usepackage{tikz, lineno, natbib, url, csvsimple, lscape, longtable}
-\linenumbers
-% \usepackage{graphicx}
-% \usepackage{epsfig}
-% \usepackage{subfig}
-% \captionsetup{belowskip=12pt}
-% \hfuzz=5.002pt  %suppress overfull warnings
-
-\author{Michael Vlah}
 \title{Determining multi-scale controls on river temperature: a time series approach}
-
-\begin{document}
-\pagenumbering{gobble}
-
-\maketitle
-\clearpage
-
-\pagenumbering{arabic}
 
@@
 
-\clearpage
-
 \section*{Introduction}
@@
 The ecological condition of a stream or river, the life it supports, and the goods and services it provides, are influenced by the timing and magnitude of seasonal changes in water temperature.
-Temperature is a chief consideration in the management of fisheries, as it affects species distribution \citep{Boisneau2008}, growth and reproduction \citep{mccullough1999review}, and migration timing \citep{boscarino2007effects}.
-In particular, In the Puget Sound watershed of the American Pacific Northwest, several salmonid species spawn, migrate, and emerge only within the bounds of a few degrees Celsius, and thrive under even greater temperature constraints \citep{carter2005effects}.
+Temperature is a chief consideration in the management of fisheries, as it affects species distribution (Boisneau2008), growth and reproduction (mccullough1999review), and migration timing (boscarino2007effects).
+In particular, In the Puget Sound watershed of the American Pacific Northwest, several salmonid species spawn, migrate, and emerge only within the bounds of a few degrees Celsius, and thrive under even greater temperature constraints (carter2005effects).
 As a result, the success of commercial and recreational fisheries that depend on the region's riverine habitat rests on many precarious factors.
@@
 River networks, being fractal in structure, are naturally governed by environmental processes at multiple scales.
-Seasonal variation in water temperature in rivers of the Pacific Northwest is a function of the surrounding air, as well as precipitation and snowmelt \citep{eldridge1967water}.
-These drivers may in turn be mediated or supplemented by several aspects of watershed morphology at smaller scales, including slope, elevation, and geology \citep{poole2001ecological,lisi2013association}.
+Seasonal variation in water temperature in rivers of the Pacific Northwest is a function of the surrounding air, as well as precipitation and snowmelt (eldridge1967water).
+These drivers may in turn be mediated or supplemented by several aspects of watershed morphology at smaller scales, including slope, elevation, and geology (poole2001ecological,lisi2013association).
 Taken together, this hierarchical system complicates fishery management, as the temperature regime of one river may be the direct product of climate, while that of another may depend more on within-watershed conditions.
 
-Adding to this picture, flow regimes across rivers of the Puget Sound watershed vary with latitude and elevation \citep{reidy2012hydrogeomorphic,mauger2015CIG}, and can be classified broadly into three categories by flow source and hydrograph shape.
+Adding to this picture, flow regimes across rivers of the Puget Sound watershed vary with latitude and elevation (reidy2012hydrogeomorphic,mauger2015CIG), and can be classified broadly into three categories by flow source and hydrograph shape.
 Rain-dominated (RD) rivers receive little or no input from snowmelt, and thus peak in discharge (Q) during the rainy season, usually between October and February.
@@
 
-Effective management plans must therefore integrate a diversity of factors across space and time in order to determine which rivers and watersheds are likely to see consequential changes under projected climate and land use scenarios for the Pacific Northwest \citep{mote2010future,radeloff2012economic}.
+Effective management plans must therefore integrate a diversity of factors across space and time in order to determine which rivers and watersheds are likely to see consequential changes under projected climate and land use scenarios for the Pacific Northwest (mote2010future,radeloff2012economic).
 However, the understanding required to do so is limited by knowledge of relationships among temperature drivers at scale.
-
-% The physical properties of water also change with temperature, which measurably affects the efficacy of electrical power generation (Foerster & Lillestam 2010), and drinking water production (Ramaker et al.
-2005).
-% Physically, temperature determines vapor pressure, surface tension, density (Stevens et al. 1975), gas solubility, reaction rates (Brezonik 1972) and more.
 
@@
 Our second aim was to identify watershed features that correlate with such departures, and thus provide a nuanced basis for predicting impacts of water temperature on aquatic biodiversity and fishery health.
-We hypothesized that water temperature (T\textsubscript{water} would track air temperature T\textsubscript{air} most closely in RD rivers \citep{ward1985thermal,garner2014river}.
-We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt \citep{lisi2015watershed} and with factors affecting heat capacity of water, including Q (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing; \citealt{van2013global}).
+We hypothesized that water temperature (T_water would track air temperature T_air most closely in RD rivers (ward1985thermal,garner2014river).
+We expected deviations from this relationship to correlate best with cold-water influx from snow and ice melt (lisi2015watershed) and with factors affecting heat capacity of water, including Q (volume over time) and watershed slope (which relates to turbulence, surface area, and mixing; (van2013global)).
 
@@
 
-We investigated climate and landscape controls on T\textsubscript{water} and Q, as separate response variables, from 1978 to 2015.
-Monthly time series of water temperature were obtained for 24 river sites via the Washington Department of Ecology's River and Stream Water Quality Monitoring program \citep{DoEwaterData}.
+We investigated climate and landscape controls on T_water and Q, as separate response variables, from 1978 to 2015.
+Monthly time series of water temperature were obtained for 24 river sites via the Washington Department of Ecology's River and Stream Water Quality Monitoring program (DoEwaterData).
 These sites represent 19 nonnested watersheds across 9 counties, and range from 4 to 775 m in elevation.
 For at least one site at each river, monthly Q time series were also available, either from the same location as one of the temperature monitoring sites, or from within 30 km on the same major reach.
-Q data were aggregated by monthly mean from the USGS National Water Information System database \citep{USGSdischarge}.
+Q data were aggregated by monthly mean from the USGS National Water Information System database (USGSdischarge).
 
@@
 \resizebox{\textwidth}{!}{\includegraphics{figures/map/puget_sound_labeled.png}}
-\textbf{Figure 1} Site locations (red points) in relation to combined Washington State Climate Divisions 3 and 4 (colored topography), the region across which climate data were aggregated.
+Figure 1 Site locations (red points) in relation to combined Washington State Climate Divisions 3 and 4 (colored topography), the region across which climate data were aggregated.
 See Appendix C for site information.
@@
 
-Potential climatic predictors of T\textsubscript{water} and Q included mean and max T\textsubscript{air} (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index), averaged by month across the response variable time series.
+Potential climatic predictors of T_water and Q included mean and max T_air (\degree C), total precipitation (cm), snowmelt (cm), and hydrological drought (Palmer Hydrological Drought Index), averaged by month across the response variable time series.
 All but snowmelt were available through the U.S.
-Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI; \citealt{climateData}).
+Climate Divisional Dataset, developed by the National Centers for Environmental Information (NCEI; (climateData)).
 We acquired climatic predictor data grouped by Washington State climate division, and all but two of our sites fell within divisions 3 (Puget Sound Lowland) and 4 (East Olympic/Cascade Foothills; see Fig.
@@
 We therefore aggregated these data by monthly mean across the two regions (after verifying their post-standardization similarity), resulting in a single dataset of four climatic predictor variables.
-A snowmelt time series was then added to this dataset, using monthly mean records from six SNOTEL sites (Bumping Ridge, Elbow Lake, Mount Crag, Park Creek Ridge, Stevens Pass, White Pass) listed by the USDA's Natural Resources Conservation Service; \citealt{snowData}.
+A snowmelt time series was then added to this dataset, using monthly mean records from six SNOTEL sites (Bumping Ridge, Elbow Lake, Mount Crag, Park Creek Ridge, Stevens Pass, White Pass) listed by the USDA's Natural Resources Conservation Service; (snowData).
 We calculated monthly snowmelt for each site as the absolute value of negative differences in cumulative snow water equivalent from each month to the next.
@@
 \subsection*{Time series analysis}
-Response time series (T\textsubscript{water} and Q) were modeled using dynamic factor analysis (DFA; \citealt{zuur2003estimating}), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
+Response time series (T_water and Q) were modeled using dynamic factor analysis (DFA; (zuur2003estimating)), a multivariate technique that can be thought of as an analog to principal component analysis in the time domain.
 In DFA, response time series are fit with a linear combination of shared, random-walk trends (usually many fewer than the total number of response series), predictors (which can have unique effects on each response series), and random error.
 We chose DFA over a traditional multivariate state space approach for two reasons.
-First, it provides advantages in computational efficiency, as a small number of shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost \citep{zuur2003dynamic}.
+First, it provides advantages in computational efficiency, as a small number of shared trends often adequately capture variation across dozens of responses, and at much lower parameter cost (zuur2003dynamic).
 Second, in terms of identifying what drives the shared trends, having fewer of them allows for greater inferential parsimony.
 Being a multivariate technique, DFA also provides an advantage over univariate alternatives in that covariance structure among responses can be specified and compared.
-All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software \citep{kristensen2015tmb}, which we called using package TMB in R \citep{Rmanual,tmbPackage}.
+All models were fit using maximum likelihood estimation by automatic differentiation, with Template Model Builder software (kristensen2015tmb), which we called using package TMB in R (Rmanual,tmbPackage).
 
@@
 \begin{equation}
-    \textbf{x}_t = \textbf{x}_{t-1} + \textbf{w}_t\textrm{, where } \textbf{w}_t \sim \textrm{MVN}(0,\textbf{Q})
+    x_t = x_{t-1} + w_t, where  w_t \sim MVN(0,Q)
 \end{equation}
 \begin{equation}
-    \textbf{y}_t = \textbf{Zx}_t + \textbf{Dd}_t + \textbf{v}_t\textrm{, where } \textbf{v}_t \sim \textrm{MVN}(0,\textbf{R})
+    y_t = Zx_t + Dd_t + v_t, where  v_t \sim MVN(0,R)
 \end{equation}
 \begin{equation}
-    \textbf{x}_0 \sim \textrm{MVN}(0,\bm{\Lambda})
+    x_0 \sim MVN(0,\Lambda)
 \end{equation}
 
-At time step {\it t}, the $m \times 1$ vector of shared trends (\textbf{x}) is a function of \textbf{x} in the previous step, plus normal error (\textbf{w}; $m\times 1$; Eq.
+At time step {\it t}, the m * 1 vector of shared trends (x) is a function of x in the previous step, plus normal error (w; m* 1; Eq.
 1).
 This is the definition of a random walk.
-The $n\times 1$ response vector (\textbf{y}) at time {\it t} is a function of the shared trends and their factor loadings (\textbf{Z}; $n\times m$), covariates (\textbf{d}; $q\times 1$) and their river-specific effects (\textbf{D}; $n\times q$), and a second normal error term (\textbf{v}; $n\times 1$; Eq.
+The n* 1 response vector (y) at time {\it t} is a function of the shared trends and their factor loadings (Z; n* m), covariates (d; q* 1) and their river-specific effects (D; n* q), and a second normal error term (v; n* 1; Eq.
 2).
-\textbf{R} and \textbf{Q} are variance-covariance matrices of order m, and \textbf{Q} is set to identity for model identifiability \citep{harvey1990forecasting}.
-The initial state of the shared trend vector ($\bm{x}_0$) is multivariate-normally distributed with a mean of zero and a diagonal variance-covariance matrix with large variance (e.g.
+R and Q are variance-covariance matrices of order m, and Q is set to identity for model identifiability (harvey1990forecasting).
+The initial state of the shared trend vector (x_0) is multivariate-normally distributed with a mean of zero and a diagonal variance-covariance matrix with large variance (e.g.
 5; Eq.
@@
 
-Because we were interested in isolating the effects of climatic predictors on T\textsubscript{water} and Q, we used a fixed factor to account for recurring seasonal variation not related to the predictors, with one factor level for each month.
-This factor was incorporated into the covariate matrix (\textbf{d}).
-Thus, the coefficient in \textbf{D} relating, say, precipitation (predictor) and T\textsubscript{water} (response), represents the effect size of the former on the latter.
+Because we were interested in isolating the effects of climatic predictors on T_water and Q, we used a fixed factor to account for recurring seasonal variation not related to the predictors, with one factor level for each month.
+This factor was incorporated into the covariate matrix (d).
+Thus, the coefficient in D relating, say, precipitation (predictor) and T_water (response), represents the effect size of the former on the latter.
 In other words, it is the change in water temperature accompanying a unit change in precipitation across the whole time series.
-We call this relationship ``coupling.'' We were also interested in coupling by month for T\textsubscript{air}, which required that it be arranged as twelve separate, monthly time series.
+We call this relationship ``coupling.'' We were also interested in coupling by month for T_air, which required that it be arranged as twelve separate, monthly time series.
 Concretely,
 
-$$
-\textbf{d} = \begin{blockarray}{cccccc}
-& \textrm{Jan}_{1978} & \textrm{Feb}_{1978} & \textrm{Mar}_{1978} & \cdots & \textrm{Dec}_{2015} \\
+d = \begin{blockarray}{cccccc}
+& Jan_{1978} & Feb_{1978} & Mar_{1978} & \cdots & Dec_{2015} \\
 \begin{block}{c(ccccc)}
@@
 \end{blockarray}
-$$
 
-\noindent
-is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and the effects of precipitation (row 13) and snowmelt (row 14), while also yielding the effect of T\textsubscript{air} by month (rows 15-26) on the response (\textbf{y}; Eq.
+is the covariate matrix structure necessary to account for seasonal variation of unknown origin (rows 1-12), and the effects of precipitation (row 13) and snowmelt (row 14), while also yielding the effect of T_air by month (rows 15-26) on the response (y; Eq.
 2).
-This is the covariate structure of the T\textsubscript{water} model we used for subsequent analyses, not including those described in Figure 5d-e, and Appendix B.
+This is the covariate structure of the T_water model we used for subsequent analyses, not including those described in Figure 5d-e, and Appendix B.
 The same form was used for the Q model.
 
-Additional, non-seasonal variation due to unknown factors manifests in the shared trends, and a portion of any residual variation is absorbed by error matrix \textbf{v}.
-We fit models using four unique error structures (\textbf{R}), to allow for multiple suites of unknown drivers affecting rivers.
+Additional, non-seasonal variation due to unknown factors manifests in the shared trends, and a portion of any residual variation is absorbed by error matrix v.
+We fit models using four unique error structures (R), to allow for multiple suites of unknown drivers affecting rivers.
 We included shared variance with zero covariance, individual variance with zero covariance, shared variance with shared covariance, and individual variance with individual covariance.
-Details on these structures and their implications can be found in \citep{holmes2012marss}.
-The best models for T\textsubscript{water} and Q were determined via AIC.
-However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, undermining common rules of thumb for admitting additional parameters under AIC \citep{burnham2003model}.
+Details on these structures and their implications can be found in (holmes2012marss).
+The best models for T_water and Q were determined via AIC.
+However, negligible likelihood improvements can be inflated when multiplied by thousands of data points, undermining common rules of thumb for admitting additional parameters under AIC (burnham2003model).
 Thus, we had reason to doubt that the ``most parsimonious'' model according to AIC alone was any better than a much simpler alternative.
-To manage this, we required that each additional trend, covariate, or seasonal structure improve the median coefficient of determination (R\textsuperscript{2}) by at least 1\% in order to justify accepting its attendant complexity.
+To manage this, we required that each additional trend, covariate, or seasonal structure improve the median coefficient of determination (R^2) by at least 1\% in order to justify accepting its attendant complexity.
 
 \subsection*{Landscape predictors and post-hoc regression}
+
 For post-hoc analyses, monitoring sites were separated into three classes based on relative areal coverage of perennial ice and/snow (hereinafter ``\% glaciation'') and mean elevation across their watersheds.
-The three classes are loosely based on the classification scheme and language of the Climate Impacts Group at the University of Washington \citep{mauger2015CIG}, and are here delineated according to Table 1.
+The three classes are loosely based on the classification scheme and language of the Climate Impacts Group at the University of Washington (mauger2015CIG), and are here delineated according to Table 1.
 
-% \begin{table}
 \begin{center}
-\textbf{Table 1} Watershed classification scheme
+Table 1 Watershed classification scheme
 \end{center}
@@
  \hline
- Rain-dominated & RD & $< 0.7$ & $< 600$ \\
- Rain-and-snow & RS & $< 0.7$ & $\geq 600$ \\
- Snow-dominated & SD & $\geq 0.7$ & - \\
+ Rain-dominated & RD & < 0.7 & < 600 \\
+ Rain-and-snow & RS & < 0.7 & \geq 600 \\
+ Snow-dominated & SD & \geq 0.7 & - \\
  \hline
@@
 \end{center}
-% \end{table}
 
-After model selection, climatic predictor effect sizes (\textbf{D}; Eq.
+After model selection, climatic predictor effect sizes (D; Eq.
 2) for each river were back-transformed to their original scales and regressed against landscape predictors in order to identify possible watershed-scale controls on coupling.
 To achieve this, we amassed an additional dataset of landscape features (Appendix C).
-These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library \citep{hill2016stream} and the National Hydrography Dataset (NHDPlusV2; \citealt{mckay2012nhdplus}).
+These were collected individually for each of the watersheds corresponding to our 24 river sites, using the EPA's StreamCat (stream-catchment) data library (hill2016stream) and the National Hydrography Dataset (NHDPlusV2; (mckay2012nhdplus)).
 Each site was mapped to an individual river reach, defined as a segment bounded on each end by a stream or river source, confluence, or mouth.
 The region contributing flow to this reach (its watershed) was then fetched, along with selected areal data, from the NHDPlusV2 database.
-Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km\textsuperscript{2}), soil permeability (cm hr\textsuperscript{-1}), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo\textsuperscript{-1}), percent perennial ice and snow coverage (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km\textsuperscript{-2} within 100m of streams; 2010 census), riparian road density (km km\textsuperscript{-2}; 2010 census), and percent riparian urban land (NLCD 2011).
+Landscape attributes used as predictors were aggregated by watershed mean where applicable, and include elevation (m), total area (km^2), soil permeability (cm hr^-1), water table depth (cm), bedrock depth (cm), Base Flow Index (BFI; \%), runoff (mm mo^-1), percent perennial ice and snow coverage (National Land Cover Database [NLDC] 2006 and 2011 average), riparian population density (people km^-2 within 100m of streams; 2010 census), riparian road density (km km^-2; 2010 census), and percent riparian urban land (NLCD 2011).
 Monitoring site elevation (m) and presence of upstream dams (as full/partial/no damming of upstream mainstem and major tributaries) were also included.
 Finally, we calculated area above 1000 m (as \% watershed area), mean slope (\% rise), and mean aspect (degree from true north) by delineating and summarizing watersheds from a digital elevation model in ArcMap v.
-10.4 \citep{arcviewenvironmental}.
+10.4 (arcviewenvironmental).
 
-An additional set of post-hoc regressions was performed using factor loadings on shared trends (\textbf{Z}; Eq.2) as dependent variables, with landscape predictors again as independent variables.
+An additional set of post-hoc regressions was performed using factor loadings on shared trends (Z; Eq.2) as dependent variables, with landscape predictors again as independent variables.
 Loadings represent the degree to which each river's temperature fluctuates with the anonymous force driving the corresponding shared trend.
@@
 To facilitate inference by way of the shared trends, we made three simplifications to the model.
-We removed the monthly factor and the snowmelt predictor from the covariate matrix (\textbf{d}, rows 1-12 and 14), so that the trends would be free to express seasonal and elevational variation.
+We removed the monthly factor and the snowmelt predictor from the covariate matrix (d, rows 1-12 and 14), so that the trends would be free to express seasonal and elevational variation.
 Then, we limited the number of trends to between one and three, to avoid ``trend specialization.'' in other words, we optimized the trends for flexibility while concentrating their explanatory power.
@@
 All continuous data were then centered and scaled to unit variance before PCoA was performed.
-We used the Gower dissimilarity coefficient (Gower's distance) to account for association among both continuous and nominal variables \citep{gowerDist}.
+We used the Gower dissimilarity coefficient (Gower's distance) to account for association among both continuous and nominal variables (gowerDist).
 
@@
 RD rivers were consistently warmest throughout the year.
-In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD (Student's t: $p<0.01, F=11.9$).
+In January, RS reached a minimum of 4.4\degree C, and did not significantly differ from SD (Student's t: p<0.01, F=11.9).
 RD only attained a minimum of 5.6\degree C.
@@
 
-Meanwhile, the amplitude of T\textsubscript{air} oscillation exceeded that of any river class, dipping below T\textsubscript{water} in autumn to a minimum of 3.2\degree C in December, and rising above RS and SD in March to an August maximum of 17.4\degree C.
-T\textsubscript{air} did not overtake RD T\textsubscript{water} until August, by which time the latter had begun to decline.
+Meanwhile, the amplitude of T_air oscillation exceeded that of any river class, dipping below T_water in autumn to a minimum of 3.2\degree C in December, and rising above RS and SD in March to an August maximum of 17.4\degree C.
+T_air did not overtake RD T_water until August, by which time the latter had begun to decline.
 
-% \begin{figure}[h!]
 \begin{center}
-% \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
 \resizebox{\textwidth}{!}{\includegraphics{figures/12_air_water_discharge.pdf}}
-% \caption{Epipolar Geometry}
-% \label{fig:air_water_disch}
 \end{center}
-% \end{figure}
 
-\noindent
-\textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
-All depicted series represent discrete data.
-%\setlength{\parskip}{6pt}
-\\[\baselineskip]
+Figure 2 Monthly mean T_water by river class, and T_air and Q across classes, from 1978 to 2015. All depicted series represent discrete data.
 
 The combined hydrograph of all rivers revealed two primary peaks, one beginning in late spring and the other extending from late autumn to early winter, with a prominent trough in late summer.
-Spring peak Q coincided noticeably with a separation in water temperature between SD and RS, while the summer trough coincided with separation of RD and T\textsubscript{air}.
-On average, November marked both the autumn peak in Q and the point at which T\textsubscript{air} fell below T\textsubscript{water}.
+Spring peak Q coincided noticeably with a separation in water temperature between SD and RS, while the summer trough coincided with separation of RD and T_air.
+On average, November marked both the autumn peak in Q and the point at which T_air fell below T_water.
 
@@
 
-% \begin{tikzpicture}
-%     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-%     \draw (-4.1, -2.1) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
-Points are sized by watershed area.
-Regression lines indicate slopes significant at $\alpha=0.05$.};
-% \end{tikzpicture}
 \begin{center}
 \resizebox{\textwidth}{!}{\includegraphics{figures/19_temp_simp_reg.pdf}}
-% \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and between watershed features and factor loadings on shared trends \textbf{(d-e)}.
-Regression lines indicate slopes significant at $\alpha=0.1$.
 \resizebox{\textwidth}{!}{\includegraphics{figures/18_PCoA.pdf}}
 \end{center}
-\textbf{Figure 3} (a-c) Relationships between watershed elevation and climatic effects on T\textsubscript{water}, obtained from full model fit.
+Figure 3 (a-c) Relationships between watershed elevation and climatic effects on T_water, obtained from full model fit.
 (d-e) Relationships between watershed features and factor loadings on shared trends, from constrained model fit.
-Regression lines indicate slopes significant at $\alpha=0.1$.
+Regression lines indicate slopes significant at \alpha=0.1.
 (f) Ordination of landscape predictors by principal coordinates analysis.
 Length and direction of arrows are proportional to loading of landscape predictors onto each principal axis of their variation.
-\\[\baselineskip]
 
-DFA results, aggregated across months and years for each site, revealed a trend toward reduced $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ coupling with increasing watershed elevation ($p=0.04, \textrm{mult.} R^2=0.18$; Fig.
+DFA results, aggregated across months and years for each site, revealed a trend toward reduced T_{air} -> T_{water} coupling with increasing watershed elevation (p=0.04, mult.
+R^2=0.18; Fig.
 3a).
-On average, a 1\degree C change in T\textsubscript{air} corresponded to a $0.53\pm 0.03$\degree C change in T\textsubscript{water} at RD, a $0.51\pm 0.08$\degree C change at RS, and a $0.45\pm 0.17$\degree C change at SD sites.
-A similar trend was observed with respect to $\textrm{precip}\rightarrow \textrm{T}_{water}$ coupling ($p=0.03, \textrm{mult.} R^2=0.21$; Fig.
-3b), where a monthly change in total precipitation of 1 cm corresponded to a $0.02\pm 0.009$\degree C change in T\textsubscript{water} for RD, $-0.003\pm 0.009$\degree C for RS, and $0.004\pm 0.02$\degree C for SD.
-There was no evidence of coupling overall between snowmelt and T\textsubscript{water} (Fig.
-3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R\textsuperscript{2} (See Appendix A.).
-The strongest examples of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip}\rightarrow \textrm{T}_{water}$ coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
+On average, a 1\degree C change in T_air corresponded to a 0.53 +/- 0.03\degree C change in T_water at RD, a 0.51 +/- 0.08\degree C change at RS, and a 0.45 +/- 0.17\degree C change at SD sites.
+A similar trend was observed with respect to precip -> T_{water} coupling (p=0.03, mult.
+R^2=0.21; Fig.
+3b), where a monthly change in total precipitation of 1 cm corresponded to a 0.02 +/- 0.009\degree C change in T_water for RD, -0.003 +/- 0.009\degree C for RS, and 0.004 +/- 0.02\degree C for SD.
+There was no evidence of coupling overall between snowmelt and T_water (Fig.
+3c), but this predictor was included in the most parsimonious DFA model selected via AIC and R^2 (See Appendix A.).
+The strongest examples of T_{air} -> T_{water} and precip -> T_{water} coupling were observed in the Duckabush River, while the weakest examples are from the Elwha River.
 Both rivers drain glaciers of the Olympic Mountain Range, and both are SD.
-Among SD rivers, those influenced by dams appear to couple less strongly with T\textsubscript{air} and precip, but more so with snowmelt.
+Among SD rivers, those influenced by dams appear to couple less strongly with T_air and precip, but more so with snowmelt.
 
-% In addition to the three climate predictors above, the best T\textsubscript{water} model also included five shared trends.
-Of these, four correlated significantly with at least one known watershed predictor.
-Figure 3 depicts the strongest correlated variables with each trend (insets d-e).
-These are, in arbitrary order, mean water table depth ($p<0.001, \textrm{mult.} R^2=0.60$; Fig.
-3d), \% glaciation ($p<0.01, \textrm{mult.} R^2=0.30$; Fig.
-3e), BFI ($p=0.01, \textrm{mult.} R^2=0.25$; Fig.
-3f), and mean slope ($p<0.01, \textrm{mult.} R^2=0.29$; Fig.
-3g).
-The fifth shared trend was not correlated with any variables in the watershed predictor dataset.
 Factor loadings from a constrained, two-trend model each correlated with one of the two dominant, principal axes of variation across landscape predictors, determined by PCoA (Fig.
@@
 The first principal axis was driven by mean watershed slope, snow (\% area > 1000 m) and ice, soil permeability, and other features that vary along elevational gradients, as well as mean elevation itself.
-Watershed's scores along this axis correlated with loadings from one trend, with marginal significance ($p=0.07, \textrm{mult.} R^2=0.14$; Fig.
+Watershed's scores along this axis correlated with loadings from one trend, with marginal significance (p=0.07, mult.
+R^2=0.14; Fig.
 3d).
-The second principal axis was driven by runoff, base flow, and upstream dams, and correlated with the other trend's loadings ($p<0.01, \textrm{mult.} R^2=0.35$; Fig.
+The second principal axis was driven by runoff, base flow, and upstream dams, and correlated with the other trend's loadings (p<0.01, mult.
+R^2=0.35; Fig.
 3e).
@@
 \end{center}
-\textbf{Figure 4} Relationship between $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{T}_{air}\rightarrow \textrm{Q}$.
-Both axes are expressed per 1\degree C change in $\textrm{T}_{air}$.
-The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $\textrm{T}_{air}$.
-Regression lines indicate slopes significant at $\alpha=0.05$.
-\\[\baselineskip]
+Figure 4 Relationship between T_{air} -> T_{water} and T_{air} -> Q.
+Both axes are expressed per 1\degree C change in T_{air}.
+The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and T_{air}.
+Regression lines indicate slopes significant at \alpha=0.05.
 
-To examine possible sub-season interactions between T\textsubscript{air}, T\textsubscript{water} and Q, we performed an additional DFA with Q as the response.
-In both models, T\textsubscript{air} was allowed to have unique monthly effects.
+To examine possible sub-season interactions between T_air, T_water and Q, we performed an additional DFA with Q as the response.
+In both models, T_air was allowed to have unique monthly effects.
 These effects, taken together, can be conceptualized in relation to the four quadrants of the Cartesian coordinate system (increasing clockwise from upper right; Fig.
@@
 
-In mid-winter (exemplified by February), all river classes primarily occupy the first quadrant, signifying $\textrm{T}\textsubscript{air} \propto \textrm{T}\textsubscript{water}$ and $\textrm{T}\textsubscript{air} \propto \textrm{Q}$, where $\propto$ denotes proportionality.
+In mid-winter (exemplified by February), all river classes primarily occupy the first quadrant, signifying T_air \propto T_water and T_air \propto Q, where \propto denotes proportionality.
 RD shows the weakest Q response.
-By spring, many RS and SD sites develop an inverse relationship between T\textsubscript{air} and T\textsubscript{water}, denoted $\textrm{T}\textsubscript{air} \propto \frac{1}{\textrm{T}\textsubscript{water}}$, while RD sites change little from their winter state.
+By spring, many RS and SD sites develop an inverse relationship between T_air and T_water, denoted T_air \propto 1/T_water, while RD sites change little from their winter state.
 June and August see a procession of most sites into the near fourth quadrant, with SD trailing.
-This signifies $\textrm{T}\textsubscript{air} \propto \frac{1}{\textrm{Q}}$, though $\textrm{T}\textsubscript{air} \propto \textrm{T}\textsubscript{water}$.
+This signifies T_air \propto 1/Q, though T_air \propto T_water.
 One stark exception is again the Elwha river, which occupies quadrant three.
@@
 However, SD rivers in April divide across the x-axis according to whether they are dammed.
-Those with dams exhibit $\textrm{T}\textsubscript{air} \propto \frac{1}{\textrm{T}\textsubscript{water}}$, while $\textrm{T}\textsubscript{air} \propto \textrm{T}\textsubscript{water}$ for those without.
-Similarly, in June, dammed SD rivers display stronger coupling between T\textsubscript{air} and Q than those without dams.
+Those with dams exhibit T_air \propto 1/T_water, while T_air \propto T_water for those without.
+Similarly, in June, dammed SD rivers display stronger coupling between T_air and Q than those without dams.
 
@@
 
-The effects of climate on T\textsubscript{water}, inferred through dynamic factor analysis, suggest that nearly all rivers included in our dataset were influenced strongly by air temperature, precipitation, and/or snowmelt across 38 years of monthly data (Fig.
+The effects of climate on T_water, inferred through dynamic factor analysis, suggest that nearly all rivers included in our dataset were influenced strongly by air temperature, precipitation, and/or snowmelt across 38 years of monthly data (Fig.
 3a-c).
-At most monitoring sites, T\textsubscript{water} closely tracked changes in T\textsubscript{air}, on average responding to increases and decreases with proportional movements of up to 66\% magnitude.
-However, some rivers only weakly tracked T\textsubscript{air}, and patterns in the intensity of this coupling relate primarily to changing landscape features along an elevational gradient (Fig.
+At most monitoring sites, T_water closely tracked changes in T_air, on average responding to increases and decreases with proportional movements of up to 66\% magnitude.
+However, some rivers only weakly tracked T_air, and patterns in the intensity of this coupling relate primarily to changing landscape features along an elevational gradient (Fig.
 3f).
@@
 
-Before any analysis, a ``buffering'' effect (the inverse of coupling) of ice on river temperature can be seen in the yearly patterns of T\textsubscript{water} relative to T\textsubscript{air} (Fig.
+Before any analysis, a ``buffering'' effect (the inverse of coupling) of ice on river temperature can be seen in the yearly patterns of T_water relative to T_air (Fig.
 2).
@@
 Eight sites on five rivers included in this study are (or were until 2014, in the case of the Elwha River) interrupted by dams or embankments that release stored water from the bases of their reservoirs.
-At 33 m, even the shallowest of these reservoirs is deep enough to stratify in summer, meaning released water would be delivered from the cold hypolimnion \citep{olden_hypoCold}.
+At 33 m, even the shallowest of these reservoirs is deep enough to stratify in summer, meaning released water would be delivered from the cold hypolimnion (olden_hypoCold).
 This certainly would have affected temperature readings for the Green, Elwha, Cedar and upper Skagit River sites, whose mainstems are or were dammed upstream of the sample location.
-The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam, resetting or partially resetting natural conditions \citep{stanford2001revisiting}.
+The impact of damming on temperature readings at the Skokomish and the lower Skagit River sites should be lesser, as major, unobstructed river forks intercede between sample location and dam, resetting or partially resetting natural conditions (stanford2001revisiting).
 These sites are RS and SD, respectively, and both fall almost exactly on the regression line in Figure 4a.
-The upper Skagit site therefore occupies a middling space of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ coupling between ``fully'' obstructed and unobstructed SD sites.
+The upper Skagit site therefore occupies a middling space of T_{air} -> T_{water} coupling between ``fully'' obstructed and unobstructed SD sites.
 
 As for the unobstructed SD sites, they appear to oppose the trend exemplified overall.
-In particular, the Duckabush and Puyallup Rivers (upper white circles in Figs 3a, 3b, and 4-Apr.) noticeably break suit with the other SD sites in terms of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$, showing stronger relationships even than many of the RD rivers.
-Compared to all RS and RD rivers, and many SD, these stand out in terms of mean water table depth (Appendix C), suggesting they receive little influence from groundwater influx, which would otherwise serve to decouple $\textrm{T}_{air}$ and $\textrm{T}_{water}$.
-They also occupy smaller watersheds than most of the other SD rivers, which yield lower overall discharge and heat capacity, and thus greater susceptibility to temperature change \citep{caissie2006thermal}.
+In particular, the Duckabush and Puyallup Rivers (upper white circles in Figs 3a, 3b, and 4-Apr.) noticeably break suit with the other SD sites in terms of T_{air} -> T_{water} and precip.
+-> T_{water}, showing stronger relationships even than many of the RD rivers.
+Compared to all RS and RD rivers, and many SD, these stand out in terms of mean water table depth (Appendix C), suggesting they receive little influence from groundwater influx, which would otherwise serve to decouple T_{air} and T_{water}.
+They also occupy smaller watersheds than most of the other SD rivers, which yield lower overall discharge and heat capacity, and thus greater susceptibility to temperature change (caissie2006thermal).
 There may be additional factors at work in the SD rivers that account for the surprisingly high coupling seen in some unobstructed SD rivers.
-Another potential candidate is watershed slope, which increases with elevation and affects T\textsubscript{water} by influencing residence time and evaporative cooling (via turbulence).
+Another potential candidate is watershed slope, which increases with elevation and affects T_water by influencing residence time and evaporative cooling (via turbulence).
 High slope and elevation are also associated with lower-order tributaries, and thus lower heat capacity.
 
-The role of reservoirs in restructuring natural temperature coupling relationships is complex \citep{damsComplex,WarmingPostDam}, and here confounded with many additional variables.
+The role of reservoirs in restructuring natural temperature coupling relationships is complex (damsComplex,WarmingPostDam), and here confounded with many additional variables.
 Omitting all obstructed sites from Figure 3a, it would appear that no trend exists, yet we believe such omission is unwarranted.
 If the presence of reservoirs negated the influence of other factors, there would be no separation between obstructed sites of different river classes.
-Furthermore, though cold, hypolimnetic outflow should be expected to buffer T\textsubscript{water} in summer, it alone cannot explain an {\it inverse} relationship between T\textsubscript{air} and T\textsubscript{water}.
-Instead, reservoirs may serve to enhance the decoupling of $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{precip.}\rightarrow \textrm{T}_{water}$ brought on by snowmelt and glacial runoff, by selectively withholding warm water in their epilimnia and admitting cold water through their hypolimnia.
-Evidence for this phenomenon can be seen in the coupling of snowmelt and T\textsubscript{water}, which is generally greater in RS and SD sites downstream of obstructions (Fig.
+Furthermore, though cold, hypolimnetic outflow should be expected to buffer T_water in summer, it alone cannot explain an {\it inverse} relationship between T_air and T_water.
+Instead, reservoirs may serve to enhance the decoupling of T_{air} -> T_{water} and precip.
+-> T_{water} brought on by snowmelt and glacial runoff, by selectively withholding warm water in their epilimnia and admitting cold water through their hypolimnia.
+Evidence for this phenomenon can be seen in the coupling of snowmelt and T_water, which is generally greater in RS and SD sites downstream of obstructions (Fig.
 3c).
@@
 The Elwha in particular may slip from its current state of high resistance to seasonal climatic changes.
-We tested for changes in mean and variance of $\textrm{T}\textsubscript{air}\rightarrow \textrm{T}\textsubscript{water}$ and $\textrm{T}\textsubscript{air}\rightarrow \textrm{Q}$ coupling between 1978 and 2015, but did not detect any regular patterns (Appendix B).
+We tested for changes in mean and variance of T_air -> T_water and T_air -> Q coupling between 1978 and 2015, but did not detect any regular patterns (Appendix B).
 
-In addition to the most parsimonious DFA, we fit a simplified model, designed to focus on what variation in T\textsubscript{water} could be explained by landscape predictors.
+In addition to the most parsimonious DFA, we fit a simplified model, designed to focus on what variation in T_water could be explained by landscape predictors.
 The two trends of this model represent additional drivers responsible for structuring water temperature across some or all of the 24 sites included in the analysis.
 While the precise identities of these drivers cannot be obtained with certainty, they can be inferred through their relationships with predictor variables.
-In this way, we found elevation to be one of the dominant determinants of $\textrm{T}\textsubscript{air}\rightarrow \textrm{T}\textsubscript{water}$ and $\textrm{precip.}\rightarrow \textrm{T}\textsubscript{water}$, by driving variation in snow- and icemelt, soil permeability, and slope (Fig.
+In this way, we found elevation to be one of the dominant determinants of T_air -> T_water and precip.
+-> T_water, by driving variation in snow- and icemelt, soil permeability, and slope (Fig.
 3f).
@@
 In the longer term, changes can be expected for rivers that now receive the temperature-buffering effect of glacial runoff.
-Glaciers continue to decline across North America, with glacial ice across Western Canada projected to decline by 70\% from 2005 to 2100 \citep{clarke2015projected}.
+Glaciers continue to decline across North America, with glacial ice across Western Canada projected to decline by 70\% from 2005 to 2100 (clarke2015projected).
 
@@
 
-\bibliography{manuscript}
-\bibliographystyle{apalike}
-\clearpage
-
 \section*{Appendix A}
@@
 \subsection*{Temperature DFA output and diagnostics}
-The process of selecting the best T\textsubscript{water} and discharge models involved four climate covariates (air temperature, precipitation, snowmelt, and hydrological drought), between 1 and 15 shared trends, four within-and-among-site error structures (see methods), and two expressions of unknown seasonal variation (fixed monthly factors and Fourier series).
-The most parsimonious models were selected using the Akaike Information Criterion (AIC) in tandem with R\textsuperscript{2} (required increase of 1\% for each additional parameter), and in each case included air temperature, precipitation, and snowmelt as covariates.
-The T\textsubscript{water} model also included five shared trends and an independent and unequally distributed error structure among rivers (i.e.
+
+The process of selecting the best T_water and discharge models involved four climate covariates (air temperature, precipitation, snowmelt, and hydrological drought), between 1 and 15 shared trends, four within-and-among-site error structures (see methods), and two expressions of unknown seasonal variation (fixed monthly factors and Fourier series).
+The most parsimonious models were selected using the Akaike Information Criterion (AIC) in tandem with R^2 (required increase of 1\% for each additional parameter), and in each case included air temperature, precipitation, and snowmelt as covariates.
+The T_water model also included five shared trends and an independent and unequally distributed error structure among rivers (i.e.
 diagonal and unequal variance-covariance matrix).
 The discharge model (not shown) included six trends.
-All subsequent plots relate to the T\textsubscript{water} model, and alphabetic names correspond to sampling sites (Fig 1).
+All subsequent plots relate to the T_water model, and alphabetic names correspond to sampling sites (Fig 1).
 
@@
 \end{center}
-\textbf{Figure A1} Shared trends.
+Figure A1 Shared trends.
 
@@
 \end{center}
-\textbf{Figure A2} Factor loadings on shared trends.
+Figure A2 Factor loadings on shared trends.
 
 \begin{center}
-% \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/03_fits.pdf}}
 \includepdf[pages={1-},scale=0.75]{figures/diagnostic_plots/03_fits.pdf}
 \end{center}
-\textbf{Figure A3} Model fits (green line = overall fit; blue line = trends-only fit, points = data).
+Figure A3 Model fits (gray line = overall fit; green line = trends-only fit, points = data).
 
 \begin{center}
-% \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/04_residuals.pdf}}
 \includepdf[pages={1-},scale=0.75]{figures/diagnostic_plots/04_residuals.pdf}
 \end{center}
-\textbf{Figure A4} Residuals.
+Figure A4 Residuals.
 
 \begin{center}
-% \resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/05_acf.pdf}}
 \includepdf[pages={1-},scale=0.75]{figures/diagnostic_plots/05_acf.pdf}
 \end{center}
-\textbf{Figure A5} Autocovariance function (ACF).
+Figure A5 Autocovariance function (ACF).
 
-\clearpage
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/01_trends_simple.pdf}}
+\end{center}
+Figure A6 Shared trends from simplified model (no seasonal fixed factor, no snowmelt predictor)
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/diagnostic_plots/02_loadings_simple.pdf}}
+\end{center}
+Figure A7 Factor loadings from simplified model (no seasonal fixed factor, no snowmelt predictor)
 
@@
 
-We used an additional DFA model to test for changes in $T_{air}\rightarrow T_{water}$ coupling over time, by dividing the 1978-2015 time series into 5 intervals and comparing central tendency and variance of effect sizes for each interval.
+We used an additional DFA model to test for changes in T_{air} -> T_{water} coupling over time, by dividing the 1978-2015 time series into 5 intervals and comparing central tendency and variance of effect sizes for each interval.
 Figures B1-B3 show mean effect size for each river.
@@
 This approach yielded no trends in variation over time, and is not visualized here.
-For Bayesian analyses, we used R package ``statss'' \citep{eric_ward_2017_375646}.
+For Bayesian analyses, we used R package ``statss'' (eric_ward_2017_375646).
 
@@
 \end{center}
-\textbf{Figure B1} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
+Figure B1 Mean T_{air} -> T_{water} coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
 
@@
 \end{center}
-\textbf{Figure B2} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
+Figure B2 Mean T_{air} -> T_{water} coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
 
@@
 \end{center}
-\textbf{Figure B3} Mean $T_{air}\rightarrow T_{water}$ coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
+Figure B3 Mean T_{air} -> T_{water} coupling over time. Each plot corresponds to an individual site. Y-label colors represent mean watershed elevation (bluer=higher).
 
@@
 
-\textbf{Table C1}.
+Table C1.
 \begin{landscape}

### 2017-04-21 manuscript/vlah_chapter2_draft.tex
added draft version of manuscript

--- 2017-04-18 manuscript/manuscript.tex
@@
 
-\author{Michael Vlah}
-\title{Determining multi-scale controls on river temperature: a time series approach}
+\author{Authors: Michael Vlah* and Gordon Holtgrieve}
+\title{Title: Determining multi-scale controls on river temperature: a time series approach}
 
@@
 \maketitle
+\noindent
+Affiliation: School of  Aquatic and Fishery Sciences, University of Washington, Seattle, WA 98105
+\\[\baselineskip]
+\noindent
+*Corresponding author: vlahm13@gmail.com, 412-735-3706
 \clearpage
@@
 Q data were aggregated by monthly mean from the USGS National Water Information System database \citep{USGSdischarge}.
-
-\begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/map/puget_sound_labeled.png}}
-\textbf{Figure 1} Site locations (red points) in relation to combined Washington State Climate Divisions 3 and 4 (colored topography), the region across which climate data were aggregated.
-See Appendix C for site information.
-\end{center}
 
@@
 
-% \begin{table}
-\begin{center}
-\textbf{Table 1} Watershed classification scheme
-\end{center}
-\begin{center}
-\begin{tabular}{ |c|c|c|c| }
- \hline
- Classification & Abb. & Glaciation (\%) & Mean elev. (m) \\
- \hline
- Rain-dominated & RD & $< 0.7$ & $< 600$ \\
- Rain-and-snow & RS & $< 0.7$ & $\geq 600$ \\
- Snow-dominated & SD & $\geq 0.7$ & - \\
- \hline
-\end{tabular}
-\end{center}
-% \end{table}
-
 After model selection, climatic predictor effect sizes (\textbf{D}; Eq.
@@
 
-% \begin{figure}[h!]
-\begin{center}
-% \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
-\resizebox{\textwidth}{!}{\includegraphics{figures/12_air_water_discharge.pdf}}
-% \caption{Epipolar Geometry}
-% \label{fig:air_water_disch}
-\end{center}
-% \end{figure}
-
-\noindent
-\textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
-All depicted series represent discrete data.
-%\setlength{\parskip}{6pt}
-\\[\baselineskip]
-
 The combined hydrograph of all rivers revealed two primary peaks, one beginning in late spring and the other extending from late autumn to early winter, with a prominent trough in late summer.
@@
 Between June and July, RS and SD saw a large jump in temperature, which coincided with the decline in snowmelt.
-
-% \begin{tikzpicture}
-%     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
-%     \draw (-4.1, -2.1) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
-Points are sized by watershed area.
-Regression lines indicate slopes significant at $\alpha=0.05$.};
-% \end{tikzpicture}
-\begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/19_temp_simp_reg.pdf}}
-% \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and between watershed features and factor loadings on shared trends \textbf{(d-e)}.
-Regression lines indicate slopes significant at $\alpha=0.1$.
-\resizebox{\textwidth}{!}{\includegraphics{figures/18_PCoA.pdf}}
-\end{center}
-\textbf{Figure 3} (a-c) Relationships between watershed elevation and climatic effects on T\textsubscript{water}, obtained from full model fit.
-(d-e) Relationships between watershed features and factor loadings on shared trends, from constrained model fit.
-Regression lines indicate slopes significant at $\alpha=0.1$.
-(f) Ordination of landscape predictors by principal coordinates analysis.
-Length and direction of arrows are proportional to loading of landscape predictors onto each principal axis of their variation.
-\\[\baselineskip]
 
@@
 Combined, the first two principal axes accounted for 79.4\% of variation across landscape predictors.
-
-\begin{center}
-\resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
-\end{center}
-\textbf{Figure 4} Relationship between $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{T}_{air}\rightarrow \textrm{Q}$.
-Both axes are expressed per 1\degree C change in $\textrm{T}_{air}$.
-The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $\textrm{T}_{air}$.
-Regression lines indicate slopes significant at $\alpha=0.05$.
-\\[\baselineskip]
 
@@
 
+\clearpage
 \bibliography{manuscript}
@@
 \clearpage
+
+\section*{Tables}
+% \begin{table}
+\begin{center}
+\textbf{Table 1} Watershed classification scheme.
+\end{center}
+\begin{center}
+\begin{tabular}{ |c|c|c|c| }
+ \hline
+ Classification & Abb. & Glaciation (\%) & Mean elev. (m) \\
+ \hline
+ Rain-dominated & RD & $< 0.7$ & $< 600$ \\
+ Rain-and-snow & RS & $< 0.7$ & $\geq 600$ \\
+ Snow-dominated & SD & $\geq 0.7$ & - \\
+ \hline
+\end{tabular}
+\end{center}
+% \end{table}
+
+\section*{Figures}
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/map/puget_sound_labeled.png}}
+\textbf{Figure 1} Site locations (red points) in relation to combined Washington State Climate Divisions 3 and 4 (colored topography), the region across which climate data were aggregated.
+See Appendix C for site information.
+\end{center}
+
+% \begin{figure}[h!]
+\begin{center}
+% \includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}
+\resizebox{\textwidth}{!}{\includegraphics{figures/12_air_water_discharge.pdf}}
+% \caption{Epipolar Geometry}
+% \label{fig:air_water_disch}
+\end{center}
+% \end{figure}
+\noindent
+\textbf{Figure 2} Monthly mean T\textsubscript{water} by river class, and T\textsubscript{air} and Q across classes, from 1978 to 2015.
+All depicted series represent discrete data.
+%\setlength{\parskip}{6pt}
+\\[\baselineskip]
+
+% \begin{tikzpicture}
+%     \draw (0, 0) node[inner sep=0] {\includegraphics[width=\textwidth]{figures/16_temp_all_reg.pdf}};
+%     \draw (-4.1, -2.1) node[text width=9.5em] {\normalsize \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and other watershed predictors and factor loadings on shared trends \textbf{(d-g)}.
+Points are sized by watershed area.
+Regression lines indicate slopes significant at $\alpha=0.05$.};
+% \end{tikzpicture}
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/19_temp_simp_reg.pdf}}
+% \textbf{Figure 3} Relationships between watershed elevation and climatic effects on T\textsubscript{water} \textbf{(a-c)}, and between watershed features and factor loadings on shared trends \textbf{(d-e)}.
+Regression lines indicate slopes significant at $\alpha=0.1$.
+\resizebox{\textwidth}{!}{\includegraphics{figures/18_PCoA.pdf}}
+\end{center}
+\textbf{Figure 3} (a-c) Relationships between watershed elevation and climatic effects on T\textsubscript{water}, obtained from full model fit.
+(d-e) Relationships between watershed features and factor loadings on shared trends, from constrained model fit.
+Regression lines indicate slopes significant at $\alpha=0.1$.
+(f) Ordination of landscape predictors by principal coordinates analysis.
+Length and direction of arrows are proportional to loading of landscape predictors onto each principal axis of their variation.
+
+\begin{center}
+\resizebox{\textwidth}{!}{\includegraphics{figures/15_disch_vs_air_focal5.pdf}}
+\end{center}
+\textbf{Figure 4} Relationship between $\textrm{T}_{air}\rightarrow \textrm{T}_{water}$ and $\textrm{T}_{air}\rightarrow \textrm{Q}$.
+Both axes are expressed per 1\degree C change in $\textrm{T}_{air}$.
+The red quadrant designates proportionality between all three variables, the blue inverse proportionality between each response and $\textrm{T}_{air}$.
+Regression lines indicate slopes significant at $\alpha=0.05$.
+\\[\baselineskip]
 
